---
title: "Session 5: Linear Regression"
subtitle: "From Patterns to Defensible Policy Relationships"
format: 
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true    
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

## ⚙️ Live Setup (Run Once)

```{r setup-data, eval=TRUE, include=FALSE}
library(dplyr)
library(ggplot2)

resolve_data_path <- function(filename) {
  candidates <- c(
    file.path("data", filename),
    file.path("..", "..", "data", filename)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) stop("Could not find data file: ", filename)
  hit
}

ev <- read.csv(resolve_data_path("wa_ev_population.csv"), stringsAsFactors = FALSE)

df <- ev %>%
  filter(
    !is.na(electric_range),
    electric_range > 0,
    !is.na(model_year),
    !is.na(ev_type),
    !is.na(county),
    county != ""
  ) %>%
  mutate(
    bev = ifelse(grepl("BEV", ev_type), 1, 0),
    make = as.factor(make)
  )

set.seed(5120)
idx <- sample.int(nrow(df), size = floor(0.7 * nrow(df)))
df_train <- df[idx, ]
df_test <- df[-idx, ]
```

---

## 🧭 What's Happening Today

- Last Week: 

> Is there evidence to show that there is a relationship?

- This Week
  
> What is the magnitude of the relationship? 

## 🧭 What Is Data Modeling?

Plain definition:

* A model is a simplified version of reality.
* Reality is too complex to capture fully.
* So we use a model like a map.
  * Not perfect
  * Not complete
  * But useful for navigation

Key idea:

> A model is judged by usefulness, not perfection.

---

## ❓ Question Before We Begin

Question:

> How much impact does the car's make year have on the Electric Range of the car?

* A scatterplot might show a pattern.
* But is the pattern real?
* Or could it be random noise?

---

## 📉 Live Pattern Check

```{r hook-plot, eval=TRUE, fig.width=8, fig.height=4.5}

df_sample = df %>% filter(model_year > 2010) %>% filter(model_year < 2021)

ggplot(df_sample, aes(x = model_year, y = electric_range)) +
  geom_point(alpha = 0.07, color = "#1f5c8b") +
  geom_smooth(method = "lm", se = FALSE, color = "#b22222", linewidth = 1) +
  labs(
    x = "Model Year",
    y = "Electric Range (miles)",
    title = "Newer EV model years tend to have longer electric range"
  ) +
  theme_minimal(base_size = 13)
```

---

## 📈 What Is Regression?

So how do we actually quantify the impact we see? Regression helps us answer:

> When $X$ changes, what usually happens to $Y$?

Use regression when outcome $Y$ is numeric:

* EV adoption rate
* Emissions
* Income
* Temperature

Why it matters:

* It turns patterns into measurable relationships.

---

## ⚖️ Explanation vs Prediction

Regression has two jobs:

* **Explanation (policy story)**  
  Which factors are associated with Electric Range?
* **Prediction (forecasting)**  
  What electric range should we expect in a car produced in 2026?

Key line:

> Explaining is understanding relationships. Predicting is accuracy on new data.

---

## ➗ What Is Linear Regression?

Linear regression draws the best straight-line summary:

$$
y = \beta_0 + \beta_1 x + \epsilon
$$

Where:

* $y$: outcome we care about
* $x$: predictor
* $\beta_0$: baseline level
* $\beta_1$: expected change in $y$ when $x$ increases by 1
* $\epsilon$: everything else we did not measure

---

## 🧩 Why $\epsilon$ Matters

* The model never explains everything (because it's a simplified reality).
* Even a good model has misses.
* That leftover part is not failure, it is uncharted reality.

> Regression is not about perfect certainty.  
> It is about useful approximation.

---

## 🧮 OLS: How the Best Line Is Chosen

Residual:

$$
\text{residual} = y - \hat{y}
$$

OLS chooses the line that minimizes:

$$
\sum (y - \hat{y})^2
$$

---

## ❓ Why Squared Errors?

* Positive and negative errors cannot cancel out (square makes everything positive).
* Big misses are penalized more than small misses.
* The optimization is mathematically tractable.

So OLS is not random.
It is a clear, consistent rule for "best fit."

---

## 🏛️ Multi-Variable Regression

But reality is not impacted by a "singular" variable. It's a combination of multiple aspects:

* Model year
* EV type
* Vehicle make
* Other policy/context variables

Model:

$$
Electric Range = \beta_0 + \beta_1(\text{model year}) + \beta_2(\text{BEV}) + \beta_3(\text{make})
$$

---

## 🧠 What's Happening in the Background

- If there are **2 x variables** + **1 y variable**, we are essentially drawing a best fitted line in a **3 dimensional space**, and calculating the OLS.
- Starting 4 dimension, humans can't visualize it, but we can certainly imagine!

---

## 💻 How Results Look in R

Function:
$$
Electric Range = \beta_0 + \beta_1(\text{model year}) + \beta_2(\text{BEV}) + \beta_3(\text{make})
$$

Coefficients:
```{r multiple-results, eval=TRUE, echo = TRUE}
m_multi <- lm(electric_range ~ model_year + bev + make, data = df_train)

round(coef(summary(m_multi))[1:10, ], 3)
```

Adjusted R-Squared:
```{r, eval=TRUE, echo = TRUE}
round(summary(m_multi)$adj.r.squared, 4)
```

---

## 🧷 Holding All Else Constant

Without controls (variables), regression can confuse relationships.

Multiple regression asks:

> What is the relationship between model year and electric range, holding EV type and make constant?

That "holding constant" phrase is the core value.

---

## 🔍 Interpreting Coefficients (Numeric)

Rule:

> For a 1-unit increase in $X$, $Y$ is expected to change by $\beta$ units, holding other variables constant.

Example:

* $\beta = 5.631$ for model year

Interpretation:

* One newer model year is associated with about +5.631 miles of electric range, holding other variables constant.

---

## 🏷️ Interpreting Coefficients (Categorical)

For categories, regression sets one group as baseline.

Example:

* Baseline make = one reference make (alphabetically first in factor coding)
* Coefficient for another make = difference from baseline make

Interpretation:

* Each make coefficient tells how much higher/lower expected electric range is relative to the baseline make, all else equal.

---

## 🧪 Coefficient p-values

For each predictor, test:

$$
H_0: \beta_j = 0 \qquad H_a: \beta_j \neq 0
$$

Meaning:

* $H_0$: predictor has no relationship with outcome
* $H_a$: predictor has a non-zero association

---

## ❓ Interpreting the p-value

Plain language:

> If this predictor truly had no relationship, how likely is a coefficient this large just by chance?

* Small p-value: evidence coefficient is not just noise.
* Large p-value: data cannot distinguish effect from zero.

Warning:

> Significant does not mean important.  
> It only means "probably not zero."

---

## 📏 Confidence Intervals

Suppose:

* $\beta = 2.5$
* 95% CI = [0.5, 4.5]

Common mistake:

* "The coefficient randomly moves between 0.5 and 4.5."

Correct idea:

* The true effect is one fixed number.
* We just do not know it perfectly.

---

## 🎯 What CI Actually Means

Interpretation:

> Based on our data, the true effect is plausibly as low as 0.5 or as high as 4.5.

Analogy:

* Coefficient = best guess
* CI = uncertainty band around that guess

Why CIs matter:

* p-value gives yes/no evidence.
* CI gives effect size and uncertainty.

---

## 📊 $R^2$ vs Adjusted $R^2$

**$R^2$**

* Percent of outcome variation explained.
* Always increases when you add predictors.

**Adjusted $R^2$**

* Penalizes unnecessary complexity.
* Better honesty check in multiple regression.

---

## 🧭 Which One Should You Use?

* Simple one-predictor model -> $R^2$ is usually fine.
* Multiple regression with many predictors -> Adjusted $R^2$ is more informative.

Question to ask:

> Did this variable truly help, or just add complexity?

---

## 🧱 Regression Assumptions: L.I.N.E.

These are not trivia.
They tell us when conclusions can break.

* **L — Linearity**
* **I — Independence**
* **N — Normal residuals**
* **E — Equal variance**

---

## 🧱 L.I.N.E.: Why Each Matters

* **Linearity**: if truth is curved, straight-line model misleads.
* **Independence**: if observations are not independent, p-values are too confident.
* **Normal residuals**: improves reliability of p-values and CIs.
* **Equal variance**: if errors fan out, uncertainty estimates are wrong.

---

## 🩺 Residuals: Most Important Diagnostic

Plain definition:

$$
\text{residual} = \text{actual} - \text{predicted}
$$

Residuals are what the model got wrong.

They reveal:

* Missing variables
* Wrong model shape
* Groups where model underperforms

---

## 🩺 Residual Plot (Live Output)

```{r residual-plot, eval=TRUE, fig.width=8, fig.height=4.5}
plot(
  fitted(m_multi), resid(m_multi),
  xlab = "Fitted values",
  ylab = "Residuals",
  pch = 16,
  col = "#1f5c8b",
  main = "Residuals vs Fitted (Multiple Regression)"
)
abline(h = 0, lty = 2, col = "firebrick")
```

---

## 🧠 Reading Residual Patterns

Good residuals:

* Random cloud around zero

Bad residuals:

* Curve pattern -> nonlinearity
* Funnel pattern -> unequal variance
* One-sided pattern -> systematic bias

---

```{r bad-residual-examples, eval=TRUE, fig.width=10, fig.height=4.5, fig.show='hold', out.width='49%'}
set.seed(42)
x <- runif(250, 0, 10)
y_curve <- 3 + 2 * x + 0.8 * x^2 + rnorm(250, 0, 3)
m_curve <- lm(y_curve ~ x)

y_funnel <- 5 + 4 * x + rnorm(250, 0, 0.5 + x / 1.5)
m_funnel <- lm(y_funnel ~ x)

par(mfrow = c(1, 2))
plot(fitted(m_curve), resid(m_curve), pch = 16, cex = 0.7, col = "#1f5c8b",
     main = "Bad Pattern: Curve", xlab = "Fitted", ylab = "Residual")
abline(h = 0, lty = 2, col = "firebrick")
plot(fitted(m_funnel), resid(m_funnel), pch = 16, cex = 0.7, col = "#1f5c8b",
     main = "Bad Pattern: Funnel", xlab = "Fitted", ylab = "Residual")
abline(h = 0, lty = 2, col = "firebrick")
par(mfrow = c(1, 1))
```

Key line:

> Residuals are the model's blind spots.

---

## 🔮 Prediction Using Data Models

Up to this point, we looked at **Explainability**.
What about **Prediction**?

> How do we know if the model is a good "predictor"?

---

## ✂️ Train-Test Split

- We want to know whether the model we created is good at "prediction"
- Then the **"test"** data should never be seen by the model

* **Training set**: model learns the relationship.
* **Test set**: model is evaluated on unseen data.
* If test performance drops a lot, that is overfitting risk.
* We generally take **70/30 Rule** where we take 70% as training, 30% as testing

```{r split-check, eval=TRUE}
cat("Train rows:", nrow(df_train), "\n")
cat("Test rows:", nrow(df_test), "\n")
cat("Train share:", round(nrow(df_train) / nrow(df), 3), "\n")
```

---

## 📐 Model Evaluation on New Data
Use test-set performance, not training set only.

Three core metrics:

* **MAE**: average absolute miss
* **RMSE**: penalizes large misses more
* **MAPE**: average percent error¬

---

## 📐 MAE, RMSE, MAPE

MAE:

$$
\text{MAE} = \frac{1}{n}\sum |y - \hat{y}|
$$

RMSE:

$$
\text{RMSE} = \sqrt{\frac{1}{n}\sum (y - \hat{y})^2}
$$

MAPE:

$$
\text{MAPE} = \frac{100}{n}\sum \left(\frac{|y-\hat{y}|}{|y|}\right)
$$

---

## ⚠️ Metric Use Cases

* Use **MAE** for stakeholder-friendly average error.
* Use **RMSE** when large errors are especially costly.
* Use **MAPE** for relative percent error comparisons.

Warning:

* MAPE breaks when actual values are near zero.

```{r metrics-live, eval=TRUE}
pred_test <- predict(m_multi, newdata = df_test)
resid_test <- df_test$electric_range - pred_test

mae <- mean(abs(resid_test))
rmse <- sqrt(mean(resid_test^2))
mape <- mean(abs(resid_test / pmax(df_test$electric_range, 1))) * 100

round(c(MAE = mae, RMSE = rmse, MAPE = mape), 2)
```

---

## 🧪 Underfitting vs Overfitting

**Underfitting**

* Model is too simple.
* Misses structure.
* Bad even on training data.

**Overfitting**

* Model is too complex.
* Memorizes noise.
* Looks great on training, fails on new data.

---

## 🎯 Model Goal: Generalization

Goal:

* Balance simplicity and predictive power.
* Perform well on unseen data, not just seen data.

Tool:

* Train-test split helps detect overfitting.

---

## ⚖️ How to Compare Models

```{r model-compare, eval=TRUE}
m1 <- lm(electric_range ~ model_year, data = df_train)
m2 <- lm(electric_range ~ model_year + bev, data = df_train)
m3 <- lm(electric_range ~ model_year + bev + make, data = df_train)

score_model <- function(mod, test_df) {
  p <- predict(mod, newdata = test_df)
  e <- test_df$electric_range - p
  c(
    adj_r2 = summary(mod)$adj.r.squared,
    MAE = mean(abs(e)),
    RMSE = sqrt(mean(e^2))
  )
}

out <- rbind(
  `Model 1: year` = score_model(m1, df_test),
  `Model 2: year + bev` = score_model(m2, df_test),
  `Model 3: year + bev + make` = score_model(m3, df_test)
)

round(out, 3)
```

Pick models using both:

* Inference quality (coefficients, CIs, assumptions)
* Out-of-sample performance (MAE/RMSE on test set)

---

## 🚫 Regression Is Not Causation

Regression shows association, not policy impact.

Causal claims require:

* Experiments, or
* Stronger research designs, and
* Explicit assumptions

This is the most common project mistake.

---

## ✅ Final Workflow Checklist

Every regression analysis should:

1. Plot the relationship first.
2. Fit model.
3. Interpret coefficients in words.
4. Read p-values as evidence about $\beta \neq 0$.
5. Use CI for plausible effect size.
6. Check residuals for blind spots.
7. Evaluate MAE/RMSE/MAPE on test data.
8. Avoid causal language unless justified.

---
