﻿﻿---
title: "Lab 03 – EV Infrastructure Audit"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are part of a state policy team auditing **EV demand vs. charging supply** in Washington. Your task is to clean, summarize, join, and visualize the data to identify where infrastructure is lagging.

**Your Goal:** Build a city- and county-level audit that reveals mismatches between EV population and public charging stations.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the analytic toolkit into your session.
# readr: Fast and "smart" data loading.
# dplyr/tidyr: Cleaning, reshaping, and summarizing.
# ggplot2: Visualization.
# maps: Geographic map outlines.

library(tidyverse)
library(maps)
library(plotly)

# Define file paths (adjust if your folder structure differs)
ev_pop_path <- file.path("../../../data", "wa_ev_population.csv")
ev_stations_path <- file.path("../../../data", "wa_ev_charge_stations.csv")
```

## 1. The Evidence Intake: Load the Data

### 1.1 Load and Inspect
**functions:**  
`read_csv()` – Load data quickly and safely.  
`head()` – Preview the first rows.  
`glimpse()` – Inspect columns and data types.

**logic steps:**  
Load both CSV files into `ev_pop` and `ev_stations`.  
Preview each table with `head()` to verify columns.  
Use `glimpse()` to confirm data types and spot obvious issues.

```{r}
"<write your code here>"
```

**Question 1.1.1:** How many rows and columns does each dataset contain?

---

## 2. Data Types and Categorical Adjustments
We clean strings and convert data types so joins and summaries behave correctly.

### 2.1 Clean Population Table
**functions:**  
`mutate()` – Create or modify columns.  
`str_to_title()` – Standardize casing.  
`case_when()` – Recode categories.  
`as.character()` – Set ZIP/postal codes as text.

**logic steps:**  
Standardize `make` casing to avoid duplicate labels.  
Recode eligibility into **Eligible / Ineligible / Unknown**.  
Convert `postal_code` to character so it matches station ZIPs.

```{r}
"<write your code here>"
```

### 2.2 Clean Stations Table
**functions:**  
`mutate()` – Modify columns.  
`as.character()` – Treat ZIP codes as text.  
`as.factor()` – Convert station group to categorical.

**logic steps:**  
Convert `zip` to character for joining.  
Convert `station_group` to factor for grouping/plotting.

```{r}
"<write your code here>"
```

**Question 2.2.1:** Why must `postal_code` and `zip` have the same data type before joining?

---

## 3. EV Counts by City (Demand)

**functions:**  
`group_by()` – Group by city.  
`summarize()` – Count EVs.  
`arrange(desc())` – Sort from highest to lowest.  
`ggplot()` + `geom_col()` + `coord_flip()` – Bar chart.

**logic steps:**  
Count EVs by city.  
Sort to identify the highest-demand cities.  
Plot the top 15 to visualize concentration.

```{r}
"<write your code here>"
```

**Question 3.1.1:** Which city has the highest EV demand?

---

## 4. EV Station Counts by City (Supply)

**functions:**  
`group_by()` / `summarize()` – Count stations by city.  
`arrange(desc())` – Sort by station count.  
`geom_col()` – Bar chart.

**logic steps:**  
Count stations by city.  
Sort to find the most supplied cities.  
Plot the top 15 station counts for comparison.

```{r}
"<write your code here>"
```

**Question 4.1.1:** Do the top EV-demand cities also appear in the top station-supply list?

---

## 5. The Join Experiment
We test how different joins change the number of cities we observe.

**functions:**  
`left_join()` – Keep all EV-demand cities.  
`inner_join()` – Keep only cities in both tables.  
`full_join()` – Keep all cities from either table.  
`mutate()` / `ifelse()` – Replace missing station counts with 0.

**logic steps:**  
Try left, inner, and full joins.  
Compare row counts across joins to see what each keeps/drops.  
Create a **city audit** with EV-to-station ratios.

```{r}
"<write your code here>"
```

**Question 5.1.1:** Which join produced the most rows, and why?

**Question 5.1.2:** Which join best supports a supply-gap audit? Explain.

---

## 6. EV/Station Ratios by City
We estimate how many EVs compete per public station.

**functions:**  
`distinct()` – Unique city/ZIP mappings.  
`inner_join()` / `left_join()` – Combine county and station data.  
`mutate()` – Compute ratios.  
`geom_col()` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Aggregate station counts by county.  
Merge in EV counts by county.  
Compute EV-to-station ratios.  
Plot counties by ratio to identify shortages.

```{r}
"<write your code here>"
```

**Question 6.1.1:** Which city has the highest EVs-per-station ratio?

**Question 6.1.2:** Why is ratio more informative than raw counts here?

---

## 7. Station Groups by County
We compare **Public vs Private** station ownership in high-demand counties.

**functions:**  
`slice_max()` – Select top counties by station count.  
`pull()` – Extract a vector of county names.  
`geom_bar()` – Stacked bar chart.

**logic steps:**  
Select the top 15 counties by station count.  
Filter stations to those counties.  
Plot station groups with a cleaner legend placement.

```{r}
"<write your code here>"
```

**Question 7.1.1:** Do top counties rely more on public or private stations?

---

## 8. Charger Mix by Station Group
We compare the **share of Level 1 / Level 2 / DC Fast** ports by ownership group.

**functions:**  
`pivot_longer()` – Convert wide tech columns into long form.  
`group_by()` / `summarize()` – Aggregate total ports by group and level.  
`geom_col(position = "fill")` – Compare proportional shares.

**logic steps:**  
Pivot charger-level columns into a long table.  
Sum total ports by `station_group` and `level`.  
Plot proportional shares to compare technology mix.

```{r}
"<write your code here>"
```

**Question 8.1.1:** Which station group has the highest share of DC fast ports?

---

## 9. Geographic Map: EV-to-Station Ratio
We map **EV density vs. station density** by latitude and longitude.

**functions:**  
`group_by()` / `summarize()` – Compute EV and station counts by location.  
`full_join()` – Combine location-based counts.  
`replace_na()` – Handle missing counts.  
`geom_point()` – Plot ratio points.  
`coord_quickmap()` – Preserve map scale.

**logic steps:**  
Count EVs by latitude/longitude.  
Count stations by latitude/longitude.  
Join the two tables and compute the EV/station ratio.  
Plot the ratio on a WA map to see gaps.

```{r}
"<write your code here>"
```

**Question 9.1.1:** Where do you observe the biggest gaps between demand and supply?

---

## 10. Lab Conclusion
Final Reflection: If you were advising policymakers, which 1–2 counties would you prioritize for new public charging stations, and why?
