﻿﻿---
title: "Lab 03 – EV Infrastructure Audit"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are part of a state policy team auditing **EV demand vs. charging supply** in Washington. Your task is to clean, summarize, join, and visualize the data to identify where infrastructure is lagging.

**Your Goal:** Build a city-level audit that reveals mismatches between EV population and public charging stations.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the analytic toolkit into your session.
# readr: Fast and "smart" data loading.
# dplyr/tidyr: Cleaning, reshaping, and summarizing.
# ggplot2: Visualization.
# maps: Geographic map outlines.

library(tidyverse)
library(maps)
library(plotly)
library(ggplot2)

# Define file paths (adjust if your folder structure differs)
ev_pop_path = file.path("../../../data", "wa_ev_population.csv")
ev_stations_path = file.path("../../../data", "wa_ev_charge_stations.csv")
```

## 1. The Evidence Intake: Load the Data

### 1.1 Load and Inspect
**functions:**  
`read_csv()` – Load data quickly and safely.  
`head()` – Preview the first rows.  
`glimpse()` – Inspect columns and data types.

**logic steps:**  
Load both CSV files into `ev_pop` and `ev_stations`.  
Preview each table with `head()` to verify columns.  
Use `glimpse()` to confirm data types and spot obvious issues.

```{r}
ev_pop = read_csv(ev_pop_path)
ev_stations = read_csv(ev_stations_path)

head(ev_pop)
head(ev_stations)

glimpse(ev_pop)
glimpse(ev_stations)
```

**Question 1.1.1:** How many rows and columns does each dataset contain?

---

## 2. Data Types and Categorical Adjustments
We standardize data types and station group labels for consistent summaries and plots.

### 2.1 Clean Stations Table
**functions:**  
`mutate()` – Create cleaned columns.  
`as.character()` – Treat ZIP codes as strings.  
`case_when()` + `str_detect()` – Recode station groups.  
`as.factor()` – Store station groups as a factor.

**logic steps:**  
Convert `zip` to character so leading zeros are preserved and joins/labels are reliable.  
Use `case_when()` with `str_detect()` to collapse raw station labels into Public/Private/Unknown.  
Store the cleaned `station_group` as a factor for stable ordering and legend display.

```{r}
"<write your code here>"
```

**Question 2.2.1:** Why is it helpful to treat `zip` as a character value?

---

## 3. EV Counts by City (Demand)

**functions:**  
`group_by()` – Group by city.  
`summarize()` – Count EVs.  
`arrange(desc())` – Sort from highest to lowest.  
`ggplot()` + `geom_col()` + `coord_flip()` – Bar chart.

**logic steps:**  
Count EVs by city.  
Sort to identify the highest-demand cities.  
Plot the top 15 to visualize concentration.

```{r}
"<write your code here>"
```

**Question 3.1.1:** Which city has the highest EV demand?

---

## 4. EV Station Counts by City (Supply)

**functions:**  
`group_by()` / `summarize()` – Count stations by city.  
`arrange(desc())` – Sort by station count.  
`geom_col()` – Bar chart.

**logic steps:**  
Count stations by city.  
Sort to find the most supplied cities.  
Plot the top 15 station counts for comparison.

```{r}
"<write your code here>"
```

**Question 4.1.1:** Do the top EV-demand cities also appear in the top station-supply list?

---

## 5. The Join Experiment
We test how different joins change the number of cities we observe.

**functions:**  
`left_join()` – Keep all EV-demand cities.  
`inner_join()` – Keep only cities in both tables.  
`full_join()` – Keep all cities from either table.  
`mutate()` / `ifelse()` – Replace missing station counts with 0.

**logic steps:**  
Try left, inner, and full joins.  
Compare row counts across joins to see what each keeps/drops.  
Create a **city audit** with EV-to-station ratios.

```{r}
"<write your code here>"
```

**Question 5.1.1:** Which join produced the most rows, and why?

**Question 5.1.2:** Which join best supports a supply-gap audit? Explain.

---

## 6. EV/Station Ratios by City
We estimate how many EVs compete per public station.

**functions:**  
`mutate()` – Compute EV-to-station ratios.  
`arrange()` – Sort by ratio.  
`slice_head()` – Keep the 20 lowest ratios.  
`ggplot()` + `geom_col()` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Add a `ratio` column as `ev_count / station_count` to measure load per station.  
Sort by the ratio and keep the 20 lowest values to find best-served cities.  
Plot the ratios with ordered bars to highlight relative availability.

```{r}
"<write your code here>"
```

**Question 6.1.1:** Which city has the lowest EVs-per-station ratio?

**Question 6.1.2:** Why is ratio more informative than raw counts here?

---

## 7. Station Groups by City
We compare **Public vs Private** station ownership in high-demand cities.

**functions:**  
`slice_max()` – Select top EV-demand cities.  
`group_by()` + `summarize()` – Count stations by city and group.  
`left_join()` – Attach station groups to top cities.  
`ggplot()` + `geom_bar()` + `coord_flip()` – Plot group totals.

**logic steps:**  
Select the top 10 EV-demand cities using `slice_max()`.  
Count stations by `city` and `station_group` to get group totals.  
Left join those totals onto the top-city list so only high-demand cities appear.  
Plot grouped totals with a flipped axis and simplified legend.

```{r}
"<write your code here>"
```

**Question 7.1.1:** Do top cities rely more on public or private stations?

---

## 8. Charger Mix by Station Group
We compare the **share of Level 1 / Level 2 / DC Fast** ports by ownership group.

**functions:**  
`select()` – Keep charger columns.  
`pivot_longer()` – Reshape wide to long.  
`filter()` – Keep positive, non-missing counts.  
`group_by()` + `summarize()` – Total ports by group and level.  
`ggplot()` + `geom_col(position = "fill")` – Plot proportional shares.

**logic steps:**  
Reshape Level 1/Level 2/DC fast columns into a long `level`/`count` format.  
Filter out missing or zero counts to avoid diluting totals.  
Aggregate total ports by `station_group` and `level`.  
Plot stacked proportions to compare technology mix across ownership types.

```{r}
"<write your code here>"
```

**Question 8.1.1:** Which station group has the highest share of DC fast ports?

---

## 9. Geographic Map: EV-to-Station Ratio
We map **EV density vs. station density** by latitude and longitude.

**functions:**  
`mutate()` + `round()` – Round coordinates for aggregation.  
`group_by()` + `summarize()` – Count EVs and stations by location.  
`full_join()` – Combine EV and station locations.  
`replace_na()` + `ifelse()` – Handle missing counts and zero stations.  
`ggplot()` + `borders()` + `geom_point()` – Map demand and supply.  
`plotly::ggplotly()` – Make the map interactive.

**logic steps:**  
Round lat/long to 2 decimals to create consistent location bins.  
Count EVs and stations within each rounded coordinate pair.  
Full-join the location tables and replace missing counts with zero.  
Compute the EV-per-station ratio, guarding against zero stations.  
Map station counts (color) and EV counts (size) over WA counties to spot gaps.

```{r}
"<write your code here>"
```

**Question 9.1.1:** Where do you observe the biggest gaps between demand and supply?

---

## 10. Lab Conclusion
Final Reflection: If you were advising policymakers, which 1–2 cities would you prioritize for new public charging stations, and why?
