---
title: "Lab 03 – EV Infrastructure Audit"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are part of a state policy team auditing **EV demand vs. charging supply** in Washington. Your task is to clean, summarize, join, and visualize the data to identify where infrastructure is lagging.

**Your Goal:** Build a city-level audit that reveals mismatches between EV population and public charging stations.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the analytic toolkit into your session.
# readr: Fast and "smart" data loading.
# dplyr/tidyr: Cleaning, reshaping, and summarizing.
# ggplot2: Visualization.
# maps: Geographic map outlines.

library(tidyverse)
library(maps)
library(plotly)
library(ggplot2)

# Define file paths (adjust if your folder structure differs)
ev_pop_path = file.path("../../../data", "wa_ev_population.csv")
ev_stations_path = file.path("../../../data", "wa_ev_charge_stations.csv")
```

## 1. The Evidence Intake: Load the Data

### 1.1 Load and Inspect
We load both datasets and confirm the structure is as expected.
**functions:**  
- `read_csv()` – Load CSV files into tibbles.  
- `head()` – Preview the first rows.  
- `glimpse()` – Inspect column types and structure.

**logic steps:**  
- Read the EV population CSV into `ev_pop` and the stations CSV into `ev_stations`.  
- Check the first few rows of each table to confirm expected columns and basic formatting.  
- Inspect column types with `glimpse()` to ensure dates, numerics, and character fields are parsed correctly.

```{r}
ev_pop = read_csv(ev_pop_path)
ev_stations = read_csv(ev_stations_path)

head(ev_pop)
head(ev_stations)

glimpse(ev_pop)
glimpse(ev_stations)
```

**Question 1.1.1:** How many rows and columns does each dataset contain?

---

## 2. Data Types and Categorical Adjustments
We standardize data types and station group labels for consistent summaries and plots.

### 2.1 Clean Stations Table
**functions:**  
- `mutate()` – Create cleaned columns.  
- `as.character()` – Treat ZIP codes as strings.  
- `case_when()` + `str_detect()` – Recode station groups.  
- `as.factor()` – Store station groups as a factor.

**logic steps:**  
- Convert `zip` to character so leading zeros are preserved and joins/labels are reliable.  
- Use `case_when()` with `str_detect()` to collapse raw station labels into Public/Private/Unknown.  
- Store the cleaned `station_group` as a factor for stable ordering and legend display.

```{r}

ev_stations = ev_stations %>%
  mutate(
    zip = as.character(zip),
    station_group = case_when(
      str_detect(station_group, "Public") ~ "Public",
      str_detect(station_group, "Private") ~ "Private",
      TRUE ~ "Unknown"
    ),
    station_group = as.factor(station_group)
  )
```

**Question 2.2.1:** Why is it helpful to treat `zip` as a character value?

---

## 3. EV Counts by City (Demand)
We summarize EV registrations by city and visualize the top 15.

**functions:**  
- `group_by()` + `summarize()` – Count EVs by city.  
- `arrange()` – Sort by EV count.  
- `ggplot()` + `geom_col()` + `coord_flip()` – Plot top cities.  
- `reorder()` – Order bars by count.

**logic steps:**  
- Group registrations by `city` and count rows to get `ev_count`.  
- Sort cities in descending order to surface highest-demand locations.  
- Plot the top 15 cities, ordering bars by count to make rank comparisons clear.

```{r}
city_demand = ev_pop %>%
  group_by(city) %>%
  summarize(ev_count = n()) %>%
  arrange(desc(ev_count))

city_demand

ggplot(head(city_demand, 15), aes(x = reorder(city, ev_count), y = ev_count)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 15 Cities by EV Count", x = "City", y = "EVs") +
  theme_minimal()
```

**Question 3.1.1:** Which city has the highest EV demand?

---

## 4. EV Station Counts by City (Supply)
We count stations by city and visualize the top 15 supply locations.

**functions:**  
- `group_by()` + `summarize()` – Count stations by city.  
- `arrange()` – Sort by station count.  
- `ggplot()` + `geom_col()` + `coord_flip()` – Plot top cities.  
- `reorder()` – Order bars by count.

**logic steps:**  
- Group stations by `city` and count rows to get `station_count`.  
- Sort cities in descending order to identify the most supplied locations.  
- Plot the top 15 station counts, ordering bars by count for quick comparison.

```{r}
city_supply = ev_stations %>%
  group_by(city) %>%
  summarize(station_count = n()) %>%
  arrange(desc(station_count))

city_supply

ggplot(head(city_supply, 15), aes(x = reorder(city, station_count), y = station_count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Cities by Station Count", x = "City", y = "Stations") +
  theme_minimal()
```

**Question 4.1.1:** Do the top EV-demand cities also appear in the top station-supply list?

---

## 5. The Join Experiment
We test how different joins change the number of cities we observe.

**functions:**  
- `left_join()` – Keep all EV cities.  
- `inner_join()` – Keep only cities with both EVs and stations.  
- `full_join()` – Keep all cities from either table.  
- `mutate()` + `ifelse()` – Replace missing station counts with 0.  
- `nrow()` – Compare row counts.

**logic steps:**  
- Create left, inner, and full joins on `city` to see which cities are retained.  
- Compare `nrow()` outputs to understand what each join includes or excludes.  
- Use the left join as the audit base, then replace missing `station_count` with 0.

```{r}
# Try Left Join: Keeps all cities in Washington with EVs
test_left = left_join(city_demand, city_supply, by = "city")

# Try Inner Join: Keeps only cities that have BOTH EVs and Stations
test_inner = inner_join(city_demand, city_supply, by = "city")

# Try Full Join: Keeps every city mentioned in either dataset
test_full = full_join(city_demand, city_supply, by = "city")

city_data = test_left %>%
  mutate(
    station_count = ifelse(is.na(station_count), 0, station_count)
  )

nrow(test_left)
nrow(test_inner)
nrow(test_full)
```

**Question 5.1.1:** Which join produced the most rows, and why?

**Question 5.1.2:** Which join best supports a supply-gap audit? Explain.

---

## 6. EV/Station Ratios by City
We estimate how many EVs compete per public station.

**functions:**  
- `mutate()` – Compute EV-to-station ratios.  
- `arrange()` – Sort by ratio.  
- `slice_head()` – Keep the 20 lowest ratios.  
- `ggplot()` + `geom_col()` + `coord_flip()` – Visualize ratios.

**logic steps:**  
- Add a `ratio` column as `ev_count / station_count` to measure load per station.  
- Sort by the ratio and keep the 20 lowest values to find best-served cities.  
- Plot the ratios with ordered bars to highlight relative availability.

```{r}
city_data = city_data %>%
  mutate(
    ratio = ev_count / station_count
  ) %>%
  arrange(ratio) %>%
  slice_head(n = 20)

city_data

ggplot(city_data, aes(x = reorder(city, -ratio), y = ratio)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(title = "EVs per Public Station by City", x = "City", y = "Ratio") +
  theme_minimal()
```

**Question 6.1.1:** Which city has the lowest EVs-per-station ratio?

**Question 6.1.2:** Why is ratio more informative than raw counts here?

---

## 7. Station Groups by City
We compare **Public vs Private** station ownership in high-demand cities.

**functions:**  
- `slice_max()` – Select top EV-demand cities.  
- `group_by()` + `summarize()` – Count stations by city and group.  
- `left_join()` – Attach station groups to top cities.  
- `ggplot()` + `geom_bar()` + `coord_flip()` – Plot group totals.

**logic steps:**  
- Select the top 10 EV-demand cities using `slice_max()`.  
- Count stations by `city` and `station_group` to get group totals.  
- Left join those totals onto the top-city list so only high-demand cities appear.  
- Plot grouped totals with a flipped axis and simplified legend.

```{r}
top_cities = city_demand %>%
  slice_max(ev_count, n = 10)

grouped_ev_stations = ev_stations %>%
  group_by(city, station_group) %>%
  summarize(station_count = n())

grouped_ev_stations

stations_by_group = top_cities %>%
  left_join(grouped_ev_stations, by = "city")

stations_by_group

ggplot(stations_by_group, aes(x = city, y = station_count, fill = station_group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Station Ownership in Top 10 Cities", x = "City", y = "Total Stations") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

**Question 7.1.1:** Do top cities rely more on public or private stations?

---

## 8. Charger Mix by Station Group
We compare the **share of Level 1 / Level 2 / DC Fast** ports by ownership group.

**functions:**  
- `select()` – Keep charger columns.  
- `pivot_longer()` – Reshape wide to long.  
- `filter()` – Keep positive, non-missing counts.  
- `group_by()` + `summarize()` – Total ports by group and level.  
- `ggplot()` + `geom_col(position = "fill")` – Plot proportional shares.

**logic steps:**  
- Reshape Level 1/Level 2/DC fast columns into a long `level`/`count` format.  
- Filter out missing or zero counts to avoid diluting totals.  
- Aggregate total ports by `station_group` and `level`.  
- Plot stacked proportions to compare technology mix across ownership types.

```{r}
ev_stations[c('id','station_group','ev_lvl_1','ev_lvl_2', 'ev_dc_fast_count')]

tech_long = ev_stations %>%
  select(id, station_group, ev_lvl_1, ev_lvl_2, ev_dc_fast_count) %>%
  pivot_longer(
    cols = c(ev_lvl_1, ev_lvl_2, ev_dc_fast_count),
    names_to = "level",
    values_to = "count"
  ) %>%
  filter(!is.na(count), count > 0)

tech_long

tech_summary = tech_long %>%
  group_by(station_group, level) %>%
  summarize(total_ports = sum(count))

tech_summary

ggplot(tech_summary, aes(x = station_group, y = total_ports, fill = level)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Charger mix by station group",
    subtitle = "Share of Level 1 / Level 2 / DC Fast",
    x = "Station Group",
    y = "Share of ports"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

**Question 8.1.1:** Which station group has the highest share of DC fast ports?

---

## 9. Geographic Map: EV-to-Station Ratio
We map **EV density vs. station density** by latitude and longitude.

**functions:**  
- `mutate()` + `round()` – Round coordinates for aggregation.  
- `group_by()` + `summarize()` – Count EVs and stations by location.  
- `full_join()` – Combine EV and station locations.  
- `replace_na()` + `ifelse()` – Handle missing counts and zero stations.  
- `ggplot()` + `borders()` + `geom_point()` – Map demand and supply.  
- `plotly::ggplotly()` – Make the map interactive.

**logic steps:**  
- Round lat/long to 2 decimals to create consistent location bins.  
- Count EVs and stations within each rounded coordinate pair.  
- Full-join the location tables and replace missing counts with zero.  
- Compute the EV-per-station ratio, guarding against zero stations.  
- Map station counts (color) and EV counts (size) over WA counties to spot gaps.

```{r}
ev_loc = ev_pop %>%
  mutate(
    longitude = round(longitude, 2),
    latitude = round(latitude, 2)
  ) %>%
  group_by(longitude, latitude) %>%
  summarize(ev_count = n())

station_loc = ev_stations %>%
  mutate(
    longitude = round(longitude, 2),
    latitude = round(latitude, 2)
  ) %>%
  group_by(longitude, latitude) %>%
  summarize(station_count = n())

loc_ratio = full_join(ev_loc, station_loc, by = c("longitude", "latitude")) %>%
  mutate(
    ev_count = replace_na(ev_count, 0),
    station_count = replace_na(station_count, 0),
    ratio = ifelse(ev_count / station_count == Inf, 0, ev_count / station_count)
  )

loc_ratio

p_map = ggplot() +
  borders("county", "washington", colour = "grey80", fill = "grey95") +
  geom_point(
    data = loc_ratio,
    aes(x = longitude, y = latitude, color = station_count, size = ev_count),
    alpha = 0.6
  ) +
  scale_color_viridis_c(option = "C", name = "Stations") +
  scale_size_continuous(name = "EVs") +
  coord_quickmap() +
  theme_void() +
  labs(title = "WA EV Demand (size) and Station Supply (color)")

p_map

plotly::ggplotly(p_map)
```

**Question 9.1.1:** Where do you observe the biggest gaps between demand and supply?

---

## 10. Lab Conclusion
Final Reflection: If you were advising policymakers, which 1–2 cities would you prioritize for new public charging stations, and why?

