---
title: "Lab 03 – EV Infrastructure Audit"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are part of a state policy team auditing **EV demand vs. charging supply** in Washington. Your task is to clean, summarize, join, and visualize the data to identify where infrastructure is lagging.

**Your Goal:** Build a city-level audit that reveals mismatches between EV population and public charging stations.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the analytic toolkit into your session.
# readr: Fast and "smart" data loading.
# dplyr/tidyr: Cleaning, reshaping, and summarizing.
# ggplot2: Visualization.
# maps: Geographic map outlines.

library(tidyverse)
library(maps)
library(plotly)

# Define file paths (adjust if your folder structure differs)
ev_pop_path <- file.path("../../../data", "wa_ev_population.csv")
ev_stations_path <- file.path("../../../data", "wa_ev_charge_stations.csv")
```

## 1. The Evidence Intake: Load the Data

### 1.1 Load and Inspect
**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Load both CSV files into `ev_pop` and `ev_stations`.  
Preview each table with ``head()`` to verify columns.  
Use ``glimpse()`` to confirm data types and spot obvious issues.

```{r}
ev_pop <- read_csv(ev_pop_path)
ev_stations <- read_csv(ev_stations_path)

head(ev_pop)
head(ev_stations)

glimpse(ev_pop)
glimpse(ev_stations)
```

**Question 1.1.1:** How many rows and columns does each dataset contain?

---

## 2. Data Types and Categorical Adjustments
We clean strings and convert data types so summaries and plots behave correctly.

### 2.1 Clean Stations Table
**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Convert `zip` to character for consistency.  
Recode `station_group` into Public/Private/Unknown for grouping/plotting.

```{r}

ev_stations <- ev_stations %>%
  mutate(
    zip = as.character(zip),
    station_group = case_when(
      str_detect(station_group, "Public") ~ "Public",
      str_detect(station_group, "Private") ~ "Private",
      TRUE ~ "Unknown"
    ),
    station_group = as.factor(station_group)
  )
```

**Question 2.2.1:** Why is it helpful to treat `zip` as a character value?

---

## 3. EV Counts by City (Demand)

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Count EVs by city.  
Sort to identify the highest-demand cities.  
Plot the top 15 to visualize concentration.

```{r}
city_demand <- ev_pop %>%
  group_by(city) %>%
  summarize(ev_count = n()) %>%
  arrange(desc(ev_count))

city_demand

ggplot(head(city_demand, 15), aes(x = reorder(city, ev_count), y = ev_count)) +
  `geom_col()` +
  coord_flip() +
  labs(title = "Top 15 Cities by EV Count", x = "City", y = "EVs") +
  theme_minimal()
```

**Question 3.1.1:** Which city has the highest EV demand?

---

## 4. EV Station Counts by City (Supply)

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Count stations by city.  
Sort to find the most supplied cities.  
Plot the top 15 station counts for comparison.

```{r}
city_supply <- ev_stations %>%
  group_by(city) %>%
  summarize(station_count = n()) %>%
  arrange(desc(station_count))

city_supply

ggplot(head(city_supply, 15), aes(x = reorder(city, station_count), y = station_count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Cities by Station Count", x = "City", y = "Stations") +
  theme_minimal()
```

**Question 4.1.1:** Do the top EV-demand cities also appear in the top station-supply list?

---

## 5. The Join Experiment
We test how different joins change the number of cities we observe.

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Try left, inner, and full joins.  
Compare row counts across joins to see what each keeps/drops.  
Create a **city audit** table with EV counts and station counts.

```{r}
# Try Left Join: Keeps all cities in Washington with EVs
test_left <- left_join(city_demand, city_supply, by = "city")

# Try Inner Join: Keeps only cities that have BOTH EVs and Stations
test_inner <- inner_join(city_demand, city_supply, by = "city")

# Try Full Join: Keeps every city mentioned in either dataset
test_full <- full_join(city_demand, city_supply, by = "city")

city_data = test_left %>%
  mutate(
    station_count = ifelse(is.na(station_count), 0, station_count)
  )

nrow(test_left)
nrow(test_inner)
nrow(test_full)
```

**Question 5.1.1:** Which join produced the most rows, and why?

**Question 5.1.2:** Which join best supports a supply-gap audit? Explain.

---

## 6. EV/Station Ratios by City
We estimate how many EVs compete per public station.

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Compute EV-to-station ratios.  
Take the bottom 20 cities with the lowest EV-to-station ratio.  
Plot cities by ratio to identify where stations are most available per EV.

```{r}
city_data <- city_data %>%
  mutate(
    ratio = ev_count / station_count
  ) %>%
  arrange(ratio) %>%
  slice_head(n = 20)

city_data

ggplot(city_data, aes(x = reorder(city, -ratio), y = ratio)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(title = "EVs per Public Station by City", x = "City", y = "Ratio") +
  theme_minimal()
```

**Question 6.1.1:** Which city has the lowest EVs-per-station ratio?

**Question 6.1.2:** Why is ratio more informative than raw counts here?

---

## 7. Station Groups by City
We compare **Public vs Private** station ownership in high-demand cities.

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Select the top 10 cities by EV count.  
Count stations by city and station group.  
Left join with the top-demand cities.  
Plot station groups with a cleaner legend placement.

```{r}
top_cities <- city_demand %>%
  slice_max(ev_count, n = 10)

grouped_ev_stations <- ev_stations %>%
  group_by(city, station_group) %>%
  summarize(station_count = n())

grouped_ev_stations

stations_by_group <- top_cities %>%
  left_join(grouped_ev_stations, by = "city")

stations_by_group

ggplot(stations_by_group, aes(x = city, y = station_count, fill = station_group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Station Ownership in Top 10 Cities", x = "City", y = "Total Stations") +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

**Question 7.1.1:** Do top cities rely more on public or private stations?

---

## 8. Charger Mix by Station Group
We compare the **share of Level 1 / Level 2 / DC Fast** ports by ownership group.

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Pivot charger-level columns into a long table.  
Sum total ports by `station_group` and `level`.  
Plot proportional shares to compare technology mix.

```{r}
ev_stations

tech_long <- ev_stations %>%
  select(id, station_group, ev_lvl_1, ev_lvl_2, ev_dc_fast_count) %>%
  pivot_longer(
    cols = c(ev_lvl_1, ev_lvl_2, ev_dc_fast_count),
    names_to = "level",
    values_to = "count"
  ) %>%
  filter(!is.na(count), count > 0)

tech_long

tech_summary <- tech_long %>%
  group_by(station_group, level) %>%
  summarize(total_ports = sum(count))

tech_summary

ggplot(tech_summary, aes(x = station_group, y = total_ports, fill = level)) +
  `geom_col(position = "fill")` +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Charger mix by station group",
    subtitle = "Share of Level 1 / Level 2 / DC Fast",
    x = "Station Group",
    y = "Share of ports"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

**Question 8.1.1:** Which station group has the highest share of DC fast ports?

---

## 9. Geographic Map: EV-to-Station Ratio
We map **EV density vs. station density** by latitude and longitude.

**functions:**  
``mutate()`` – Compute ratios.  
`arrange()` / `slice_`head()`` – Select lowest ratios.  
``geom_col()`` + `coord_flip()` – Visualize ratios.

**logic steps:**  
Count EVs by latitude/longitude.  
Count stations by latitude/longitude.  
Join the two tables and compute the EV/station ratio.  
Plot the ratio on a WA map to see gaps.

```{r}
ev_loc <- ev_pop %>%
  mutate(
    longitude = round(longitude, 2),
    latitude = round(latitude, 2)
  ) %>%
  group_by(longitude, latitude) %>%
  summarize(ev_count = n())

station_loc <- ev_stations %>%
  mutate(
    longitude = round(longitude, 2),
    latitude = round(latitude, 2)
  ) %>%
  group_by(longitude, latitude) %>%
  summarize(station_count = n())

loc_ratio <- full_join(ev_loc, station_loc, by = c("longitude", "latitude")) %>%
  mutate(
    ev_count = replace_na(ev_count, 0),
    station_count = replace_na(station_count, 0),
    ratio = ev_count / ifelse(station_count == 0, 1, station_count)
  )

loc_ratio

p_map <- ggplot() +
  borders("county", "washington", colour = "grey80", fill = "grey95") +
  geom_point(
    data = loc_ratio,
    aes(x = longitude, y = latitude, color = station_count, size = ev_count),
    alpha = 0.6
  ) +
  scale_color_viridis_c(option = "C", name = "Stations") +
  scale_size_continuous(name = "EVs") +
  `coord_quickmap()` +
  theme_void() +
  labs(title = "WA EV Demand (size) and Station Supply (color)")

p_map

plotly::ggplotly(p_map)
```

**Question 9.1.1:** Where do you observe the biggest gaps between demand and supply?

---

## 10. Lab Conclusion
Final Reflection: If you were advising policymakers, which 1–2 cities would you prioritize for new public charging stations, and why?





