---
title: "Session 9: APIs in R"
subtitle: "From Endpoints to Usable Policy Data"
format: 
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

## 🧭 Opening Hook

Policy question:

> "How do we pull up-to-date climate and policy indicators without manually downloading CSV files every week?"

Today’s answer:

* Use APIs
* Pull data directly in R
* Clean it into analysis-ready tables

---

## 🎯 Session Goal

By the end of this session, students will be able to:

* Explain what an API is in plain language
* Build and send API requests in R
* Parse JSON into data frames
* Handle common API errors
* Pull data from 2-3 APIs for policy analysis

---

## 🌐 What Is an API?

API = **Application Programming Interface**

Plain language:

* An API is a waiter between your code and someone else’s data system.
* You send a request.
* The server sends a response.

---

## 📦 API Request and Response

Basic flow:

1. You call an endpoint URL.
2. You pass parameters (country, date, variable).
3. Server returns data (usually JSON).
4. You parse and analyze in R.

---

## 🧩 Terms You Must Know

* **Endpoint**: API URL location
* **Query parameters**: filters like `?country=US&year=2024`
* **Status code**: server outcome (200, 404, 429, etc.)
* **JSON**: common response format

---

## 🚫 Why This Matters for Policy Work

Without APIs:

* Manual downloads
* Stale files
* Hard-to-repeat workflows

With APIs:

* Fresh data
* Reproducible pulls
* Easy updates for reports/dashboards

---

## 🛠️ API Workflow in R

Core pipeline:

1. Build URL
2. Send request
3. Check status
4. Parse JSON
5. Convert to dataframe
6. Validate columns and types

---

## 💻 Live Demo Pattern (Minimal)

```{r api-pattern, eval=FALSE}
library(jsonlite)

url <- "https://api.example.com/data?country=US&year=2024"
raw <- fromJSON(url)
str(raw)
```

---

## ☀️ API Demo 1: Open-Meteo (No Key)

Use case:

* Pull recent weather for a location
* Connect weather to policy planning

```{r open-meteo-demo, eval=TRUE}
library(jsonlite)
library(dplyr)

lat <- 38.9072
lon <- -77.0369

meteo_url <- paste0(
  "https://api.open-meteo.com/v1/forecast?",
  "latitude=", lat,
  "&longitude=", lon,
  "&daily=temperature_2m_max,precipitation_sum",
  "&timezone=auto"
)

meteo_raw <- fromJSON(meteo_url)
weather_df <- as.data.frame(meteo_raw$daily)
head(weather_df)
```

---

## 🌋 API Demo 2: USGS Earthquake (No Key)

Use case:

* Pull events and summarize risk activity

```{r usgs-demo, eval=TRUE}
library(jsonlite)
library(dplyr)

usgs_url <- "https://earthquake.usgs.gov/fdsnws/event/1/query?format=geojson&limit=200"
usgs_raw <- fromJSON(usgs_url, flatten = TRUE)

quake_df <- usgs_raw$features %>%
  select(properties.mag, properties.place, properties.time) %>%
  rename(magnitude = properties.mag, place = properties.place, timestamp = properties.time)

head(quake_df)
```

---

## 🔐 Authentication Basics

Some APIs require keys/tokens.

Where keys usually go:

* Query string: `?api_key=...`
* Header: `Authorization: Bearer ...`

Best practice:

* Store secrets in `.Renviron`, not in scripts.

---

## 🔑 Common Authentication Methods

1. **No auth**
   - Anyone can use this API.
2. **API key**
   - A password for your app.
3. **Header key**
   - Same key, just sent in a different place (request header).
4. **Bearer token**
   - A temporary access pass.
5. **Basic auth**
   - Username + password.
6. **OAuth2**
   - Login through another service (like Google), then get a token.

---

## 🗂️ Where to Store Secrets

Use `.Renviron`:

```{r env-example, eval=FALSE}
MY_API_KEY=abc123
MY_API_TOKEN=xyz456
```

Then in R:

```{r env-use, eval=FALSE}
Sys.getenv("MY_API_KEY")
```

Plain rule:

* Keep keys out of your code.
* Put keys in `.Renviron`.
* Use `Sys.getenv()` to read them.
* Never push keys to GitHub.

---

## ⚠️ Common Status Codes

* `200` OK
* `400` Bad request (wrong parameters)
* `401/403` Unauthorized/Forbidden
* `404` Endpoint not found
* `429` Too many requests (rate limit)
* `500` Server error

---

## 🧯 Debug Checklist

When API call fails:

1. Check URL and endpoint spelling
2. Check required parameters
3. Check authentication
4. Check status code
5. Test with small query first

---

## 📉 Rate Limits and Reliability

Good habits:

* Don’t spam endpoints
* Cache raw pulls locally
* Add short delays for loops
* Log pull date/time

---

## 🧪 Clean API Function Pattern

```{r api-function-pattern, eval=TRUE}
get_weather <- function(lat, lon) {
  url <- paste0(
    "https://api.open-meteo.com/v1/forecast?",
    "latitude=", lat, "&longitude=", lon,
    "&daily=temperature_2m_max&timezone=auto"
  )
  raw <- jsonlite::fromJSON(url)
  as.data.frame(raw$daily)
}
```

---

## 🧠 Policy Interpretation Reminder

API pulls give data access, not automatic conclusions.

You still need:

* data checks
* context
* careful interpretation

---

## 🧪 Lab Transition

In lab, you will:

1. Connect to 2-3 APIs in R
2. Parse JSON into dataframes
3. Build quick summaries/plots
4. Explain what each pull is useful for

---

## ✅ Final Summary

APIs let you:

* Access live data
* Build reproducible pipelines
* Move from manual files to automated analysis workflows

Key line:

> Analysts who can pull API data can answer policy questions faster and better.
