---
title: "Lab 01 – Data Anatomy & Type Checks"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
Welcome to your first day as a Junior Analyst for the **Laser League**. You’ve been handed a raw data export of player performance. Before the Commissioner can use this to rank the world's best players, you need to prove that the data is structured correctly and free of "Gremlins."

**Your Goal:** Perform a forensic audit. Identify what data types R assigned, check the table's structural integrity, and document any quality issues.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the forensic toolkit into your session.
# readr: Part of the Tidyverse, used for fast and "smart" data loading.
# dplyr: Provides the tools for inspection (glimpse) and manipulation.

library(readr)
library(dplyr)

# Setting global options to control code visibility via params
knitr::opts_chunk$set(echo = params$show_code, message = FALSE, warning = FALSE)

# Define the root path (Adjust if your folder structure is different)
ENIA_ROOT <- "~/ENIA-5120"
league_path <- file.path("../../../data", "laser_league_players.csv")
```

## 1. The Blueprint: Loading & Inspection
Our first task is to bring the evidence onto our digital workbench (the R Environment).

### 1.1 Load the Evidence
**functions:**  
`read_csv()`: Loads data much faster and smarter than base R's read.csv().  
`head()`: Displays the first 6 rows of a dataset.

**logic steps:**  
Use `read_csv()` to load the file located at `league_path`.  
Store the result in an object named `players`.  
Use `head()` to verify that the table looks correct.


```{r}
players <- read_csv(league_path)
head(players)
```

### 1.2 Identify the Data Structure
**functions:**  
`class()`: Tells you the "Object Type" (e.g., tibble/data.frame).  
`glimpse()`: Provides a clean, vertical overview of column names and variable types.

**logic steps:**  
Check the `class()` of your `players` object to see if R treats it as a table.  
Use `glimpse()` to see every column name and its assigned data type (e.g., <chr>, <dbl>).


```{r}
"<write your code here>"
```

**Question 1.2.1:** What R class is players? How does R internally store this table?

**Question 1.2.2:** Based on glimpse(), how many rows (observations) and columns (variables) are present?

## 2. The Language: Data Type Checks
In R, the type of data determines what math is allowed. You cannot calculate the "average" of a word.

### 2.1 Inspect Individual Columns
**functions:**  
`class(df$column)`: Extracts a specific column using the $ sign and reveals its type.

**logic steps:**  
Use the `$` operator to check the class of `player_id`.  
Check the class of `season_score`.  
Check the class of `reaction_time_sec`.


```{r}
"<write your code here>"
```

**Question 2.1.1:** assists shows up as <dbl> in glimpse(). Is this Nominal, Ordinal, or Numeric? Why?

**Question 2.1.2:** role is stored as a character. Given there's no ranking among roles, is this Nominal or Ordinal? Why does that matter for our future analysis?

### 2.2 Numerical Summaries
**functions:**  
`summary()`: Provides the Min, Max, Median, and Mean.

**logic steps:**  
Run a `summary()` on the `season_score` column.  
Look at the Max value—if it is impossibly high, you've found a "Gremlin."


```{r}
"<write your code here>"
```

**Question 2.2.1:** What is the median and maximum season_score? Could a massive max value indicate a data quality issue or just a very good player?

## 3. Forensic Audit: Finding the Gremlins
Now we look for the quality issues we discussed: Missingness, Duplicates, and Inconsistent Casing.

### 3.1 Missing Values (NA)
**functions:**  
`is.na()`: Returns TRUE for every missing value.  
`colSums()`: Adds up all the TRUEs to give a total count of missing values per column.

**logic steps:**  
Combine `is.na(players)` with `colSums()` to get a report of all missing values in the entire table.


```{r}
"<write your code here>"
```

**Question 3.1.1:** Which variable has the most missing values? If this were the primary $CO_2$ measurement in a climate study, why would that be a problem?

### 3.2 Inconsistent Casing
**functions:**  
`count()`: Groups the rows by a column and counts how many times each label appears.  
`arrange(desc(n))`: Sorts the counts from highest to lowest.

**logic steps:**  
Take the `players` data.  
AND THEN (%>%) `count()` the `home_arena` column.  
AND THEN (%>%) `arrange()` by descending count (`n`).


```{r}
"<write your code here>"
```

**Question 3.2.1:** Do you notice inconsistent casing or spelling (e.g., neon yard vs Neon Yard)? How would this mess up a bar chart of arena popularity?

### 3.3 Duplicate IDs
**functions:**  
`duplicated()`: Flags rows that have appeared before in the dataset.

**logic steps:**  
Use `sum(duplicated())` on the `player_id` column to see if any unique identifier appears more than once.


```{r}
"<write your code here>"
```

## 4. Challenge: Spot the Issues
Look at this slice of "Raw Data" from the League's backup server.

Plaintext

| player_id | home_arena  | season_score | preferred_loadout |
|-----------|-------------|--------------|-------------------|
| P-0220    | neon yard   | 905          | Hyper Beam        |
| P-0455    | Glow Hall   | NA           | Pulse Shield      |
| P-0220    | Neon Yard   | 905          | Hyper Beam        |
| P-0512    | Photon Dome | -10          | " "               |

**Question 4.1.1:** List three distinct data quality issues from this snippet (e.g., Impossible numbers, Duplicates, Casing).

**Question 4.1.2:** Which specific R functions from this lab would you use to confirm each issue if it appeared in a massive dataset of 100,000 players?
