---
title: "Lab 01 – Data Anatomy & Type Checks"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
# Load necessary libraries (we're focusing on base R + readr/dplyr today)
# install.packages(c("readr","dplyr"), repos = "https://cloud.r-project.org")  # Uncomment + run once if needed
library(readr)   # read_csv() reads data into tibbles quickly
library(dplyr)   # glimpse(), count(), and pipes for light wrangling
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Define the root path (students should adjust if their ENIA folder lives elsewhere)
ENIA_ROOT <- "~/ENIA-5120"
```

## 1. Loading and Inspecting the Data
In Session 1 we learned that the first step is always to bring the data into R and check its shape.

### 1.1 Load Laser League Data

```{r}
league_path <- file.path("../../../data/laser_league_players.csv")
players <- read_csv(league_path)
table_preview <- head(players, 5)
table_preview
```

### 1.2 Identify the Data Structure

```{r}
players_class <- class(players)    # Shows the top-level object type (tibble/data.frame)
players_class

players_glimpse <- glimpse(players)  # Glance at each column name + type

row_count <- nrow(players)          # Count how many observations (rows) we have
col_count <- ncol(players)          # Count how many fields (columns) we have

column_names <- names(players)      # Store column names for later reference
column_names
```

**Question 1.1:** What R class is `players`? How does R internally store this table?  
**Question 1.2:** Based on `glimpse()`, how many rows/columns are present?

## 2. Understanding Data Types
Variable types determine what analysis we can perform (math vs grouping).

### 2.1 Inspect Individual Columns

```{r}
player_id_type <- class(players$player_id)[1]          # text identifier
season_score_type <- class(players$season_score)[1]    # numeric score
reaction_time_type <- class(players$reaction_time_sec)[1]  # numeric timing

player_id_type
season_score_type
reaction_time_type
```

**Question 2.1:** `assists` shows up as `<dbl>` in `glimpse()`. Is this Nominal, Ordinal, or Numeric? Why?  
**Question 2.2:** `role` is stored as character. Given there's no ranking among roles, is this Nominal or Ordinal? Why does that matter?

### 2.2 Numerical Summaries

```{r}
season_score_summary <- summary(players$season_score)  # min, median, max, etc.
season_score_summary
```

**Question 2.3:** What is the median and maximum `season_score`? Could a huge max indicate a data quality issue?

## 3. Initial Data Quality Checks

### 3.1 Missing Values

```{r}
missing_totals <- colSums(is.na(players))  # totals NA count for every column at once
missing_totals

missing_checks <- data.frame(
  column = c("season_score", "favorite_drink", "glow_rating"),
  missing_n = c(
    sum(is.na(players$season_score)),
    sum(is.na(players$favorite_drink)),
    sum(is.na(players$glow_rating))
  ),
  row.names = NULL
)
missing_checks
```

**Question 3.1:** Which variable has the most missing values? If it were critical for analysis, why is that a problem?

### 3.2 Inconsistent Categories

```{r}
home_arena_counts <- players %>%
  count(home_arena) %>%             # how many rows per arena label
  arrange(desc(n))                  # sort to scan for casing/spelling problems
home_arena_counts
```

**Question 3.2:** Do you notice inconsistent casing/spelling among `home_arena` values? If so, what kind of data issue is that?

### 3.3 Duplicate IDs

```{r}
duplicate_ids <- sum(duplicated(players$player_id))   # >0 means the same player_id appears twice
duplicate_ids
```

**Question 3.3:** If duplicates exist, what does that mean for analysis (e.g., double-counting policy events)?

## 4. Challenge: Spot the Issues

```
| player_id | home_arena  | season_score | preferred_loadout |
|-----------|-------------|--------------|-------------------|
| P-0220    | neon yard   | 905          | Hyper Beam        |
| P-0455    | Glow Hall   | NA           | Pulse Shield      |
| P-0220    | Neon Yard   | 905          | Hyper Beam        |
| P-0512    | Photon Dome | -10          | " "               |
```

**Question 4.1:** List three data quality issues from this snippet.  
**Question 4.2:** Which R functions from this lab could you use to confirm each issue if it appeared in the full dataset?
