---
title: "Lab 02: The Surgery"
subtitle: "Answer Key"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🔧 Mission Background
Last week, you audited the climate stations dataset and flagged every row with at least one issue. You found all six Gremlins hiding inside: missing values, duplicates, inconsistent casing, impossible numbers, invalid categories, and timestamps stored as text.

**Your Goal:** Fix every one of them. Unlike last week, the order of these sections matters, several fixes depend on the ones before them. Follow the sections in order.

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)

climate_path <- file.path("../../data", "climate_stations.csv")
```

---

## 0. Setup: Reload the Evidence
**functions:**
`read_csv()`: Loads the dataset.
`dim()`: Reports rows and columns.

**logic steps:**
Load `climate_stations.csv` into an object named `climate`.
Save its starting dimensions as `dim_before`.

```{r}
climate <- read_csv(climate_path)
dim_before <- dim(climate)
dim_before
```

---

## 1. Ruling Out Outliers
**functions:**
`mean()` / `sd()`, `abs()`, `filter()`

**logic steps:**
Compute the mean and SD of `temp_c` on the full dataset.
Keep only rows within 3 SDs of the mean.
Then apply the domain-knowledge bound from Lab 1 (no surface reading is physically possible outside ±60°C) as a second pass, a backstop for anything the statistical rule alone misses.

```{r}
mu <- mean(climate$temp_c, na.rm = TRUE)
sd_val <- sd(climate$temp_c, na.rm = TRUE)

removed_by_3sd <- climate %>%
  filter(abs(temp_c - mu) > 3 * sd_val) %>%
  pull(temp_c)

climate <- climate %>%
  filter(abs(temp_c - mu) <= 3 * sd_val)

removed_by_3sd
dim(climate)

# Backstop: a couple of extreme values can inflate the SD enough to hide
# each other from the 3-SD rule. The physical bound from Lab 1 catches
# what's left.
still_impossible <- climate %>%
  filter(temp_c > 60 | temp_c < -60) %>%
  pull(temp_c)

climate <- climate %>%
  filter(temp_c <= 60, temp_c >= -60)

still_impossible
dim(climate)
```

**Question 1.1:** How many rows were removed in total, and in which of the two passes? What were the removed `temp_c` values?

---

## 2. Fixing Duplicate Station-Years
**functions:**
`duplicated()` on two columns together

**logic steps:**
A real duplicate is the same `station_id` and the same `reading_date`. Keep only the first occurrence of each combination.

```{r}
n_before <- nrow(climate)

climate <- climate[!duplicated(climate[c("station_id", "reading_date")]), ]

n_before - nrow(climate)
dim(climate)
```

**Question 2.1:** How many rows were removed this time? Does that match what you'd expect, given what Section 1 already resolved?

---

## 3. Fixing Missing Values
**functions:**
`mean(na.rm = TRUE)` / `median(na.rm = TRUE)`, `coalesce()`, `mutate()`

**logic steps:**
Compare mean and median of `co2_ppm`, pick the safer one, impute.

```{r}
mean_co2 <- mean(climate$co2_ppm, na.rm = TRUE)
median_co2 <- median(climate$co2_ppm, na.rm = TRUE)

mean_co2
median_co2

climate <- climate %>%
  mutate(co2_ppm = coalesce(co2_ppm, mean_co2))

sum(is.na(climate$co2_ppm))
```

**Question 3.1:** Which did you use, mean or median, and why?

---

## 4. Fixing Inconsistent Categories
**functions:**
`str_to_lower()`, `n_distinct()`

**logic steps:**
Check distinct counts before, standardize casing, check again.

```{r}
n_distinct(climate$energy_source)
n_distinct(climate$location)

climate <- climate %>%
  mutate(
    energy_source = str_to_lower(str_trim(energy_source)),
    location = str_to_lower(str_trim(location))
  )

n_distinct(climate$energy_source)
n_distinct(climate$location)
```

**Question 4.1:** Combined across both columns, how many fake extra categories existed purely because of casing?

---

## 5. Factors: Nominal vs. Ordinal in R
**functions:**
`factor()`, `levels =`, `ordered = TRUE`, `sum(is.na())`

**logic steps:**
Convert `energy_source` to a nominal factor. Attempt `climate_risk_level` as an ordered factor and see what breaks.

```{r}
climate <- climate %>%
  mutate(energy_source = factor(energy_source))

levels(climate$energy_source)

risk_test <- factor(
  climate$climate_risk_level,
  levels = c("Low", "Medium", "High"),
  ordered = TRUE
)

sum(is.na(risk_test))
```

**Question 5.1:** How many values became `NA`? Do they match the invalid values you flagged in Lab 1?

```{r}
climate <- climate %>%
  select(-climate_risk_level)

dim(climate)
```

**Question 5.2:** In 2-3 sentences, justify your decision.

---

## 6. Fixing Broken Types

### 6.1 `sensor_active`
**functions:**
`case_when()`, `as.logical()`

```{r}
class(climate$sensor_active)

climate <- climate %>%
  mutate(sensor_active = case_when(
    sensor_active == "Yes" ~ "True",
    sensor_active == "No"  ~ "False",
    TRUE ~ sensor_active
  )) %>%
  mutate(sensor_active = as.logical(sensor_active))

class(climate$sensor_active)
```

**Question 6.1:** Confirm the class of `sensor_active` is now logical. What was it before?

### 6.2 `reading_date`
**functions:**
`case_when()`, `as.Date()`

```{r}
class(climate$reading_date)

climate <- climate %>%
  mutate(reading_date = case_when(
    reading_date == "Jan 15th, 2016" ~ "2016-01-15",
    reading_date == "Jan 15th, 2018" ~ "2018-01-15",
    reading_date == "Jan 15th, 2019" ~ "2019-01-15",
    reading_date == "Jan 15th, 2020" ~ "2020-01-15",
    reading_date == "Jan 15th, 2022" ~ "2022-01-15",
    TRUE ~ reading_date
  )) %>%
  mutate(reading_date = as.Date(reading_date))

class(climate$reading_date)
```

**Question 6.2:** Confirm the class of `reading_date` is now `Date`. What was it before?

---

## 7. Diagnosing and Fixing Skew
**functions:**
`filter()`, `log()`, `geom_histogram()`

```{r}
climate %>% filter(precipitation_mm < 0)

climate <- climate %>%
  filter(precipitation_mm >= 0)

ggplot(climate, aes(x = precipitation_mm)) +
  geom_histogram(bins = 30, fill = "#274873", color = "white") +
  labs(title = "Raw Precipitation", x = "Precipitation (mm)", y = "Count") +
  theme_minimal()

# log_precip is only needed for this one comparison plot, so it's computed
# inline rather than saved permanently into climate.
climate %>%
  mutate(log_precip = log(precipitation_mm)) %>%
  ggplot(aes(x = log_precip)) +
  geom_histogram(bins = 30, fill = "#B8842E", color = "white") +
  labs(title = "Log-Transformed Precipitation", x = "log(Precipitation)", y = "Count") +
  theme_minimal()
```

**Question 7.1:** Why didn't the 3-SD rule necessarily catch the negative values on its own? Compare the two histograms, did the shape move closer to normal?

---

## 8. Visual Confirmation

### 8.1 Histogram: `temp_c`
```{r}
ggplot(climate, aes(x = temp_c)) +
  geom_histogram(bins = 30, fill = "#274873", color = "white") +
  labs(title = "Temperature Readings (Outliers Removed)", x = "Temp (C)", y = "Count") +
  theme_minimal()
```

### 8.2 Bar Chart: `energy_source`
```{r}
climate %>%
  count(energy_source) %>%
  ggplot(aes(x = reorder(energy_source, -n), y = n)) +
  geom_col(fill = "#B8842E") +
  labs(title = "Energy Source Counts (Casing Fixed)", x = "Energy Source", y = "Count") +
  theme_minimal()
```

### 8.3 Line Chart: `temp_c` Over Time
```{r}
climate %>%
  filter(station_id == "STN-002") %>%
  arrange(reading_date) %>%
  ggplot(aes(x = reading_date, y = temp_c)) +
  geom_line(color = "#274873") +
  geom_point() +
  labs(title = "Temperature Over Time, STN-002", x = "Reading Date", y = "Temp (C)") +
  theme_minimal()
```

**Question 8.1:** What would have gone wrong with this line chart if you'd tried it before fixing `reading_date`'s type in Section 6?

---

## 9. Write the Cleaning Report
```{r}
dim_before
dim(climate)

climate_clean <- climate
```

**Question 9.1:** In 4 to 6 sentences, summarize your cleaning pass. Compare `dim_before` to your final dimensions, note which fix you think mattered most, and explain why a clean dataset matters for what comes next (joins, in Session 3).
