---
title: "Lab 02: The Surgery"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🔧 Mission Background
Last week, you audited the climate stations dataset and flagged every row with at least one issue. You found all six Gremlins hiding inside: missing values, duplicates, inconsistent casing, impossible numbers, invalid categories, and timestamps stored as text.

**Your Goal:** Fix every one of them. Unlike last week, the order of these sections matters, several fixes depend on the ones before them. Follow the sections in order.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# readr: fast, smart data loading.
# dplyr: filtering, mutating, and summarizing.
# stringr: cleaning up text columns.
# ggplot2: visual confirmation at the end.

library(readr)
library(dplyr)
library(stringr)
library(ggplot2)

# Define the root path (adjust if your folder structure is different)
climate_path <- file.path("../../data", "climate_stations.csv")
```

---

## 0. Setup: Reload the Evidence
**functions:**
`read_csv()`: Loads the dataset.
`dim()`: Reports rows and columns.

**logic steps:**
Load `climate_stations.csv` into an object named `climate`.
Save its starting dimensions as `dim_before`, before any fixes happen. You'll compare against this number in Section 9.

```{r}
"<write your code here>"
```

---

## 1. Ruling Out Outliers
We use the 3-SD rule from lecture to catch physically impossible temperature readings.

**functions:**
`mean()` / `sd()`: Center and spread of `temp_c`.
`abs()`: Distance from the mean, regardless of direction.
`filter()`: Keeps only rows that pass a condition.

**logic steps:**
Compute the mean and SD of `temp_c` on the full dataset.
Keep only rows where the temperature is within 3 SDs of the mean.
Then, as a backstop, apply the physical bound from Lab 1 (no real surface reading falls outside ±60°C). A couple of very extreme values can inflate the SD enough to hide themselves, or each other, from the 3-SD rule, so this second pass catches whatever the statistical check alone missed.
Overwrite `climate` with the result after each pass.

```{r}
"<write your code here>"
```

**Question 1.1:** How many rows were removed in total, and in which of the two passes? What were the removed `temp_c` values?

```
<write your answer here>
```

---

## 2. Fixing Duplicate Station-Years
**functions:**
`duplicated()`: Flags repeat rows. Can check two columns together instead of one.

**logic steps:**
Reminder from Lab 1: a real duplicate here means the same `station_id` *and* the same `reading_date`, not `station_id` alone.
Use `duplicated()` on `station_id` and `reading_date` together to flag repeats.
Keep only the rows where that check is `FALSE`.

```{r}
"<write your code here>"
```

**Question 2.1:** How many rows were removed this time? Does that match what you'd expect, given what Section 1 already resolved?

```
<write your answer here>
```

---

## 3. Fixing Missing Values
**functions:**
`mean(na.rm = TRUE)` / `median(na.rm = TRUE)`: Center of `co2_ppm`, ignoring gaps.
`coalesce()`: Fills a missing value with a backup value.
`mutate()`: Adds or replaces a column.

**logic steps:**
Compute the mean and median of `co2_ppm`.
Decide which is the safer choice to impute with.
Fill missing values using `coalesce()`.

```{r}
"<write your code here>"
```

**Question 3.1:** Which did you use, mean or median, and why?

```
<write your answer here>
```

---

## 4. Fixing Inconsistent Categories
**functions:**
`str_to_lower()`: Forces every value in a column to lowercase.
`n_distinct()`: Counts how many different values exist in a column.

**logic steps:**
Check `n_distinct()` on `energy_source` and `location` before fixing.
Standardize the casing on both columns.
Check `n_distinct()` again to confirm the count dropped.

```{r}
"<write your code here>"
```

**Question 4.1:** Combined across both columns, how many fake extra categories existed purely because of casing?

```
<write your answer here>
```

---

## 5. Factors: Nominal vs. Ordinal in R
**functions:**
`factor()`: Converts a column into a categorical factor.
`levels =`, `ordered = TRUE`: Defines the valid categories, and whether they have a real order.
`sum(is.na())`: Counts how many values became missing.

**logic steps:**
Convert `energy_source` into a nominal factor, no order needed.
Attempt to convert `climate_risk_level` into an ordered factor with levels `"Low"`, `"Medium"`, `"High"`.
Count how many values became `NA` as a result.

```{r}
"<write your code here>"
```

**Question 5.1:** How many values became `NA`? Do they match the invalid values you flagged in Lab 1?

```
<write your answer here>
```

**logic steps (continued):**
Given how few values are salvageable, decide whether `climate_risk_level` is worth recoding or better off dropped entirely. Drop the column if that's your call.

```{r}
"<write your code here>"
```

**Question 5.2:** In 2-3 sentences, justify your decision.

```
<write your answer here>
```

---

## 6. Fixing Broken Types

### 6.1 `sensor_active`
**functions:**
`case_when()`: Recodes values based on conditions, checked top to bottom.
`as.logical()`: Converts a column to `TRUE`/`FALSE`.

**logic steps:**
Recode `"Yes"` to `"True"` and `"No"` to `"False"`, leaving everything else as is.
Convert the column to logical.

```{r}
"<write your code here>"
```

**Question 6.1:** Confirm the class of `sensor_active` is now logical. What was it before?

```
<write your answer here>
```

### 6.2 `reading_date`
**functions:**
`case_when()`: Swaps each known bad string for its correct value.
`as.Date()`: Converts a column to a real date type.

**logic steps:**
Swap each of the 5 known bad date strings (see the workbook table) for its correct ISO-format equivalent.
Convert the column with `as.Date()`.

```{r}
"<write your code here>"
```

**Question 6.2:** Confirm the class of `reading_date` is now `Date`. What was it before?

```
<write your answer here>
```

---

## 7. Diagnosing and Fixing Skew
**functions:**
`filter()`: Removes impossible values before transforming.
`log()`: Compresses a long right tail.
`geom_histogram()`: Shows distribution shape.

**logic steps:**
`precipitation_mm` has a long right tail, but first, filter out any negative values, rainfall can't be negative, and the 3-SD rule alone won't necessarily catch it.
Plot the raw distribution.
For the log-transformed plot, compute `log(precipitation_mm)` inline rather than saving it as a permanent column, it's only needed for this one comparison.

```{r}
"<write your code here>"
```

**Question 7.1:** Why didn't the 3-SD rule necessarily catch the negative values on its own? Compare the two histograms, did the shape move closer to normal?

```
<write your answer here>
```

---

## 8. Visual Confirmation
**functions:**
`ggplot()`, `aes()`: Set up the data and what goes on which axis.
`geom_histogram()`: Numeric shape.
`geom_col()`: Category counts.
`geom_line()`: Change over time.
`labs()`, `theme_minimal()`: Titles and a clean look.

### 8.1 Histogram: `temp_c`
Confirm the outliers are gone.

```{r}
"<write your code here>"
```

### 8.2 Bar Chart: `energy_source`
Confirm there's no longer a split between two versions of the same category.

```{r}
"<write your code here>"
```

### 8.3 Line Chart: `temp_c` Over Time
Filter to one `station_id` of your choosing, and plot its temperature across `reading_date`.

```{r}
"<write your code here>"
```

**Question 8.1:** What would have gone wrong with this line chart if you'd tried it before fixing `reading_date`'s type in Section 6?

```
<write your answer here>
```

---

## 9. Write the Cleaning Report
No new functions here. Just synthesize what you did.

```{r}
"<write your code here>"
```

**Question 9.1:** In 4 to 6 sentences, summarize your cleaning pass. Compare `dim_before` to your final dimensions, note which fix you think mattered most, and explain why a clean dataset matters for what comes next (joins, in Session 3).

```
<write your answer here>
```
