---
title: "Session 2: The Analytical Detective"
subtitle: "Moving from Cleaning to Discovery"
format: 
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true    
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

## You cleaned your dataset - Now What? 🕵️

* You have scrubbed the typos and filled the **NA** values.
* But clean data is just a silent spreadsheet.
* **Exploratory Data Analysis (EDA)** is the process of "interviewing" your data to see what stories it is trying to tell you.
* **The Goal:** Understand the **landscape** of the data to avoid building analysis on a flawed foundation.

---

## Setup Check (R Packages) 🧰

```{r, label="r_setup", echo=TRUE, eval=FALSE}
library(jsonlite)
library(rmarkdown)
library(knitr)
library(ggplot2)
```

---

## The Transition: Why EDA? 🧭

* **Cleaning** asked: *Is the data broken?*
* **EDA** asks: *What is the data's personality?*
* **The Why:** You cannot build ML models on top of what you don't know. You have to understand the data before you jump into "predict" something.
* It prevents you from applying a **Normal** model to a **Skewed** world.

---

## The Pre-Analysis Audit ✅
### What do we check before analysis?

* **Balance:** Are our categories representative, or is 90% of our data from one source? (Critical for policy equity).
* **Feasibility:** Do the ranges make sense? (e.g., Does the data display a state of "normalcy"? Or is it showing a skewed nature of a problem?).
* **Anomalies:** Extreme "impossible" values that survived the first cleaning pass.

---

### Outliers 🎯

* **The Why:** If the mean (red) is far from the median (green), outliers are hijacking your summary.
* **The What:** However, it's always important to consider: "Does this represent the reality?"
* **Resolution:** Verify errors, report median, or cap extremes.

----

```{r, echo=TRUE, eval=TRUE}
library(ggplot2)
set.seed(42)
df <- data.frame(emissions = rgamma(1000, shape = 2, scale = 5000))

mu <- mean(df$emissions, na.rm = TRUE)
med <- median(df$emissions, na.rm = TRUE)

ggplot(df, aes(x = emissions)) +
  geom_histogram(bins = 40, fill = "#959fe847", color = "white") +
  geom_vline(xintercept = mu, color = "#B71C1C", linewidth = 1.4) +
  geom_vline(xintercept = med, color = "#1B5E20", linewidth = 1.4) +
  labs(
    title = "Outliers shift the mean away from the median",
    subtitle = "Green = Median (robust), Red = Mean (sensitive)",
    x = "Annual CO2 emissions",
    y = "Count"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

----

### How do we know if it's an outlier? 📏

* **Rule of thumb:** A point more than **3 standard deviations** from the mean is often flagged as an outlier.
* **Why “normal” matters:** The 3‑SD rule only makes sense when data is **bell‑shaped**. In a normal distribution, almost everything falls near the center, so 3‑SD points are truly rare.
* **If data is skewed:** Long tails are expected, so the 3‑SD rule can label too many “outliers.” Use visuals or robust methods instead.

---

#### Example: Normal Distribution of Student Grades 📊

```{r, echo=TRUE, eval=TRUE}
set.seed(42)
grades <- pmin(pmax(rnorm(300, mean = 80, sd = 8), 0), 100)

ggplot(data.frame(grades = grades), aes(x = grades)) +
  geom_histogram(bins = 20, fill = "#2a9d8f", color = "white") +
  labs(
    title = "Student Grades (Approximately Normal)",
    x = "Score",
    y = "Number of Students"
  ) +
  theme_minimal()
```

---

### Normal Distribution 🔔

* What does **typical** mean to you?
* In real world settings, the intuition around **typical** is built around **average**.
* So if something were to be **normal**, most things would be around the average right?
* That's what Normal Distribution shows. 
* What if you are comparing 2 different things? And one is typical and one is not? 
* Because Normal Distribution shows "the usual", it allows for fair comparison, and therefore is commonly used in data science settings. 

---

### Data Skewness 📉

* **The Why:** Skewed data breaks "normal" assumptions.
* **Resolution:** Log transforms turn huge differences into a readable scale.

```{r, echo=TRUE, eval=TRUE}
set.seed(42)
df <- data.frame(emissions = rlnorm(1000, meanlog = 8, sdlog = 0.7))

df_long <- data.frame(
  value = c(df$emissions, log(df$emissions)),
  scale = rep(c("Raw CO2", "Log(CO2)"), each = nrow(df))
)

ggplot(df_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#43A047", color = "white") +
  facet_wrap(~ scale, scales = "free_x") +
  labs(
    title = "Log transformation reduces skew in emissions",
    subtitle = "Log makes the tail readable",
    x = "CO2 emissions",
    y = "Count"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

---

### What is transformation? 🔍

* Imagine you are looking at a city map while standing next to a skyscraper. 
* The skyscraper fills up your view right? 
* Now, let's zoom out. If you zoom out, tall buildings are still taller than shorter ones, but you can see the whole map now. 
* Transformation, although a lot of other usage, mainly is used in data analysis to "right size" values that are not in proportionate scale. 

---

### Class Imbalance (The Majority Bias) ⚖️

* **The Why:** If 95% of cases are one class, the "average" just describes the majority.
* **Risk:** The minority becomes invisible in your conclusions.
* **Resolution:** Emphasize **precision** and **recall**, not just accuracy. 

---

### Precision and Recall 🎯
* Imagine 1000 transactions and only 10 are fraud
* If we build a model that says "not fraud" everytime, how accurate is it?
* Is this a good model?
* **Accuracy** measures how well we predictg the majority case, not whether we catch what we care about, which is the 1%
* **Precision**: When the model flags something as fraud, how often is it actually fraud?
* **Recall**: Out of all the real fraud cases, how many did we actually catch?

---

## Univariate Health Check 🩺
### Why do these summaries matter? 🧠

* **Mean vs. Median:** Where is the true "center" of experience?
* **Standard Deviation:** How risky or unstable is the outcome?
* **Percentiles:** Who is being helped or harmed at the edges?

----

### Mean vs. Median ⚖️

* **Mean (Average):** The *center of gravity*. Uses every data point.
  * **Why:** Efficient when data is symmetrical, but fragile.
* **Median (50th percentile):** The *common-sense center*.
  * **Why:** Robust to outliers; often the truer policy story.

----

### Standard Deviation (SD) 📏

* **Concept:** Average distance of points from the mean.
* **Why it matters:** SD is your **uncertainty meter**.
  * **Low SD:** The average is reliable.
  * **High SD:** The average is a gamble. Outcomes are inconsistent.

----

### Percentiles 🧩

* **Concept:** Slicing the data into 100 buckets.
* **Why it matters:** Policy happens at the edges.
  * **10th percentile:** Who is most vulnerable?
  * **90th/99th:** Who benefits most, or who should be taxed?

---

## Relationships between Variables 💃🕺
### Numeric vs. Numeric 🔗

* **Why:** Is Variable A a lever for Variable B? In quantifiable amount, how much does Var B impact Var A?
* **Interpretation:** Slope is the `amount`, constant is the `default`

----

$\text{Air Quality Index} = 0.5 \times \text{Tree Cover} + 10$

```{r, echo=TRUE, eval=TRUE}
set.seed(42)
df <- data.frame(
  TreeCover = runif(120, 0, 100)
)
df$AirQuality <- 0.5 * df$TreeCover + 10 + rnorm(120, 0, 8)

ggplot(df, aes(x = TreeCover, y = AirQuality)) +
  geom_point(alpha = 0.7, color = "#1E88E5") +
  geom_smooth(method = "lm", se = FALSE, color = "#2E7D32", linewidth = 1) +
  labs(
    title = "Tree cover vs. air quality",
    subtitle = "Slope shows leverage",
    x = "Tree cover (%)",
    y = "Air quality index"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

----

### Numeric vs. Categorical 📦

* **Why:** Does structural difference have impact on performance?
* **Clue:** If box middles don't overlap, groups behave differently. (Mean is different!)

----

```{r, echo=TRUE, eval=TRUE}
set.seed(42)
df <- data.frame(
  Region = rep(c("Coastal", "Inland", "Mountain"), each = 60),
  WaterQuality = c(
    rnorm(60, mean = 70, sd = 8),
    rnorm(60, mean = 78, sd = 7),
    rnorm(60, mean = 85, sd = 6)
  )
)

ggplot(df, aes(x = Region, y = WaterQuality, fill = Region)) +
  geom_boxplot(width = 0.6, alpha = 0.85) +
  scale_fill_viridis_d(option = "C", end = 0.85) +
  labs(
    title = "Water quality by region",
    subtitle = "Non-overlap suggests structural differences",
    x = "Region",
    y = "Water quality score"
  ) +
  theme_minimal() +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

----

### Multicollinearity (The Redundancy Trap) 🔁

* **What it is:** Two predictors tell the same story.
* **Why it matters:** The model cannot tell which variable is the real driver.
* **Detection:** Look for hot spots in the correlation heatmap (|r| > 0.8).

----

```{r, echo=TRUE, eval=TRUE}
set.seed(42)
df <- data.frame(
  Energy_Use = rnorm(200, mean = 100, sd = 15)
)
# Create strongly related predictors
df$CO2 <- 0.9 * df$Energy_Use + rnorm(200, 0, 6)
df$Industry_Output <- 0.8 * df$Energy_Use + rnorm(200, 0, 7)
df$Vehicle_Miles <- 0.7 * df$Energy_Use + rnorm(200, 0, 8)
df$Electricity_Use <- 0.85 * df$Energy_Use + rnorm(200, 0, 5)
df$Renewables <- rnorm(200, mean = 30, sd = 8)

cor_mat <- round(cor(df), 2)
cor_long <- as.data.frame(as.table(cor_mat))

colnames(cor_long) <- c("Var1", "Var2", "Correlation")

ggplot(cor_long, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#3B4CC0", mid = "white", high = "#eede49ff",
    midpoint = 0, limits = c(-1, 1)
  ) +
  geom_text(aes(label = Correlation), size = 3) +
  labs(
    title = "Correlation matrix (multicollinearity)",
    subtitle = "Hot spots show redundant predictors"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

---

## The Pivot to Visuals 👀
### Why use ggplot? 👓

* Summary stats can be identical for datasets that look completely different.
* Visuals reveal **clusters**, **gaps**, and **curves** that tables hide.
* If you do not visualize, you are flying blind.

---

## Visual Design 🎨
### What Makes a Good Visual? ✨

* **The Issue of Scale:** Bar charts must start at zero. Non-zero baselines exaggerate small differences to mislead viewers.
* **Color Blindness:** ~8% of men have red-green color blindness.
* **The Fix:** Use colorblind-safe palettes like **viridis** or **cividis**.

----

### Example: Cluttered & Misleading 🚧

```{r, echo=TRUE, eval=TRUE}
df <- data.frame(
  Region = factor(c("A", "B", "C", "D"), levels = c("A", "B", "C", "D")),
  ProtectedLand = c(68, 71, 70, 69),
  Budget = c(12.4, 12.6, 12.5, 12.7)
)

# Distortion: truncated view + cluttered labels + color-blind-unfriendly palette
bad_plot <- ggplot(df, aes(x = Region, y = ProtectedLand, fill = Region)) +
  geom_col(width = 0.6) +
  coord_cartesian(ylim = c(50, 72)) +
  scale_fill_manual(values = c("#E41A1C", "#E41A1C", "#E41A1C", "#A65628")) +
  geom_text(
    aes(label = paste0(
      "Protected land: ", ProtectedLand, "%\n",
      "Budget: $", Budget, "B\n",
      "Policy score: ", ProtectedLand + 12
    )),
    vjust = -0.2,
    size = 3
  ) +
  labs(
    title = "Protected land by region",
    subtitle = "Crowded labels + truncated axis",
    x = "Region",
    y = "Protected land (%)"
  ) +
  theme_bw() +
  theme(panel.grid.major = element_line(color = "gray40", size = 0.6))

bad_plot
```

----

### Example: Clear & Accessible ✅

```{r, echo=TRUE, eval=TRUE}
# Clean scale + accessible palette + clear narrative

good_plot <- ggplot(df, aes(x = Region, y = ProtectedLand, fill = Region)) +
  geom_col(width = 0.6) +
  scale_fill_viridis_d(option = "C", end = 0.85) +
  scale_y_continuous(limits = c(0, 80)) +
  labs(
    title = "Protected land is similar across regions",
    subtitle = "All regions cluster around ~70% coverage",
    x = "Region",
    y = "Protected land (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

good_plot
```

---

## ggplot2 Nomenclature 🧩
### The Grammar of Graphics 🧩

| Component | Nomenclature | Description |
| :--- | :--- | :--- |
| Data | `data = df` | The raw spreadsheet. |
| Aesthetics | `aes()` | Mapping columns to visual properties ($X, Y,$ Color). |
| Geometries | `geom_...()` | The "Shape" (Point, Bar, Boxplot). |
| Facets | `facet_wrap()` | Splitting one chart into many by category. |
| Themes | `theme_...()` | The styling (fonts, backgrounds). |
