---
title: "HW02 Model Answer – Advanced Data Visualization Storytelling"
author: "Instructor"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Optional package for bonus interactivity
has_plotly <- requireNamespace("plotly", quietly = TRUE)
```

# 0. Load, Clean, and Join Data

```{r load-clean-join}
sites_raw <- read_csv("../../../../data/air_quality_sites.csv")
policy_raw <- read_csv("../../../../data/policy_actions.csv")

sites <- sites_raw %>%
  mutate(
    site_id = str_trim(str_to_upper(site_id)),
    region = str_trim(str_to_title(region))
  )

policy <- policy_raw %>%
  mutate(
    site_id = str_trim(str_to_upper(site_id)),
    policy_strength = factor(str_trim(str_to_title(policy_strength)), levels = c("None", "Pilot", "Partial", "Full"), ordered = TRUE),
    policy_type = str_trim(str_to_title(policy_type)),
    policy_adopted = as.character(policy_adopted),
    grant_amount = as.numeric(grant_amount),
    compliance_score = as.numeric(compliance_score),
    estimated_reduction_pct = as.numeric(estimated_reduction_pct)
  )

policy_site <- policy %>%
  filter(!is.na(policy_strength), !is.na(policy_type)) %>%
  arrange(site_id, start_date) %>%
  group_by(site_id) %>%
  summarize(
    policy_strength = max(policy_strength),
    policy_type = first(policy_type),
    grant_amount = sum(grant_amount, na.rm = TRUE),
    compliance_score = mean(compliance_score, na.rm = TRUE),
    estimated_reduction_pct = mean(estimated_reduction_pct, na.rm = TRUE),
    policy_adopted = as.character(any(policy_adopted == "TRUE" | policy_adopted == "True")),
    .groups = "drop"
  )

analysis_df <- left_join(sites, policy_site, by = "site_id") %>%
  filter(
    !is.na(pm25),
    !is.na(policy_strength),
    !is.na(traffic_index),
    !is.na(compliance_score),
    !is.na(estimated_reduction_pct),
    !is.na(grant_amount)
  )

glimpse(analysis_df)
```

# Q1 (Violin Plot)

**Question this chart answers:** How does PM2.5 distribution differ across policy strength levels?

```{r q1-violin, fig.width=8, fig.height=5}
p1 <- analysis_df %>%
  ggplot(aes(x = policy_strength, y = pm25, fill = policy_strength)) +
  geom_violin(alpha = 0.75, trim = FALSE, color = "gray35") +
  stat_summary(fun = median, geom = "point", color = "black", size = 2) +
  labs(
    title = "PM2.5 Distribution by Policy Strength",
    x = "Policy Strength",
    y = "PM2.5",
    fill = "Policy Strength"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

p1
```

**Why this visual:** A violin plot shows distribution shape, spread, and center in one chart.

**What it shows:** PM2.5 distributions differ by policy strength. The median markers help compare central tendency while the violin width shows where values are concentrated.

**Caveat:** This is descriptive and does not establish a causal policy effect.

# Q2 (Facet Wrap)

**Question this chart answers:** Does the traffic-PM2.5 relationship look different across policy types?

```{r q2-facet-wrap, fig.width=10, fig.height=6}
p2 <- analysis_df %>%
  ggplot(aes(x = traffic_index, y = pm25, color = policy_strength)) +
  geom_point(alpha = 0.45, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~policy_type) +
  labs(
    title = "Traffic vs PM2.5 by Policy Type",
    x = "Traffic Index",
    y = "PM2.5",
    color = "Policy Strength"
  ) +
  theme_minimal(base_size = 12)

p2
```

**Why this visual:** Faceting isolates subgroup patterns without mixing them into one panel.

**What it shows:** The traffic-pollution relationship is positive across most panels, but slope/dispersion vary by policy type.

**Caveat:** Differences across facets can reflect sample composition, not only policy dynamics.

# Q3 (Bubble Scatter Plot)

**Question this chart answers:** Which site-policy combinations look high priority for funding and reduction potential?

```{r q3-bubble, fig.width=8, fig.height=5}
p3 <- analysis_df %>%
  ggplot(
    aes(
      x = compliance_score,
      y = estimated_reduction_pct,
      size = grant_amount,
      color = policy_adopted
    )
  ) +
  geom_point(alpha = 0.65) +
  scale_size_continuous(range = c(1.5, 10)) +
  labs(
    title = "Policy Reduction Potential vs Compliance (Bubble Size = Grant Amount)",
    x = "Compliance Score",
    y = "Estimated Reduction (%)",
    size = "Grant Amount",
    color = "Policy Adopted"
  ) +
  theme_minimal(base_size = 12)

p3
```

**Why this visual:** Bubble scatter allows 4 dimensions in one view (x, y, size, color).

**What it shows:** Larger grants and higher reduction potential are visible simultaneously; adopted vs non-adopted clusters can be compared directly.

**Caveat:** Bubble size can dominate attention, so scale choices affect interpretation.

# Q4 (Ridgeline Density Plot)

**Question this chart answers:** How does the full PM2.5 distribution shift across regions?

```{r q4-ridgeline, fig.width=9, fig.height=5}
if (requireNamespace("ggridges", quietly = TRUE)) {
  p4 <- analysis_df %>%
    ggplot(aes(x = pm25, y = region, fill = region)) +
    ggridges::geom_density_ridges(alpha = 0.7, scale = 1.1, color = "gray20") +
    labs(
      title = "Ridgeline Density of PM2.5 by Region",
      x = "PM2.5",
      y = "Region"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")

  p4
} else {
  ggplot(analysis_df, aes(x = pm25, fill = region)) +
    geom_density(alpha = 0.35) +
    facet_wrap(~region, ncol = 1) +
    labs(
      title = "Fallback View (Install ggridges for Ridgeline): PM2.5 Density by Region",
      x = "PM2.5",
      y = "Density"
    ) +
    theme_minimal(base_size = 12)
}
```

**Why this visual:** Ridgeline plots compare multiple full distributions compactly.

**What it shows:** Regions differ not only in mean PM2.5 but in spread and tail behavior.

**Caveat:** Requires `ggridges`; fallback density facets are less compact.

# Final Summary

If briefing leadership: (1) distribution differences suggest uneven exposure across policy contexts, and (2) the bubble plot reveals where high-reduction opportunities align with funding and compliance. Main caution: these visuals are associational; they should guide prioritization, not claim causal impact by themselves.
