---
title: "Lab 02 – Data Cleaning Strategies and Exploratory Data Analysis (EDA)"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

## 🧭 Mission Background
You are the lead analyst for a public policy team reviewing global fishing effort. Before the team makes policy recommendations, you must inspect the data, diagnose missingness, and test how different imputation strategies change the story.

**Your Goal:** Use EDA to understand the shape of the data and apply deletion, mean/median imputation, and mode imputation to handle missing values.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads external packages into your session.
# readr: Part of the Tidyverse, used for fast and "smart" data loading.
# dplyr: Provides the tools for inspection and manipulation.

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)

# Define the root path (Adjust if your folder structure is different)
ENIA_ROOT <- "~/ENIA-5120"
efforts_path <- file.path("../../../data", "gfw_outputs", "efforts_by_flag_gear.csv")
```

## 1. The Evidence Intake: Load & Inspect
We start by pulling the dataset into R and confirming its structure.

### 1.1 Load the Data
**functions:**  
`read_csv()`: Reads data into R quickly and reliably.  
`head()`: Displays the first 6 rows for a quick sanity check.

**logic steps:**  
Locate the file path in `efforts_path`.  
Load the CSV into a data frame named `df`.  
Print the first rows to confirm columns and values look reasonable.

```{r}
df <- read_csv(efforts_path)
head(df)
```

### 1.2 Inspect the Structure
**functions:**  
`class()`: Tells you the object type (e.g., tibble/data.frame).  
`glimpse()`: Gives a vertical view of columns and data types.  
`nrow()` / `ncol()`: Counts rows and columns.

**logic steps:**  
Confirm `df` is a data frame/tibble.  
Scan column names and data types with `glimpse()`.  
Record row and column counts to know dataset size.

```{r}
class(df)
glimpse(df)

nrow(df)
ncol(df)
```

### 1.3 Basic Statistical Summary
**functions:**  
`summary()`: Provides min, max, mean, and quartiles for numeric columns.  
`select(where(is.numeric))`: Keeps only numeric variables.

**logic steps:**  
Keep only numeric columns to avoid text summaries.  
Run `summary()` to inspect min, max, and quartiles.  
Use this to spot unusually large or small values.

```{r}
numeric_summary <- summary(df %>% select(where(is.numeric)))
numeric_summary
```

## 2. Missingness + Balance
We check missingness and whether categories are balanced.

### 2.1 Missingness by Column
**functions:**  
`is.na()`: Checks every cell; returns TRUE if a value is missing (NA).  
`colSums()`: Adds up TRUE values to count missingness by column.

**logic steps:**  
Use `is.na()` inside `colSums()` to count missing values by column.

```{r}
missing_totals <- colSums(is.na(df))
missing_totals
```

**Question 2.1.1:** Which column has the most missing values?

### 2.2 Class Imbalance Check
**functions:**  
`count()`: Tallies how many observations fall in each category.  
`arrange(desc(n))`: Sorts counts from highest to lowest.  
`filter(!is.na())`: Drops missing categories when needed.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to visual channels.  
`geom_col()`: Draws bars with heights from data.  
`labs()`: Adds plot titles and axis labels.  
`theme_minimal()`: Applies a clean theme.  
`theme()`: Fine-tunes text and layout.  
`element_text()`: Styles axis labels.
`reorder()`: Orders categories by a numeric value.

**logic steps:**  
Remove missing flags for a fair comparison table.  
Count records per `flag` and per `geartype`.  
Sort counts to see which categories dominate.  
Plot each count table as a bar chart for visual balance.

```{r}
df_noflag <- df %>%
  filter(!is.na(flag))

flag_counts <- df_noflag %>%
  count(flag) %>%
  arrange(desc(n))

geartype_counts <- df %>%
  count(geartype) %>%
  arrange(desc(n))

# Base R fallback (uncomment if dplyr::count() fails)
# flag_counts <- as.data.frame(sort(table(df_noflag$flag), decreasing = TRUE))
# names(flag_counts) <- c("n", "flag")
# flag_counts <- flag_counts[, c("flag", "n")]
# geartype_counts <- as.data.frame(sort(table(df$geartype), decreasing = TRUE))
# names(geartype_counts) <- c("n", "geartype")
# geartype_counts <- geartype_counts[, c("geartype", "n")]

flag_counts
geartype_counts

flag_bar_plot <- ggplot(flag_counts, aes(x = reorder(flag, -n), y = n)) +
  geom_col(fill = "#0072B2") +
  labs(title = "Flag Counts", x = "Flag", y = "Records") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

gear_bar_plot <- ggplot(geartype_counts, aes(x = reorder(geartype, -n), y = n)) +
  geom_col(fill = "#009E73") +
  labs(title = "Gear Type Counts", x = "Gear Type", y = "Records") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

flag_bar_plot
gear_bar_plot
```

**Question 2.2.1:** Is the dataset balanced across flags, or does one flag dominate?

**Question 2.2.2:** Why can class imbalance distort policy conclusions?

## 3. Imputation Comes First
We will use **imputed data moving forward**.

### 3.1 Mean vs. Median (Numeric)
**functions:**  
`filter(!is.na(x))`: Deletion tool that drops missing rows.  
`mean()` / `median()`: Measures central tendency; `na.rm = TRUE` ignores NAs.  
`coalesce()`: Imputation tool that fills NAs with a replacement value.  
`sd()`: Calculates Standard Deviation (spread).
`mutate()`: Adds or replaces a column in the data frame.

**logic steps:**  
Create a deletion-only version by removing missing `crew_size`.  
Compute the mean and median to see if they diverge.  
Create two imputed versions using mean and median.  
Compare resulting mean and SD across strategies to judge impact.

```{r}
# Strategy A: Deletion
df_deleted <- df %>% filter(!is.na(crew_size))

# Strategy B: Mean imputation
mean_val <- mean(df$crew_size, na.rm = TRUE)
df_mean_imp <- df %>%
  mutate(crew_size = coalesce(crew_size, mean_val))

# Strategy C: Median imputation
median_val <- median(df$crew_size, na.rm = TRUE)
df_median_imp <- df %>%
  mutate(crew_size = coalesce(crew_size, median_val))

comparison <- data.frame(
  Strategy = c("Deletion", "Mean Imputation", "Median Imputation"),
  Resulting_Mean = c(
    mean(df_deleted$crew_size),
    mean(df_mean_imp$crew_size),
    mean(df_median_imp$crew_size)
  ),
  Resulting_SD = c(
    sd(df_deleted$crew_size),
    sd(df_mean_imp$crew_size),
    sd(df_median_imp$crew_size)
  )
)
comparison
```

**Question 3.1.1:** Which strategy produced the lowest SD (variability)?

**Question 3.1.2:** Why might lower variability after imputation be misleading?

**Question 3.1.3:** When the mean and median differ, which is safer to use and why?

### 3.2 Mode Imputation (Categorical)
**functions:**  
`count()`: Counts category frequencies.  
`arrange(desc(n))`: Sorts from most common to least.  
`slice(1)`: Selects the top row (the mode).  
`pull()`: Extracts a single value from a table.  
`coalesce()`: Fills missing values with the replacement.
`filter(!is.na())`: Removes missing categories before counting.  
`mutate()`: Replaces missing values in a column.  
`table()`: Base R frequency check for before/after.

**logic steps:**  
Filter to non-missing flags for a clean frequency count.  
Identify the most frequent `flag` (the mode).  
Impute missing flags using the mode.  
Compare the mode frequency before and after imputation.

```{r}
mode_flag <- df %>%
  filter(!is.na(flag)) %>%
  count(flag) %>%
  arrange(desc(n)) %>%
  slice(1) %>%
  pull(flag)

df_mode_imp <- df_median_imp %>%
  mutate(flag = coalesce(flag, mode_flag))

# Compare frequency of the mode
# Before:
table(df$flag)[mode_flag]
# After:
table(df_mode_imp$flag)[mode_flag]
```

**Question 3.2.1:** By how many observations did the mode category increase after imputation?

**Question 3.2.2:** Why can mode imputation reduce diversity in categorical data?

### 3.3 Imputed Dataset for Analysis
```{r}
df_imputed <- df_mode_imp
```

## 4. Outliers: 3 Standard Deviations Away
We flag extreme values using the 3-SD rule.

**functions:**  
`mean()` / `sd()`: Compute center and spread.  
`abs()`: Calculates distance from the mean.
`filter()`: Keeps only rows that meet a condition.

**logic steps:**  
Compute the mean and SD of `apparent_fishing_hours`.  
Measure each value’s distance from the mean.  
Flag rows whose distance exceeds 3 SD.  
Count how many rows are flagged.

```{r}
hours_mean <- mean(df_imputed$apparent_fishing_hours, na.rm = TRUE)
hours_sd <- sd(df_imputed$apparent_fishing_hours, na.rm = TRUE)

hours_outliers <- df_imputed %>%
  filter(abs(apparent_fishing_hours - hours_mean) > 3 * hours_sd)

nrow(hours_outliers)
```

**Question 4.1.1:** How many rows are flagged as outliers by the 3-SD rule?

**Question 4.1.2:** Why can the 3-SD rule be misleading if the data is skewed?

### 4.2 Overall Boxplot: With vs. Without Outliers
**functions:**  
`geom_boxplot()`: Highlights median, quartiles, and outliers.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Create a trimmed dataset by removing 3-SD outliers.  
Plot the full distribution to show raw spread.  
Plot the trimmed distribution to show the typical range.

```{r}
df_trim <- df_imputed %>%
  filter(abs(apparent_fishing_hours - hours_mean) <= 3 * hours_sd)

box_overall <- ggplot(df_imputed, aes(x = "All", y = apparent_fishing_hours)) +
  geom_boxplot(fill = "#56B4E9") +
  labs(title = "Overall Fishing Hours", x = "", y = "Fishing Hours") +
  theme_minimal()

box_trim <- ggplot(df_trim, aes(x = "Trimmed", y = apparent_fishing_hours)) +
  geom_boxplot(fill = "#009E73") +
  labs(title = "Overall Fishing Hours (Trimmed)", x = "", y = "Fishing Hours") +
  theme_minimal()

box_overall
box_trim
```

## 5. Distribution Shape + Log Transformation
We use `catch_value_usd` to compare raw vs. log distributions.

**functions:**  
`pivot_longer()`: Reshapes data for faceting.  
`geom_histogram()`: Shows distribution shape.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps values to axes.  
`facet_wrap()`: Places raw and log plots side by side.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Gather raw and log values into a long format.  
Plot the raw distribution of `catch_value_usd`.  
Plot the log-transformed distribution side by side for comparison.

```{r}
catch_long <- df_imputed %>%
  select(catch_value_usd, log_catch_value) %>%
  pivot_longer(cols = everything(), names_to = "scale", values_to = "value")

ggplot(catch_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "#009E73", color = "white") +
  facet_wrap(~ scale, scales = "free_x") +
  labs(
    title = "Pre vs. Post Log Transform",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()
```

**Question 5.1.1:** Which distribution looks closer to normal: raw or log-transformed?

## 6. Relationships + Multicollinearity
We examine correlation and linear relationships among numeric variables.

### 6.1 Correlation Heatmap (Raw Units)
**functions:**  
`cor()`: Calculates correlations between numeric variables.  
`geom_tile()`: Draws a heatmap.
`as.table()`: Converts a matrix to long form.  
`as.data.frame()`: Turns a table into a data frame.  
`scale_fill_gradient2()`: Sets a diverging color scale.  
`geom_text()`: Adds correlation labels to tiles.  
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to visual channels.  
`theme_minimal()`: Applies a clean theme.  
`theme()`: Adjusts axis text.

**logic steps:**  
Select numeric columns in raw units.  
Compute the correlation matrix.  
Reshape the matrix for plotting.  
Draw a heatmap to spot large correlations (multicollinearity).

```{r}
corr_mat <- round(cor(df_imputed %>% select(apparent_fishing_hours, effort_index, fuel_consumption, catch_value_usd, crew_size), use = "complete.obs"), 2)

corr_long <- as.data.frame(as.table(corr_mat))
colnames(corr_long) <- c("Var1", "Var2", "Correlation")

corr_plot <- ggplot(corr_long, aes(Var1, Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0, limits = c(-1, 1)) +
  geom_text(aes(label = Correlation), size = 3) +
  labs(title = "Correlation Matrix (Raw Units)") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

corr_plot
```

**Question 6.1.1:** Which two variables are most strongly correlated?

**Question 6.1.2:** If you were predicting `catch_value_usd`, do `fuel_consumption` and `apparent_fishing_hours` tell a similar story? Why might that be a problem?

### 6.2 Scatterplot with Linear Model
**functions:**  
`geom_point()`: Plots raw data points.  
`geom_smooth(method = "lm")`: Adds a linear trend line.  
`lm()`: Fits a line with a slope and intercept.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Plot `catch_value_usd` against `apparent_fishing_hours`.  
Fit a linear model and extract slope/intercept.  
Add the fitted line to the plot.  
Display the equation in the subtitle for interpretation.

```{r}
lm_fit <- lm(catch_value_usd ~ apparent_fishing_hours, data = df_imputed)
coef_vals <- coef(lm_fit)

intercept <- round(coef_vals[1], 2)
slope <- round(coef_vals[2], 2)
eq_text <- paste0("catch_value_usd = ", intercept, " + ", slope, " * apparent_fishing_hours")

scatter_plot <- ggplot(df_imputed, aes(x = apparent_fishing_hours, y = catch_value_usd)) +
  geom_point(alpha = 0.5, color = "#0072B2") +
  geom_smooth(method = "lm", se = FALSE, color = "#009E73", linewidth = 1) +
  labs(
    title = "Catch Value vs. Fishing Hours",
    subtitle = eq_text,
    x = "Apparent Fishing Hours",
    y = "Catch Value (USD)"
  ) +
  theme_minimal()

scatter_plot
```

**Question 6.2.1:** Does the scatterplot suggest a strong linear relationship? Why?

## 7. Boxplot by Gear (Trimmed Dataset)
We compare gear types using the trimmed dataset.

**functions:**  
`geom_boxplot()`: Highlights median, quartiles, and outliers.  
`theme()`: Used here to rotate axis labels.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.  
`element_text()`: Styles axis labels.

**logic steps:**  
Use the trimmed dataset from Section 4.  
Map `geartype` to the x-axis and fishing hours to the y-axis.  
Render a boxplot to compare distributions across gear types.

```{r}
box_plot <- ggplot(df_trim, aes(x = geartype, y = apparent_fishing_hours, fill = geartype)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Fishing Hours by Gear", x = "Gear Type", y = "Fishing Hours")

box_plot
```

**Question 7.1.1:** Which gear types show the most extreme outliers after trimming?

**Question 7.1.2:** Why is a boxplot more informative than a mean alone for spotting outliers?

## 8. Lab Conclusion
Final Reflection: If you were writing a policy report, which imputation strategy would you defend and why?
