---
title: "Lab 02 – Tidy Data Cleaning and Foundational EDA"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide # This tells the engine to hide the code by default
    toc_float: true
---

## 💻 Environment Setup

If you do not have these packages installed, please uncomment and run the following commands:

```{r install-packages, eval=FALSE}
# install.packages("readr")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr") # We will use this in later labs
```

```{r}
# Load the essential Tidyverse libraries for data manipulation and visualization
library(readr)
library(dplyr)
library(ggplot2)

# Set global options for the code chunks
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Define the root path (Students must ensure this path points to their data directory)
ENIA_ROOT <- "~/ENIA-5120"
league_path <- file.path("../../../data", "laser_league_players.csv")
```

## 1. Load Data and Pipe Practice

### 1.1 Essential Load Functions
**functions:**  
`read_csv()`: Loads data much faster and smarter than base R's `read.csv()`.

`glimpse()`: Provides a clean, vertical overview of data structure and variable types.

`paste()`: A base R function that concatenates strings and variables together.

```{r}
players <- read_csv(league_path)
glimpse(players)
```

### 1.2 First Pipe Practice (`%>%`)
**functions:**  
`%>%` (The Pipe): Takes the output of the code on the left and passes it as the first argument to the code on the right ("AND THEN...").

`filter()`: Selects rows based on a logical condition (e.g., role is 'Sniper').

`arrange()`: Sorts the rows of the data frame.

`desc()`: Used inside `arrange()` to specify descending order (high to low).

**logic steps:**

1. Start with the `players` data frame.
2. Filter for players whose role is `'Sniper'`.
3. Arrange the results by `season_score` from highest to lowest.

```{r}
players %>%
  filter(role == "Sniper") %>%
  arrange(desc(season_score))
```

## 2. Diagnosing and Cleaning Missing Data (NA)

### 2.1 Diagnose Missing Values
**functions:**  
`is.na()`: Returns TRUE for every missing value in the data frame.

`colSums()`: Sums the TRUE values (which R treats as 1s) for each column.

```{r}
colSums(is.na(players))
```

### 2.2 Mode Imputation (Categorical Data: preferred_loadout)
We will focus on the `preferred_loadout` column and fill its NAs using the Mode.

**functions:**  
`count()`: A powerful dplyr function that counts the occurrences of a categorical variable.

`slice(1)`: Selects the first row of the data frame (which, after sorting, will contain the mode).

`pull()`: Extracts a single column as a vector (simple list of values) for assignment.

`coalesce()`: Replaces NA values in the first argument with the corresponding value from the second argument (our mode).

**logic steps (Finding Mode):**

1. Start with `players`.
2. Count the occurrences of each unique `preferred_loadout`.
3. Arrange the results in descending order (n is the count column).
4. Slice the top row to isolate the mode.
5. Pull the `preferred_loadout` name from that row.

```{r}
mode_loadout <- players %>%
  count(preferred_loadout) %>%
  arrange(desc(n)) %>%
  slice(1) %>%
  pull(preferred_loadout)

paste("The Mode is:", mode_loadout)
```

**logic steps (Applying Mode):**

1. Start with `players`.
2. Mutate (modify) the `preferred_loadout` column.
3. Use `coalesce()` to fill any NA with the calculated `mode_loadout`.
4. Store the result in `players_clean`.
5. Check the NA count again to confirm the fix.

```{r}
players_clean <- players %>%
  mutate(
    preferred_loadout = coalesce(preferred_loadout, mode_loadout)
  )

colSums(is.na(players_clean))
```

## 3. Foundational EDA: Summarizing Data

### 3.1 Summarize Scores by Role
**functions:**  
`group_by()`: Prepares the data for calculations on groups (e.g., 'Captain', 'Scout'). Must be used before `summarise()`.

`summarise()`: Calculates aggregate statistics for each group (e.g., mean, count, sum).

`n()`: Used inside `summarise()` to count the number of observations.

`mean()`: Calculates the average score.

`sd()`: Calculates the standard deviation (measure of spread/variability).

`na.rm = TRUE`: An argument used in `mean`/`sd`/`sum` to ignore any remaining NA values.

**logic steps:**

1. Start with the `players_clean` data.
2. Group the data by `role`.
3. Summarise the following for each role: total count (n), average score, score variability (standard deviation).
4. Arrange the final table by average score.

```{r}
role_summary <- players_clean %>%
  group_by(role) %>%
  summarise(
    total_players = n(), 
    avg_score = mean(season_score, na.rm = TRUE),
    score_variability = sd(season_score, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_score))

role_summary
```

*Question 3.1:* Based on the table, which player role has the highest average score? Which role has the most variability?

### 3.2 Summarize Stamina by Squad
**logic steps:**

1. Start with the clean data.
2. Group the data by `squad_name`.
3. Summarise the count and average `stamina_score` for each squad.
4. Arrange the table by average stamina score.

```{r}
squad_summary <- players_clean %>%
  group_by(squad_name) %>%
  summarise(
    total_players = n(),
    avg_stamina = mean(stamina_score, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_stamina))

squad_summary
```

## 4. Simple Visual Diagnostics

### 4.1 Checking Distribution and Outliers (Box Plot)
We will use a box plot to visually diagnose outliers and compare the median scores across the different `squad_name`s.

**functions:**  
- `ggplot()`: Initializes the plot.
- `aes()`: Defines the aesthetics.
- `geom_boxplot()`: Creates a box-and-whisker plot.
- `labs()`: Sets the plot labels.
- `theme(axis.text.x = ...)`: Rotates labels to prevent overlap.

```{r}
players_clean %>%
  ggplot(aes(x = squad_name, y = season_score)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(
    title = "Season Score Distribution by Squad",
    y = "Season Score",
    x = "Squad Name"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Question 4.1:* Identify the squad(s) that appear to have the highest number of outliers. What might that signal?

### 4.2 Checking Frequencies for Inconsistencies (Bar Chart)
We will use a bar chart to look for inconsistencies in the `home_arena` data (e.g., different capitalization).

```{r}
players_clean %>%
  ggplot(aes(x = home_arena)) +
  geom_bar(fill = "darkgreen", alpha = 0.7) +
  labs(
    title = "Player Counts by Home Arena",
    y = "Number of Players (Frequency)",
    x = "Home Arena"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Question 4.2:* Based on the bar chart, do you see potential inconsistencies in `home_arena` (like capitalization differences)? If yes, which tidyverse verb would you use to fix them?

## 5. Lab Conclusion

You have successfully loaded, diagnosed, cleaned a categorical variable, generated core summary statistics, and created simple visual diagnostics using the tidyverse framework. This is the foundation for your analysis!
