---
title: "Lab 02 – Data Cleaning Strategies and Exploratory Data Analysis (EDA)"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

## 🧭 Mission Background
You are the lead analyst for a public policy team reviewing global fishing effort. Before the team makes policy recommendations, you must inspect the data, diagnose missingness, and test how different imputation strategies change the story.

**Your Goal:** Use EDA to understand the shape of the data and apply deletion, mean/median imputation, and mode imputation to handle missing values.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads external packages into your session.
# readr: Part of the Tidyverse, used for fast and "smart" data loading.
# dplyr: Provides the tools for inspection and manipulation.

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)

# Define the root path (Adjust if your folder structure is different)
ENIA_ROOT <- "~/ENIA-5120"
efforts_path <- file.path("../../../data", "gfw_outputs", "efforts_by_flag_gear.csv")
```

## 1. The Evidence Intake: Load & Inspect
We start by pulling the dataset into R and confirming its structure.

### 1.1 Load the Data
**functions:**  
`read_csv()`: Reads data into R quickly and reliably.  
`head()`: Displays the first 6 rows for a quick sanity check.

**logic steps:**  
Locate the file path in `efforts_path`.  
Load the CSV into a data frame named `df`.  
Print the first rows to confirm columns and values look reasonable.

```{r}
df <- read_csv(efforts_path)
head(df)
```

### 1.2 Inspect the Structure
**functions:**  
`class()`: Tells you the object type (e.g., tibble/data.frame).  
`glimpse()`: Gives a vertical view of columns and data types.  
`nrow()` / `ncol()`: Counts rows and columns.

**logic steps:**  
Confirm `df` is a data frame/tibble.  
Scan column names and data types with `glimpse()`.  
Record row and column counts to know dataset size.

```{r}
"<write your code here>"
```

### 1.3 Basic Statistical Summary
**functions:**  
`summary()`: Provides min, max, mean, and quartiles for numeric columns.  
`select(where(is.numeric))`: Keeps only numeric variables.

**logic steps:**  
Keep only numeric columns to avoid text summaries.  
Run `summary()` to inspect min, max, and quartiles.  
Use this to spot unusually large or small values.

```{r}
"<write your code here>"
```

## 2. Missingness + Balance
We check missingness and whether categories are balanced.

### 2.1 Missingness by Column
**functions:**  
`is.na()`: Checks every cell; returns TRUE if a value is missing (NA).  
`colSums()`: Adds up TRUE values to count missingness by column.

**logic steps:**  
Use `is.na()` inside `colSums()` to count missing values by column.

```{r}
"<write your code here>"
```

**Question 2.1.1:** Which column has the most missing values?

### 2.2 Class Imbalance Check
**functions:**  
`count()`: Tallies how many observations fall in each category.  
`arrange(desc(n))`: Sorts counts from highest to lowest.  
`filter(!is.na())`: Drops missing categories when needed.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to visual channels.  
`geom_col()`: Draws bars with heights from data.  
`labs()`: Adds plot titles and axis labels.  
`theme_minimal()`: Applies a clean theme.  
`theme()`: Fine-tunes text and layout.  
`element_text()`: Styles axis labels.
`reorder()`: Orders categories by a numeric value.

**logic steps:**  
Remove missing flags for a fair comparison table.  
Count records per `flag` and per `geartype`.  
Sort counts to see which categories dominate.  
Plot each count table as a bar chart for visual balance.

```{r}
"<write your code here>"
```

**Question 2.2.1:** Is the dataset balanced across flags, or does one flag dominate?

**Question 2.2.2:** Why can class imbalance distort policy conclusions?

## 3. Imputation Comes First
We will use **imputed data moving forward**.

### 3.1 Mean vs. Median (Numeric)
**functions:**  
`filter(!is.na(x))`: Deletion tool that drops missing rows.  
`mean()` / `median()`: Measures central tendency; `na.rm = TRUE` ignores NAs.  
`coalesce()`: Imputation tool that fills NAs with a replacement value.  
`sd()`: Calculates Standard Deviation (spread).
`mutate()`: Adds or replaces a column in the data frame.

**logic steps:**  
Create a deletion-only version by removing missing `crew_size`.  
Compute the mean and median to see if they diverge.  
Create two imputed versions using mean and median.  
Compare resulting mean and SD across strategies to judge impact.

```{r}
"<write your code here>"
```

**Question 3.1.1:** Which strategy produced the lowest SD (variability)?

**Question 3.1.2:** Why might lower variability after imputation be misleading?

**Question 3.1.3:** When the mean and median differ, which is safer to use and why?

### 3.2 Mode Imputation (Categorical)
**functions:**  
`count()`: Counts category frequencies.  
`arrange(desc(n))`: Sorts from most common to least.  
`slice(1)`: Selects the top row (the mode).  
`pull()`: Extracts a single value from a table.  
`coalesce()`: Fills missing values with the replacement.
`filter(!is.na())`: Removes missing categories before counting.  
`mutate()`: Replaces missing values in a column.  
`table()`: Base R frequency check for before/after.

**logic steps:**  
Filter to non-missing flags for a clean frequency count.  
Identify the most frequent `flag` (the mode).  
Impute missing flags using the mode.  
Compare the mode frequency before and after imputation.

```{r}
"<write your code here>"
```

**Question 3.2.1:** By how many observations did the mode category increase after imputation?

**Question 3.2.2:** Why can mode imputation reduce diversity in categorical data?

### 3.3 Imputed Dataset for Analysis
```{r}
"<write your code here>"
```

## 4. Outliers: 3 Standard Deviations Away
We flag extreme values using the 3-SD rule.

**functions:**  
`mean()` / `sd()`: Compute center and spread.  
`abs()`: Calculates distance from the mean.
`filter()`: Keeps only rows that meet a condition.

**logic steps:**  
Compute the mean and SD of `apparent_fishing_hours`.  
Measure each value’s distance from the mean.  
Flag rows whose distance exceeds 3 SD.  
Count how many rows are flagged.

```{r}
"<write your code here>"
```

**Question 4.1.1:** How many rows are flagged as outliers by the 3-SD rule?

**Question 4.1.2:** Why can the 3-SD rule be misleading if the data is skewed?

### 4.2 Overall Boxplot: With vs. Without Outliers
**functions:**  
`geom_boxplot()`: Highlights median, quartiles, and outliers.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Create a trimmed dataset by removing 3-SD outliers.  
Plot the full distribution to show raw spread.  
Plot the trimmed distribution to show the typical range.

```{r}
"<write your code here>"
```

## 5. Distribution Shape + Log Transformation
We use `catch_value_usd` to compare raw vs. log distributions.

**functions:**  
`pivot_longer()`: Reshapes data for faceting.  
`geom_histogram()`: Shows distribution shape.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps values to axes.  
`facet_wrap()`: Places raw and log plots side by side.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Gather raw and log values into a long format.  
Plot the raw distribution of `catch_value_usd`.  
Plot the log-transformed distribution side by side for comparison.

```{r}
"<write your code here>"
```

**Question 5.1.1:** Which distribution looks closer to normal: raw or log-transformed?

## 6. Relationships + Multicollinearity
We examine correlation and linear relationships among numeric variables.

### 6.1 Correlation Heatmap (Raw Units)
**functions:**  
`cor()`: Calculates correlations between numeric variables.  
`geom_tile()`: Draws a heatmap.
`as.table()`: Converts a matrix to long form.  
`as.data.frame()`: Turns a table into a data frame.  
`scale_fill_gradient2()`: Sets a diverging color scale.  
`geom_text()`: Adds correlation labels to tiles.  
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to visual channels.  
`theme_minimal()`: Applies a clean theme.  
`theme()`: Adjusts axis text.

**logic steps:**  
Select numeric columns in raw units.  
Compute the correlation matrix.  
Reshape the matrix for plotting.  
Draw a heatmap to spot large correlations (multicollinearity).

```{r}
"<write your code here>"
```

**Question 6.1.1:** Which two variables are most strongly correlated?

**Question 6.1.2:** If you were predicting `catch_value_usd`, do `fuel_consumption` and `apparent_fishing_hours` tell a similar story? Why might that be a problem?

### 6.2 Scatterplot with Linear Model
**functions:**  
`geom_point()`: Plots raw data points.  
`geom_smooth(method = "lm")`: Adds a linear trend line.  
`lm()`: Fits a line with a slope and intercept.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.

**logic steps:**  
Plot `catch_value_usd` against `apparent_fishing_hours`.  
Fit a linear model and extract slope/intercept.  
Add the fitted line to the plot.  
Display the equation in the subtitle for interpretation.

```{r}
"<write your code here>"
```

**Question 6.2.1:** Does the scatterplot suggest a strong linear relationship? Why?

## 7. Boxplot by Gear (Trimmed Dataset)
We compare gear types using the trimmed dataset.

**functions:**  
`geom_boxplot()`: Highlights median, quartiles, and outliers.  
`theme()`: Used here to rotate axis labels.
`ggplot()`: Initializes a plot object.  
`aes()`: Maps variables to axes.  
`labs()`: Adds titles and axis labels.  
`theme_minimal()`: Applies a clean theme.  
`element_text()`: Styles axis labels.

**logic steps:**  
Use the trimmed dataset from Section 4.  
Map `geartype` to the x-axis and fishing hours to the y-axis.  
Render a boxplot to compare distributions across gear types.

```{r}
"<write your code here>"
```

**Question 7.1.1:** Which gear types show the most extreme outliers after trimming?

**Question 7.1.2:** Why is a boxplot more informative than a mean alone for spotting outliers?

## 8. Lab Conclusion
Final Reflection: If you were writing a policy report, which imputation strategy would you defend and why?
