---
title: "Lab 02 – Tidy Data Cleaning and Foundational EDA"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    code_folding: hide # This tells the engine to hide the code by default
    toc_float: true
---

## 💻 Environment Setup
**functions:** 

`library()`: Loads external packages into your R session.  
`read_csv()`: A tidyverse function that reads data into R; it is faster and more reliable than the base read.csv().

```{r}
library(readr)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
players <- read_csv("../../../data/laser_league_players.csv")
```

## 1. The Detective Phase: Initial EDA Diagnosis
**functions:** 

`is.na()`: Checks every cell in the data; returns TRUE if a value is missing (NA).  
`colSums()`: Adds up values in a column; used here to count the total number of TRUE (missing) values.  
`geom_histogram()`: A ggplot2 layer that visualizes the distribution of a single numeric variable to check for skewness.

**logic steps:**

- Quantify the Flaws: Use `is.na()` inside `colSums()` to get a count of missingness for every variable.
- Check the Shape: Plot a histogram of `season_score` to determine if the data is symmetrical (normal) or skewed.

```{r}
colSums(is.na(players))

ggplot(players, aes(x = season_score)) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(title = "Diagnostic: Checking for Skewness", x = "Season Score")
```

**Question 1.1:** Based on the histogram, is the season_score variable symmetrical or skewed?

**Question 1.2:** According to the "Rule of Thumb" in the slides, does this shape suggest we should use the Mean or the Median for imputation? Why?

## 2. Numerical Imputation: Comparing the Consequences
**functions:** 

`filter(!is.na(x))`: The Deletion tool; it removes any row where the specified column is missing.  
`mean()` / `median()`: Calculates the average or the middle value; `na.rm = TRUE` tells R to ignore NAs during the calculation.  
`coalesce()`: The Imputation tool; it looks at a column and fills any NA with the value you provide (the "replacement").  
`sd()`: Calculates Standard Deviation; a measure of how spread out the data is (Dispersion).

**logic steps:**

- Strategy A (Deletion): Create a version of the data where missing rows are simply dropped.
- Strategy B (Mean Imputation): Fill NAs with the average (vulnerable to outliers).
- Strategy C (Median Imputation): Fill NAs with the middle value (robust against outliers).
- Compare: Evaluate how each strategy changes the "Spread" (SD) of the data.

```{r}
# Execute strategies
players_deleted <- players %>% filter(!is.na(season_score))

mean_val <- mean(players$season_score, na.rm = TRUE)
players_mean_imp <- players %>% mutate(season_score = coalesce(season_score, mean_val))

median_val <- median(players$season_score, na.rm = TRUE)
players_median_imp <- players %>% mutate(season_score = coalesce(season_score, median_val))

# Compare spread
comparison <- data.frame(
  Strategy = c("Deletion", "Mean Imputation", "Median Imputation"),
  Resulting_Mean = c(mean(players_deleted$season_score), mean(players_mean_imp$season_score), mean(players_median_imp$season_score)),
  Resulting_SD = c(sd(players_deleted$season_score), sd(players_mean_imp$season_score), sd(players_median_imp$season_score))
)
comparison
```

**Question 2.1:** Look at the Resulting_SD (Standard Deviation). Which strategy resulted in the lowest variability?

**Question 2.2:** Your slides mention a "Trade-Off" for imputation. Looking at these results, why might a researcher be worried about "artificially reducing variability"?

## 3. Categorical Imputation: Mode vs. Deletion
**functions:** 

`count()`: Groups unique text labels and counts how many times each appears.  
`slice(1)`: Grabs only the very first row of a data frame (useful after sorting).  
`pull()`: Extracts a single value out of a table so it can be saved as a variable.

**logic steps:**

- Find the Mode: Count frequencies, sort from high to low, and pull the top result.
- Apply Strategy: Use `mutate` and `coalesce` to fill missing text labels with that Mode.

```{r}
# Find the mode
mode_val <- players %>% 
  count(preferred_loadout) %>% 
  arrange(desc(n)) %>% 
  slice(1) %>% 
  pull(preferred_loadout)

players_clean <- players_median_imp %>%
  mutate(preferred_loadout = coalesce(preferred_loadout, mode_val))

# Compare count of the mode
# Before:
table(players$preferred_loadout)[mode_val]
# After:
table(players_clean$preferred_loadout)[mode_val]
```

**Question 3.1:** By how many observations did the Mode category increase after imputation?

**Question 3.2:** If this were a policy dataset, how might imputing the Mode bias your final report's conclusions about policy diversity?

## 4. Deep EDA: Summary Statistics & Grouping
**functions:** 

`group_by()`: The "Split" step; it tells R to treat different categories (like player roles) as separate groups.  
`summarise()`: The "Aggregate" step; it reduces many rows of data into a single summary row per group.  
`n()`: A helper function for `summarise` that counts how many observations are in each group.

**logic steps:**

- Split the cleaned data by role.
- Calculate the Mean, Median, and Count for each role stack.
- Compare the results to find skewed roles.

```{r}
role_eda <- players_clean %>%
  group_by(role) %>%
  summarise(
    n_players = n(),
    avg_score = mean(season_score),
    med_score = median(season_score),
    score_diff = mean(season_score) - median(season_score)
  ) %>%
  arrange(desc(avg_score))

role_eda
```

**Question 4.1:** Which role has the largest difference (score_diff) between its Mean and Median?

**Question 4.2:** Based on your slides, if the Mean is significantly higher than the Median, what does that tell us about the presence of outliers in that group?

## 5. Visual Diagnostics: Spotting Anomalies
**functions:** 

`geom_boxplot()`: Creates a box-and-whisker plot; the "box" is the middle 50% of data, and "dots" are outliers.  
`theme()`: Used here to rotate x-axis labels (`element_text(angle = 45)`) so they don't overlap.

**logic steps:**

- Visualize: Map squads to the x-axis and scores to the y-axis.
- Identify: Look for "dots" far outside the whiskers—these represent the Outliers mentioned in your slides.

```{r}
ggplot(players_clean, aes(x = squad_name, y = season_score, fill = squad_name)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "EDA Diagnostic: Identifying Outliers by Squad",
       subtitle = "Visuals show what summary tables hide")
```

**Question 5.1:** Which squad(s) show the most extreme outliers?

**Question 5.2:** Why is it "impossible to tell" this information from the summary table in Section 4 alone?

## 6. Lab Conclusion
Final Reflection: You have performed both Deletion and Imputation. If you were writing a policy brief for a government agency, which of these two strategies would you feel more comfortable defending, and why?
