---
title: "Lab 08 - Clustering with Abalone"
subtitle: "K-means, Elbow Plot, and Hierarchical Clustering"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## Mission Background
You are exploring whether abalone measurements naturally form groups. The observed `Type` labels (`M`, `F`, `I`) will be used only for comparison after clustering.

**Your Goal:** Build one K-means workflow and one hierarchical workflow, then compare what each method discovers.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

abalone_path_candidates <- c(
  "../../../data/abalone.csv",
  "data/abalone.csv"
)
abalone_path <- abalone_path_candidates[file.exists(abalone_path_candidates)][1]
```

---

## Dataset Variables (Used in This Lab)

- `Type`: observed label for comparison only (`M`, `F`, `I`)
- `LongestShell`, `Diameter`, `Height`: size measurements
- `WholeWeight`, `ShuckedWeight`, `VisceraWeight`, `ShellWeight`: weight measurements
- `Rings`: approximate age proxy

---

## 0. Clustering Question

> Do the numeric abalone measurements naturally form a few groups, and do those groups line up with `M`, `F`, and `I` in a meaningful way?

---

## 1. Load Raw Data

**functions:**
- `read.csv()` - load dataset.
- `glimpse()` - inspect column types.

**logic steps:**
- Load the abalone dataset.
- Confirm that `Type` is present for later comparison and that the measurements are numeric.

```{r}
abalone_raw <- read.csv(abalone_path, stringsAsFactors = FALSE)
glimpse(abalone_raw)
```

---

## 2. Keep Numeric Features for Clustering

**functions:**
- `select()` - keep the columns needed for clustering.
- `mutate()` - make `Type` a factor if needed.

**logic steps:**
- Keep `Type` in the main data frame for later comparison.
- Create a numeric-only data frame for clustering.
- Do not do extra cleaning beyond the needed column selection.

```{r}
abalone_df <- abalone_raw %>%
  mutate(Type = as.factor(Type))

abalone_numeric <- abalone_df %>%
  select(-Type)

glimpse(abalone_numeric)
```

**Question 2.1:** Why should `Type` be kept out of the clustering features but still be saved for later comparison?

---

## 3. Standardize the Clustering Variables

**functions:**
- `scale()` - standardize numeric columns.
- `summary()` / `colMeans()` / `apply()` - check the result.

**logic steps:**
- Standardize all clustering variables.
- Save the scaled matrix as `abalone_scaled`.
- Verify that the variables are now on comparable scales.

```{r}
abalone_scaled <- scale(abalone_numeric)

round(colMeans(abalone_scaled), 4)
round(apply(abalone_scaled, 2, sd), 4)
```

**Question 3.1:** Why is standardization necessary before Euclidean-distance clustering on this dataset?

---

## 4. Candidate K-Means Solutions

**functions:**
- `kmeans()`
- `map_dbl()` or `sapply()`
- `set.seed()`

**logic steps:**
- Fit K-means for `K = 1` through `K = 8`.
- Save total within-cluster sum of squares for each value of `K`.
- Store the results in `wcss_df`.

```{r}
"<write your code here>"
```

---

## 5. Elbow Plot and K Choice

**functions:**
- `ggplot()`
- `geom_line()`
- `geom_point()`

**logic steps:**
- Plot `K` against total within-cluster sum of squares.
- Choose one final `K`.
- Save your chosen value as `chosen_k`.

```{r}
"<write your code here>"
```

**Question 5.1:** Why did you choose this value of `K`? Your answer should refer to the elbow plot, not just preference.

---

## 6. Final K-Means Model

**functions:**
- `kmeans()`
- `mutate()`
- `prcomp()`

**logic steps:**
- Fit a final K-means model with `chosen_k` and multiple random starts.
- Save the model as `kmeans_final`.
- Add cluster assignments back to a plotting data frame.

```{r}
"<write your code here>"
```

---

## 7. Plot K-Means Clusters and Centroids

**functions:**
- `ggplot()`
- `geom_point()`
- `labs()`

**logic steps:**
- Use the first two principal components for a 2D visualization.
- Plot observations colored by K-means cluster.
- Plot centroids on the same figure.

```{r}
"<write your code here>"
```

**Question 7.1:** Looking at the K-means plot, what seems to be driving the cluster separation?

---

## 8. Hierarchical Clustering

**functions:**
- `dist()`
- `hclust()`
- `cutree()`

**logic steps:**
- Compute Euclidean distances on the standardized data.
- Fit hierarchical clustering with complete linkage.
- Cut the dendrogram into `chosen_k` clusters.

```{r}
"<write your code here>"
```

**Question 8.1:** Why is it useful to use the same `K` here when comparing hierarchical clustering to K-means?

---

## 9. Dendrogram and Hierarchical Cluster Plot

**functions:**
- `plot()`
- `rect.hclust()`
- `ggplot()`

**logic steps:**
- Plot the dendrogram.
- Build a second 2D plot using the same principal-component coordinates, now colored by hierarchical cluster.

```{r}
"<write your code here>"
```

**Question 9.1:** Compared with K-means, what looks different about the hierarchical clustering result?

---

## 10. Compare Clusters to Type (`M`, `F`, `I`)

**functions:**
- `table()`
- `prop.table()`
- `as.data.frame()`

**logic steps:**
- Build one comparison table for K-means clusters vs `Type`.
- Build one comparison table for hierarchical clusters vs `Type`.
- Convert to row proportions to make interpretation easier.

```{r}
"<write your code here>"
```

**Question 10.1:** Does either clustering method line up cleanly with `M`, `F`, and `I`? Explain with evidence from your comparison tables.

**Question 10.2:** If the clusters do not match `Type` very well, what might the clusters be capturing instead?

---

## 11. Final Reflection

**Summary Question 11.1:**
If you had to explain this analysis to a non-technical audience, which method would you present first, and what is one limitation you would emphasize?
