---
title: "Lab 01: Climate Station Forensic Audit"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
Welcome to your first day as a Junior Auditor for the **Global Climate Monitoring Network**. You've been handed a raw export from 20 monitoring stations around the world. Before anyone downstream (scientists, policymakers, journalists) can trust a single chart built from this data, you need to prove it's structurally sound and free of "Gremlins."

**Your Goal:** Perform a forensic audit. Confirm what data types R assigned, check the table's structural integrity, and document every quality issue you find. These are the same six categories from lecture.

```{r setup, message=FALSE, warning=FALSE}
# --- UTILITY REFERENCE ---
# library(): Loads the forensic toolkit into your session.
# readr: Part of the Tidyverse, used for fast and "smart" data loading.
# dplyr: Provides the tools for inspection (glimpse, count) and manipulation (mutate).

library(readr)
library(dplyr)

# Define the root path (adjust if your folder structure is different)
climate_path <- file.path("../../data", "climate_stations.csv")
```

---

## 1. The Blueprint: Loading & Inspection
Our first task is to bring the evidence onto our digital workbench (the R Environment).

### 1.1 Load the Evidence
**functions:**
`read_csv()`: Loads data much faster and smarter than base R's `read.csv()`.
`head()`: Displays the first 6 rows of a dataset.

**logic steps:**
Use `read_csv()` to load the file located at `climate_path`.
Store the result in an object named `climate`.
Use `head()` to verify the table looks correct.

```{r}
"<write your code here>"
```

### 1.2 Identify the Data Structure
**functions:**
`class()`: Tells you the "Object Type" (e.g., tibble/data.frame).
`glimpse()`: Provides a clean, vertical overview of column names and variable types.

**logic steps:**
Check the `class()` of your `climate` object to confirm R treats it as a table.
Use `glimpse()` to see every column name and its assigned data type.

```{r}
"<write your code here>"
```

**Question 1.2.1:** What R class is `climate`? How does R internally store this table?

```
<write your answer here>
```

**Question 1.2.2:** Based on `glimpse()`, how many rows (observations) and columns (variables) are present?

```
<write your answer here>
```

---

## 2. The Language: Data Type Checks
In R, the type of data determines what math is allowed. You cannot calculate the "average" of a word.

### 2.1 Inspect Individual Columns
**functions:**
`class(df$column)`: Extracts a specific column using the `$` sign and reveals its type.

**logic steps:**
Check the class of `temp_c` and `extreme_weather_events`. Both are numeric, but not the same *kind* of numeric.
Check the class of `energy_source` and `climate_risk_level`. Both are categorical, but not the same *kind* of categorical.

```{r}
"<write your code here>"
```

**Question 2.1.1:** `temp_c` and `extreme_weather_events` are both numeric. Which one is **continuous** and which is **discrete**? Why?

```
<write your answer here>
```

**Question 2.1.2:** `energy_source` and `climate_risk_level` are both categorical. Which one is **nominal** and which is **ordinal**? Why does that distinction matter for later analysis?

```
<write your answer here>
```

### 2.2 Numerical Summaries
**functions:**
`summary()`: Provides the Min, Max, Median, and Mean.

**logic steps:**
Run `summary()` on `temp_c` and `co2_ppm`.
Look closely at the Min and Max. An impossible value hides in plain sight here.

```{r}
"<write your code here>"
```

**Question 2.2.1:** What is the max value of `temp_c`? Is that physically possible for a surface temperature reading?

```
<write your answer here>
```

---

## 3. Forensic Audit: Missingness, Duplicates, Casing
Now we look for three of the six quality issues from lecture: gaps, repeats, and label inconsistency.

### 3.1 Missing Values (NA)
**functions:**
`is.na()`: Returns TRUE for every missing value.
`colSums()`: Adds up all the TRUEs to give a total count of missing values per column.

**logic steps:**
Combine `is.na(climate)` with `colSums()` to get a report of all missing values across the entire table.

```{r}
"<write your code here>"
```

**Question 3.1.1:** Which variable (besides `analyst_notes`, which is expected to be mostly empty) has the most missing values? Why is that a problem if it's a key climate measurement?

```
<write your answer here>
```

### 3.2 Inconsistent Casing
**functions:**
`count()`: Groups rows by a column and counts how many times each label appears.
`arrange(desc(n))`: Sorts the counts from highest to lowest.

**logic steps:**
Take `climate`, count `location`, and arrange by descending count. Repeat for `energy_source`.

```{r}
"<write your code here>"
```

**Question 3.2.1:** Do you notice inconsistent casing in either column (e.g., `Arlington` vs. `ARLINGTON`, or `Wind` vs. `WIND`)? Why would this mess up a bar chart of station locations or energy sources?

```
<write your answer here>
```

### 3.3 Duplicate Station-Year Entries
**functions:**
`count(col1, col2)`: Counts how many times each combination of two columns appears together.
`filter(n > 1)`: Keeps only combinations that appear more than once.

**logic steps:**
`station_id` alone repeats naturally, since there's one row per station per year, so a duplicate isn't "the same station_id twice." It's "the same station and the same year twice." Count `station_id` and `reading_date` together, then filter down to combinations that appear more than once.

```{r}
"<write your code here>"
```

**Question 3.3.1:** How many station-year combinations are duplicated? Why is this a bigger problem for a yearly summary than it would be for a table with no natural repeats?

```
<write your answer here>
```

---

## 4. Logical Detective Work
Time to reuse the bracket-filtering skills from Lab 0. Instead of reading a printed summary, we'll pull out the exact offending rows ourselves.

### 4.1 Isolate the Impossible Temperatures
**functions:**
`df[condition, ]`: Keeps only the rows where `condition` is TRUE.

**logic steps:**
No surface station on Earth reads above 60°C or below -60°C. Filter `climate` down to rows where `temp_c` is outside that range.

```{r}
"<write your code here>"
```

**Question 4.1.1:** How many rows have a physically impossible temperature? What's the most extreme value you found?

```
<write your answer here>
```

### 4.2 Catch the Malformed Sensor Readings
**functions:**
`class()`, `df[!df$column %in% c(...), ]`: `%in%` checks membership. `!` negates it, keeping rows that do not match.

**logic steps:**
First check what class R actually assigned to `sensor_active`. Then filter to rows where the value isn't `"True"` or `"False"`.

```{r}
"<write your code here>"
```

**Question 4.2.1:** What class did R assign to `sensor_active`? Given what you just found in the filter above, why did it end up that way instead of being a clean TRUE/FALSE logical column?

```
<write your answer here>
```

### 4.3 Catch the Invalid Risk Category
**functions:**
`df[!df$column %in% c(...), ]`

**logic steps:**
`climate_risk_level` should only ever be "Low," "Medium," or "High." Filter to rows where it isn't.

```{r}
"<write your code here>"
```

**Question 4.3.1:** What value(s) showed up that don't belong to any of the three defined levels? Why is an invalid category worse than a missing one for an ordinal variable?

```
<write your answer here>
```

### 4.4 Check the Date Column
**functions:**
`class()`

**logic steps:**
Check what class `reading_date` currently has.

```{r}
"<write your code here>"
```

**Question 4.4.1:** What class is `reading_date` right now? Given that some entries look like `"Jan 15th, 2020"` instead of `"2020-01-15"`, why can't R sort this chronologically yet?

```
<write your answer here>
```

---

## 5. Flagging with `mutate()`
We're not fixing anything yet. That's next week's "surgery." But we can build a flag that marks every row with at least one issue, so Session 2 knows exactly where to start.

**functions:**
`mutate()`: Adds a new column to the data frame.
`|` : Logical OR. TRUE if *any* of the conditions is TRUE.

**logic steps:**
Create a new logical column `has_issue` that is TRUE if the row has a missing `co2_ppm`, an impossible `temp_c`, a malformed `sensor_active`, or an invalid `climate_risk_level`.

```{r}
"<write your code here>"
```

**Question 5.1.1:** How many rows got flagged? Does that roughly match what you'd expect from adding up the individual checks above (keeping in mind some rows might trip more than one check)?

```
<write your answer here>
```

---

## 6. Write the Audit Report
No new functions here. Just synthesize what you found.

**Question 6.1:** In 4 to 6 sentences, summarize your audit. Include roughly how many rows have at least one issue, which issue type was most common, and (this is the part that takes judgment, not a lookup) if you only had time to fix one issue type before Session 2, which would you fix first and why?

```
<write your answer here>
```
