---
title: "Lab 05 – Linear Regression in Practice"
subtitle: "Agricultural Yield: Simple, Expanded, and Trimmed Models"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are an analyst studying what drives crop yield.

**Your Goal:** Build three regression versions (simple -> expanded -> trimmed) and defend your final model using both interpretability and prediction evidence.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Data path
yield_path <- "../../../data/agricultural_yield_combined.csv"
```

---

## Dataset Variables

- `soil_quality`: Soil quality index (50 to 100)
- `seed_variety`: Binary indicator (1 = high-yield variety)
- `fertilizer_amount_kg_per_hectare`: Fertilizer amount
- `sunny_days`: Number of sunny days
- `rainfall_mm`: Total rainfall during growing season
- `irrigation_schedule`: Number of irrigations
- `soil_ph`: Soil acidity/alkalinity indicator
- `farm_size_hectares`: Farm size in hectares
- `pest_management_score`: Operational pest-management score
- `harvest_delay_days`: Days delayed before harvest
- `yield_kg_per_hectare`: Target outcome

---

## 0. Policy Question

> How is crop yield associated with fertilizer use, weather, and seed variety?

---

## 1. Data Cleaning

**functions:**
- `read_csv()` – load data.
- `filter()` – remove invalid rows.
- `mutate()` – set data types.

**logic steps:**
- Load the combined dataset.
- Keep rows with valid outcome and predictor values.
- Set `seed_variety` as the only categorical predictor.
- Create `ag_clean`.

```{r}
"<write your code here>"
```

---

## 2. Train/Test Split (70/30)

**functions:**
- `set.seed()` – reproducible split.
- `sample.int()` – random row split.

**logic steps:**
- Set seed.
- Assign 70% to training and 30% to testing.
- Save as `train_df` and `test_df`.

```{r}
"<write your code here>"
```

**Question 2.1:** Why is random splitting necessary before you compare model versions?

---

## 3. Model Version 1 (Simple Baseline)

V1 should be intentionally simple.

**Required rule:**
- Use only one predictor in V1.

**functions:**
- `lm()`
- `summary()`
- `confint()`

**logic steps:**
- Fit `model_v1` with `fertilizer_amount_kg_per_hectare` only.
- Interpret direction and magnitude.

```{r}
"<write your code here>"
```

**Question 3.1:** Interpret the fertilizer coefficient in one sentence with units.

**Question 3.2:** What does the V1 confidence interval for fertilizer imply about uncertainty?

---

## 4. Model Version 2 (Expanded)

V2 adds extra predictors to test whether they add value.

**Required rule:**
- Include all predictor columns in the dataset **except** `soil_quality` and `yield_kg_per_hectare`.

**functions:**
- `lm()`
- `summary()`

**logic steps:**
- Fit `model_v2` with all listed predictors.
- Compare V2 fit against V1 fit.

```{r}
"<write your code here>"
```

**Question 4.1:** Compared with V1, how much do V2's `R²` and adjusted `R²` change? Is that increase meaningful?

**Question 4.2 (Hypothesis Test on One Variable):**  
Using the `soil_ph` coefficient in V2:
- Write the null hypothesis ($H_0$) and alternative hypothesis ($H_a$).
- State whether you reject or fail to reject $H_0$ at $\alpha = 0.05$.
- Explain your decision using the reported p-value.

---

## 5. Model Version 3 (Trimmed)

V3 should be between V1 and V2 in complexity.

**Required rule:**
- Keep more predictors than V1.
- Keep fewer predictors than V2.

**functions:**
- `lm()`
- `summary()`

**logic steps:**
- Remove weak predictors from V2.
- Fit `model_v3`.
- Compare V3 vs V2 on adjusted R² and clarity.

```{r}
"<write your code here>"
```

**Question 5.1:** Why is V3 a better compromise (or not) than V2 for practical use?

---

## 6. Residual Diagnostics (Model V3)

**functions:**
- `predict()` – generate fitted values.
- `ggplot()` + `geom_point()` – visualize residual pattern.
- `geom_hline()` – add zero-reference line.

**logic steps:**
- Compute residuals from `model_v3` as `actual - predicted`.
- Plot residuals against fitted values.
- Check whether residuals look like a random cloud around 0.

```{r}
"<write your code here>"
```

**Question 6.1:** Based on your residual plot for V3, explain why this is (or is not) a good residual pattern.

---

## 7. Model Comparison (Train + Test Evidence)

**functions:**
- `predict()` – out-of-sample predictions.
- `mean()`, `abs()`, `sqrt()` – MAE/RMSE/MAPE.
- `summary()` – extract R² and Adjusted R².
- `data.frame()` – store model comparison results in a simple table.
- `rbind()` – stack per-model rows into one comparison table.
- `order()` – rank models by lowest RMSE.

**logic steps:**
- Define one helper function that takes a model and test data, then returns all metrics.
- Apply the same function to V1, V2, and V3 so comparison is fair.
- Build one `data.frame` with one row per model.
- Rank by RMSE (lowest first), then inspect MAE/MAPE and adjusted R² to break close ties.

```{r}
"<write your code here>"
```

**Question 7.1:** Which model has the best RMSE? Does it also have the best adjusted R²?

**Question 7.2:** If V2 and V3 have very similar metrics, why might you still prefer V3?

**Question 7.3:** Which model would you recommend to farm planners and why? Use at least two metrics and one interpretation argument.

---

## 8. Summary Question

**Summary Question 8.1:**  
If you had to brief farm planners in two minutes, which model would you recommend, what is the single most useful predictor, and what is one limitation of your recommendation?
