---
title: "Lab 05 – Linear Regression in Practice"
subtitle: "Agricultural Yield: Simple, Expanded, and Trimmed Models"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are an analyst studying what drives crop yield.

**Your Goal:** Build three regression versions (simple -> expanded -> trimmed) and defend your final model using both interpretability and prediction evidence.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# Data path
yield_path <- "../../../data/agricultural_yield_combined.csv"
```

---

## Dataset Variables

- `soil_quality`: Soil quality index (50 to 100)
- `seed_variety`: Binary indicator (1 = high-yield variety)
- `fertilizer_amount_kg_per_hectare`: Fertilizer amount
- `sunny_days`: Number of sunny days
- `rainfall_mm`: Total rainfall during growing season
- `irrigation_schedule`: Number of irrigations
- `soil_ph`: Soil acidity/alkalinity indicator
- `farm_size_hectares`: Farm size in hectares
- `pest_management_score`: Operational pest-management score
- `harvest_delay_days`: Days delayed before harvest
- `yield_kg_per_hectare`: Target outcome

---

## 0. Policy Question

> How is crop yield associated with fertilizer use, weather, and seed variety?

---

## 1. Data Cleaning

**functions:**
- `read_csv()` – load data.
- `filter()` – remove invalid rows.
- `mutate()` – set data types.

**logic steps:**
- Load the dataset.
- Keep rows with valid outcome and predictor values.
- Set `seed_variety` as the only categorical predictor.
- Create `ag_clean`.

```{r}
ag_raw <- read_csv(yield_path)

ag_clean <- ag_raw %>%
  filter(
    !is.na(soil_quality),
    !is.na(seed_variety),
    !is.na(fertilizer_amount_kg_per_hectare),
    !is.na(sunny_days),
    !is.na(rainfall_mm),
    !is.na(irrigation_schedule),
    !is.na(soil_ph),
    !is.na(farm_size_hectares),
    !is.na(pest_management_score),
    !is.na(harvest_delay_days),
    !is.na(yield_kg_per_hectare)
  ) %>%
  mutate(seed_variety = as.factor(seed_variety))

glimpse(ag_clean)
```

---

## 2. Train/Test Split (70/30)

**functions:**
- `set.seed()` – reproducible split.
- `sample.int()` – random row split.

**logic steps:**
- Set seed.
- Assign 70% to training and 30% to testing.
- Save as `train_df` and `test_df`.

```{r}
set.seed(5120)
idx <- sample.int(nrow(ag_clean), size = floor(0.7 * nrow(ag_clean)))

train_df <- ag_clean[idx, ]
test_df <- ag_clean[-idx, ]

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")
cat("Train share:", round(nrow(train_df) / nrow(ag_clean), 3), "\n")
```

**Question 2.1:** Why is random splitting necessary before you compare model versions?

---

## 3. Model Version 1 (Simple Baseline)

V1 should be intentionally simple.

**Required rule:**
- Use only one predictor in V1.

**functions:**
- `lm()`
- `summary()`
- `confint()`

**logic steps:**
- Fit `model_v1` with `fertilizer_amount_kg_per_hectare` only.
- Interpret direction and magnitude.

```{r}
model_v1 <- lm(
  yield_kg_per_hectare ~ fertilizer_amount_kg_per_hectare,
  data = train_df
)

summary(model_v1)
confint(model_v1)
```

**Question 3.1:** Interpret the fertilizer coefficient in one sentence with units.

**Question 3.2:** What does the V1 confidence interval for fertilizer imply about uncertainty?

---

## 4. Model Version 2 (Expanded)

V2 adds extra predictors to test whether they add value.

**Required rule:**
- Include all predictor columns in the dataset **except** `soil_quality` and `yield_kg_per_hectare`.

**functions:**
- `lm()`
- `summary()`

**logic steps:**
- Fit `model_v2` with broader predictor set.
- Compare V2 fit against V1 fit.

```{r}
model_v2 <- lm(
  yield_kg_per_hectare ~ fertilizer_amount_kg_per_hectare + sunny_days + rainfall_mm + irrigation_schedule + soil_ph + farm_size_hectares + pest_management_score + harvest_delay_days + seed_variety,
  data = train_df
)

summary(model_v2)
```

**Question 4.1:** Compared with V1, how much do V2's `R²` and adjusted `R²` change? Is that increase meaningful?

**Question 4.2 (Hypothesis Test on One Variable):**  
Using the `soil_ph` coefficient in V2:
- Write the null hypothesis ($H_0$) and alternative hypothesis ($H_a$).
- State whether you reject or fail to reject $H_0$ at $\alpha = 0.05$.
- Explain your decision using the reported p-value.

---

## 5. Model Version 3 (Trimmed)

V3 should be between V1 and V2 in complexity.

**functions:**
- `lm()`
- `summary()`

**logic steps:**
- Remove weak predictors from V2.
- Fit `model_v3`.
- Compare V3 vs V2 on adjusted R² and clarity.

```{r}
model_v3 <- lm(
  yield_kg_per_hectare ~ fertilizer_amount_kg_per_hectare + soil_quality + sunny_days + rainfall_mm + irrigation_schedule + seed_variety,
  data = train_df
)

summary(model_v3)
```

**Question 5.1:** Why is V3 a better compromise (or not) than V2 for practical use?

---

## 6. Residual Diagnostics (Model V3)

**functions:**
- `predict()` – generate fitted values.
- `ggplot()` + `geom_point()` – visualize residual pattern.
- `geom_hline()` – add zero-reference line.

**logic steps:**
- Compute residuals from `model_v3` as `actual - predicted`.
- Plot residuals against fitted values.
- Check whether residuals look like a random cloud around 0.

```{r}
fitted_v3 <- predict(model_v3, newdata = train_df)
resid_v3 <- train_df$yield_kg_per_hectare - fitted_v3

resid_df <- data.frame(
  fitted = fitted_v3,
  residual = resid_v3
)

ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.45, color = "steelblue") +
  geom_hline(yintercept = 0, color = "firebrick", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Residual Plot for Model V3",
    x = "Fitted values (V3)",
    y = "Residuals (actual - predicted)"
  ) +
  theme_minimal()
```

**Question 6.1:** Based on your residual plot for V3, explain why this is (or is not) a good residual pattern.

---

## 7. Model Comparison (Train + Test Evidence)

**functions:**
- `predict()` – out-of-sample predictions.
- `mean()`, `abs()`, `sqrt()` – MAE/RMSE/MAPE.
- `summary()` – extract R² and Adjusted R².
- `data.frame()` – store model comparison results in a simple table.
- `rbind()` – stack per-model rows into one comparison table.
- `order()` – rank models by lowest RMSE.

**logic steps:**
- Define one helper function that takes a model and test data, then returns all metrics.
- Apply the same function to V1, V2, and V3 so comparison is fair.
- Build one `data.frame` with one row per model.
- Rank by RMSE (lowest first), then inspect MAE/MAPE and adjusted R² to break close ties.

```{r}
compute_metrics <- function(model, test_data, model_name) {
  pred <- predict(model, newdata = test_data)
  err <- test_data$yield_kg_per_hectare - pred

  return(data.frame(
    model = model_name,
    MAE = mean(abs(err), na.rm = TRUE),
    RMSE = sqrt(mean(err^2, na.rm = TRUE)),
    MAPE = mean(abs(err / pmax(test_data$yield_kg_per_hectare, 1)), na.rm = TRUE) * 100,
    R2 = summary(model)$r.squared,
    Adjusted_R2 = summary(model)$adj.r.squared
  ))
}

model_performance <- rbind(
  compute_metrics(model_v1, test_df, "V1: simple"),
  compute_metrics(model_v2, test_df, "V2: expanded"),
  compute_metrics(model_v3, test_df, "V3: trimmed")
)

model_performance <- model_performance[order(model_performance$RMSE), ]

model_performance
```

**Question 7.1:** Which model has the best RMSE? Does it also have the best adjusted R²?

**Question 7.2:** If V2 and V3 have very similar metrics, why might you still prefer V3?

**Question 7.3:** Which model would you recommend to farm planners and why? Use at least two metrics and one interpretation argument.

---

## 8. Summary Question

**Summary Question 8.1:**  
If you had to brief farm planners in two minutes, which model would you recommend, what is the single most useful predictor, and what is one limitation of your recommendation?
