---
title: "Session 6: Logistic Regression"
subtitle: "From Probability to Policy Decisions"
format: 
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true    
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

## 🧭 Opening Decision Problem

Question:

> Is this car a BEV or HPEV? **Yes or No**


This is not a "how much" question.
It is a **decision** question.

What we really need:

* Is this car a BEV? **Yes**.

---

## 🎯 Core Idea

But how do we get to **"Yes"** or **"No"**?

* Calculate **Probabilities** first
* See what level of probability you are comfortable with
* Then labels (Yes/No) using a threshold

Key line:

> Logistic regression predicts probabilities, not labels.

---

## 🚫 Why Linear Regression Fails for Yes/No

Linear regression can output values like:

* 1.3
* -0.2

But probabilities must stay between 0 and 1.

Plain language:

* A straight line has no boundaries.
* Probability must have boundaries.

---

## ✅ What We Need Instead

We need a model that:

* Uses predictors linearly
* But always returns valid probabilities (bounded between 0 and 1)
* Works naturally for binary outcomes

That model is **logistic regression**.

---

## 🤔 No..? But You Said..?

- Yes, I did say regressions are for **Numeric Outputs** 
- But if the **outcome** is binary, it's not numeric, it's categorical!
- But, notice that **outcome** is binary, but **output** is numeric (probability)
- So logistic regression, although may seem like non-regression, is classified as a "regression".


## 🧮 Logistic Regression: Step 1 (Linear Score)

Build a score:

$$
z = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots
$$

This score:

* Can be any real number
* Is **not** a probability yet
* Think of **Y** as a temporary score for the actual binary outcome
* Think of $/beta$ as just a random point (for now)

---

## 🧮 Logistic Regression: Step 2 (Sigmoid)

Convert score to probability:

$$
p = \frac{1}{1 + e^{-z}}
$$

This is the logistic (sigmoid) function.

---

## 📉 From Linear Score to Sigmoid

```{r sigmoid-demo, eval=TRUE, fig.width=8, fig.height=4.5}
library(dplyr)
library(ggplot2)

b0 <- -4
b1 <- 0.18

df <- tibble(x = seq(0, 50, by = 0.1)) %>%
  mutate(
    z = b0 + b1 * x,
    p = 1 / (1 + exp(-z))
  )

anchors <- df %>%
  filter(round(x) %in% c(5, 15, 25, 35, 45)) %>%
  mutate(id = paste0("x=", round(x)))

ggplot(df, aes(x = x, y = z)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(data = anchors, aes(color = id), size = 3) +
  labs(
    title = "Step 1: Linear Score",
    x = "x",
    y = "z = b0 + b1x"
  ) +
  theme_minimal(base_size = 12)

ggplot(df, aes(x = x, y = p)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_point(data = anchors, aes(color = id), size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "firebrick") +
  labs(
    title = "Step 2: Sigmoid Probability",
    x = "x",
    y = "p = 1/(1+e^-z)"
  ) +
  theme_minimal(base_size = 12)
```

---

## 📈 What Sigmoid Does

$$
p = \frac{1}{1 + e^{-z}}
$$

**Quick checkpoints:**

* If $z$ is very negative -> $p$ is near 0.
* If $z = 0$ -> $p = 0.5$.
* If $z$ is very positive -> $p$ is near 1.

**Why:**

* Large negative $z$ makes $e^{-z}$ very large, so denominator is large -> probability shrinks.
* At $z=0$, $e^0=1$, so $p=1/(1+1)=0.5$.
* Large positive $z$ makes $e^{-z}$ very small, so denominator approaches 1 -> probability approaches 1.

**Why the S-shape matters:**

* Middle is steep: small score changes can move decisions.
* Extremes are flat: probabilities stabilize near 0 or 1.

---

## 🧠 Plain Summary

> Logistic regression is a straight line fed through an S-curve.

---

## 🌐 Three Languages You Must Know

Logistic regression is discussed in:

1. Probability
2. Odds
3. Log-odds (logit)

---

## 📊 1. Probability

$$
p = P(Y=1)
$$

Interpretation:

* Out of 100 similar cases, how many are expected to be Yes?

---

## ⚖️ 2. Odds

$$
\text{odds} = \frac{p}{1-p}
$$

Odds compare:

* How much more likely is Yes than No?

Example:

* If $p=0.20$, then odds $= 0.20/0.80 = 0.25$
* Meaning: 1 Yes for every 4 No.

---

## 🧮 3. Log-Odds (Logit)

$$
\log\left(\frac{p}{1-p}\right)
$$

So what is the **z** that we saw in the original regression like function?

$$
\log(\text{odds}) = \beta_0 + \beta_1 x
$$

---

## ❓ Why Log-Odds?

Some of you may have noticed that sigmoid function looks **Logarithmic**

Because:

* Probability space is curved
* Log-odds space becomes linear

Key line:

> Logistic regression is linear in log-odds space.

---

## 📉 Why Log-Odds?

```{r logodds-demo, eval=TRUE, fig.width=8, fig.height=4.5}
library(dplyr)
library(ggplot2)

b0 <- -4
b1 <- 0.18

df <- tibble(x = seq(0, 50, by = 0.1)) %>%
  mutate(
    z = b0 + b1 * x,
    p = 1 / (1 + exp(-z)),
    odds = p / (1 - p),
    log_odds = log(odds)
  )

ggplot(df, aes(x = x, y = p)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  labs(
    title = "Probability (Sigmoid): Curved",
    x = "x",
    y = "p"
  ) +
  theme_minimal(base_size = 12)

ggplot(df, aes(x = x, y = log_odds)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_abline(intercept = b0, slope = b1, linetype = "dashed", color = "firebrick") +
  labs(
    title = "Log-Odds: Linear",
    subtitle = "Dashed line is b0 + b1x",
    x = "x",
    y = "log(p/(1-p))"
  ) +
  theme_minimal(base_size = 12)
```

---

## 🔍 Two Views of Same Model

Linear-score view:

$$
z = \beta_0 + \beta_1 x
$$

Probability view:

$$
p(x) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x)}}
$$

Same model, different lens.

---

## ⚙️ How Logistic Regression Is Fit

Linear regression:

* Minimizes squared errors

Logistic regression:

* Maximizes likelihood (MLE)

Meaning:

* Choose coefficients that make observed Yes/No outcomes most probable.

---

## 📉 Log-Loss Intuition

MLE behavior:

* Rewards confident correct predictions
* Punishes confident wrong predictions heavily

Log-likelihood (maximize):

$$
\ell(\beta)=\sum_{i=1}^{n}\left[y_i\log(p_i)+(1-y_i)\log(1-p_i)\right]
$$

Equivalent log-loss / negative log-likelihood (minimize):

$$
\text{LogLoss} = -\frac{1}{n}\sum_{i=1}^{n}\left[y_i\log(p_i)+(1-y_i)\log(1-p_i)\right]
$$

This objective is equivalent to minimizing log-loss.

---

## 💻 Logistic Output Example
```{r logit-output-demo, eval=TRUE}
set.seed(5120)
n <- 300

demo_df <- data.frame(
  chargers = rnorm(n, mean = 10, sd = 3),
  income = rnorm(n, mean = 65, sd = 12)
)

z <- -6 + 0.35 * demo_df$chargers + 0.04 * demo_df$income
p <- 1 / (1 + exp(-z))
demo_df$funded_yes <- rbinom(n, size = 1, prob = p)

m_logit <- glm(funded_yes ~ chargers + income, data = demo_df, family = binomial())

summary(m_logit)

exp(coef(m_logit))
```

---

## 🧾 Coefficient Interpretation

Raw logistic coefficient:

* Change in **log-odds** (hard to explain directly)

So we exponentiate:

$$
OR = e^{\beta}
$$

This gives the **odds ratio (OR)**.

---

## 🔁 Odds Ratio Meaning

If OR = 2:

* Odds of Yes are multiplied by 2 for a 1-unit increase in X

If OR = 0.7:

* Odds are multiplied by 0.7 (30% decrease)

Important:

* OR talks about odds, not direct probability change.

---

## 🔢 Odds to Probability Example

Suppose:
$$
p(x_1) = \frac{1}{1 + e^{-(-1.386 + 0.693 x_1)}}
$$

Step 1: Baseline probability at \(x_1=0\)

$$
p_0 = \frac{1}{1 + e^{-(-1.386 + 0.693 \cdot 0)}}
= \frac{1}{1 + e^{1.386}}
= 0.20
$$

Step 2: Convert baseline probability to odds

$$
\text{odds}_0 = \frac{p_0}{1-p_0} = \frac{0.20}{0.80} = 0.25
$$

Step 3: Increase \(x_1\) by 1 -> odds multiply by OR

$$
\text{odds}_1 = \text{odds}_0 \times OR = 0.25 \times 2 = 0.50
$$

Step 4: Convert updated odds back to probability

$$
p_1 = \frac{\text{odds}_1}{1+\text{odds}_1}
= \frac{0.50}{1.50}
= 0.333
$$

Step 5: Probability impact of a +1 increase in \(x_1\)

$$
\Delta p = p_1 - p_0 = 0.333 - 0.20 = 0.133
$$

Result summary:

* A +1 increase in \(x_1\) moves probability from **0.20 -> 0.333**.
* That is a **+13.3 percentage-point** increase in this baseline scenario.

---

## ⚠️ Key Lesson

> Doubling odds does not double probability.

---


## 🧪 Coefficient Significance

For each coefficient, test:

$$
H_0: \beta = 0
$$

Meaning:

* Predictor has no association with outcome

If p-value is small:

* Evidence of association

Reminder:

* Significant != large effect
* Significant != causal

---

## 📏 Confidence Intervals for OR

Example:

* OR = 1.5
* 95% CI = [1.1, 2.1]

Interpretation:

* Odds likely increase between 10% and 110%

If CI includes 1:

* Effect may be indistinguishable from no effect

CI gives magnitude uncertainty, not just yes/no.

---

## 🧩 Confusion Matrix

|               | Pred Yes | Pred No |
|---|---:|---:|
| Actual Yes | TP | FN |
| Actual No  | FP | TN |

---

## 📐 Derived Metrics

Accuracy (overall percent correct):

$$
\text{Accuracy} = \frac{TP + TN}{\text{Total}}
$$

Precision (out of predicted Yes, how many were truly Yes):

$$
\text{Precision} = \frac{TP}{TP + FP}
$$

Recall (out of actual Yes, how many we caught):

$$
\text{Recall} = \frac{TP}{TP + FN}
$$

---

## 🧭 Precision vs Recall in Practice

| Context | Usually prioritize | Why |
|---|---|---|
| Medical screening | Recall | Missing a true case is costly |
| Fraud enforcement | Precision | False accusations are costly |

From the example:

* Precision = 60% -> 40% of flagged counties were false alarms.
* Recall = 75% -> model caught 3 out of 4 counties that truly needed funding.

---

## 🎚️ Thresholds: 0.5 Is Not Sacred

Model gives probability $p$.
You choose threshold $t$:

* If $p \ge t$ -> Yes
* Else -> No

Lower threshold:

* Higher recall, lower precision

Higher threshold:

* Higher precision, lower recall

---

## 🏛️ Policy Interpretation of Threshold

Threshold is not only technical.
It reflects cost of mistakes.

* If FN is very costly -> lower threshold
* If FP is very costly -> higher threshold

---

## 📈 ROC

ROC stands for **Receiver Operating Characteristic**.

Think of ROC as:

> A report card of model behavior at every possible threshold.

| Piece | Meaning |
|---|---|
| X-axis | False Positive Rate (how many No cases are wrongly flagged) |
| Y-axis | True Positive Rate / Recall (how many Yes cases are correctly caught) |
| Better curve | Closer to top-left corner |
| Diagonal line | Random guessing baseline |
| Stronger model | Stays farther above the diagonal |

> Furthest point from the diagonal line is considered the best cutoff!

---

## 📉 ROC Plot

```{r roc-demo, eval=TRUE, fig.width=8, fig.height=5}
library(ggplot2)

set.seed(5120)
n <- 350

# Simulated true labels
y_true <- rbinom(n, size = 1, prob = 0.35)

# Three score sets with different quality
score_weak <- 0.30 + 0.18 * y_true + rnorm(n, 0, 0.28)   # weak model
score_mid  <- 0.25 + 0.45 * y_true + rnorm(n, 0, 0.24)   # medium model
score_strong <- 0.20 + 0.80 * y_true + rnorm(n, 0, 0.20) # strong model

to_prob <- function(s) pmin(pmax(s, 0.001), 0.999)

roc_from_probs <- function(p_hat, y, label) {
  thresholds <- seq(0, 1, by = 0.01)
  out <- data.frame(
    threshold = thresholds,
    TPR = NA_real_,
    FPR = NA_real_
  )

  for (i in seq_along(thresholds)) {
    t <- thresholds[i]
    pred <- ifelse(p_hat >= t, 1, 0)
    TP <- sum(pred == 1 & y == 1)
    FP <- sum(pred == 1 & y == 0)
    TN <- sum(pred == 0 & y == 0)
    FN <- sum(pred == 0 & y == 1)
    out$TPR[i] <- TP / (TP + FN)
    out$FPR[i] <- FP / (FP + TN)
  }

  out <- out[order(out$FPR, out$TPR), ]
  auc <- sum(diff(out$FPR) * (head(out$TPR, -1) + tail(out$TPR, -1)) / 2)
  out$model <- paste0(label, " (AUC=", round(auc, 3), ")")
  out
}

roc_all <- rbind(
  roc_from_probs(to_prob(score_weak), y_true, "Weak"),
  roc_from_probs(to_prob(score_mid), y_true, "Medium"),
  roc_from_probs(to_prob(score_strong), y_true, "Strong")
)

ggplot(roc_all, aes(x = FPR, y = TPR, color = model)) +
  geom_line(linewidth = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick") +
  labs(
    title = "ROC Curves: Weak vs Medium vs Strong Models",
    subtitle = "Higher curves and larger AUC indicate better ranking quality",
    x = "False Positive Rate",
    y = "True Positive Rate (Recall)",
    color = "Model"
  ) +
  theme_minimal(base_size = 12)
```

---

## 📈 AUC

AUC (Area Under the Curve) tells us how well the model separates Yes cases from No cases.

Think of it like ranking:

* The model sorts cases from highest risk to lowest risk.
* AUC asks: how often does a true Yes case get ranked above a true No case?
* Higher AUC means better separation between the two groups.

Quick scale:

* 0.50 -> random ranking
* 0.70-0.80 -> useful discrimination
* 0.80-0.90 -> strong discrimination
* 1.00 -> perfect separation

Important reminder:

* AUC is about ranking quality.
* AUC does not tell us if probabilities are calibrated.

---

## 🧪 AUC Example

Concrete example:

* A: actual Yes, predicted 0.90
* B: actual No, predicted 0.80
* C: actual Yes, predicted 0.60
* D: actual No, predicted 0.20

Compare each Yes with each No:

* A vs B -> correct (Yes above No)
* A vs D -> correct
* C vs B -> incorrect
* C vs D -> correct

So 3 out of 4 pairwise comparisons are correct -> AUC = 0.75.

---

## ⚖️ Class Imbalance

If Yes cases are rare:

* Accuracy can be misleading

So emphasize:

* Precision
* Recall
* Threshold strategy

Brief tools:

* Class weights
* Resampling
* Threshold tuning

---

## 🔁 Bootstrapping (Brief)

Bootstrapping:

* Resample data many times with replacement
* Recompute metrics repeatedly

Used for:

* AUC uncertainty
* Metric stability checks

Plain language:

> Run the analysis many times on slightly different versions of the dataset.

---

## 🧪 Comparing Logistic Models

If goal is prediction:

* Compare AUC
* Compare precision/recall at chosen threshold
* Compare log-loss
* Compare calibration

If goal is explanation:

* Coefficient stability
* CI width
* Interpretability
* Prefer simpler model when performance is close

---

## ✅ Final Summary

Logistic regression:

* Builds a linear score
* Converts score to probability via sigmoid
* Fits coefficients using MLE
* Interprets effects with odds ratios
* Needs threshold choice for labels
* Evaluates with confusion matrix + ROC/AUC
* Needs calibration if probabilities are used operationally

---
