---
title: "Session 10: Storytelling, Dashboards, and Delivery"
subtitle: "How to make analysis usable"
format:
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

## ⚙️ Live Setup

```{r setup-data, eval=TRUE, include=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)

resolve_data_path <- function(filename) {
  candidates <- c(
    file.path("data", filename),
    file.path("..", "..", "data", filename)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) stop("Could not find data file: ", filename)
  hit
}

ev_raw <- read_csv(resolve_data_path("wa_ev_population.csv"), show_col_types = FALSE)

story_df <- ev_raw %>%
  filter(!is.na(county), county != "", !is.na(ev_type)) %>%
  count(county, ev_type, name = "ev_registrations") %>%
  group_by(county) %>%
  summarize(
    total_evs = sum(ev_registrations),
    bev_share = sum(ev_registrations[ev_type == "Battery Electric Vehicle (BEV)"]) / total_evs,
    .groups = "drop"
  ) %>%
  arrange(desc(total_evs)) %>%
  slice_head(n = 10) %>%
  mutate(county = reorder(county, total_evs))
```

---

## 🎯 The Problem

Most weak storytelling does not fail on analysis.

It fails on translation:

* too much process
* too many equal-priority visuals
* no default takeaway
* no clear decision at the end

The audience rarely asks for "more charts."
They ask for less ambiguity.

---

## 🔍 What Actually Matters

Three questions decide whether a story works:

1. What should this audience notice first?
2. What should they believe after 60 seconds?
3. What should they do differently after seeing it?

That is the real filter.
Not "is the graphic pretty?"

---

## 🧠 Storytelling for Analysts

The audience already assumes:

* you can calculate
* you can visualize
* you can summarize

What they are testing instead:

* judgment
* prioritization
* sequencing
* restraint

---

## 🧭 A Better Mental Model

Treat storytelling as attention design.

You are deciding:

* where attention starts
* where it moves next
* where it ends

That is true in slides, dashboards, memos, and briefings.

---

## ⚖️ The Real Difference: Story vs Dashboard

**Story**

* you control sequence
* you control emphasis
* you decide what matters now

**Dashboard**

* user controls sequence
* user explores branches
* default state matters more than spoken explanation

So the design challenge changes.

---

## 🧱 Why Dashboards Get Overbuilt

Teams often confuse flexibility with value.

Typical failure mode:

* more filters
* more tabs
* more cards
* less clarity

A good dashboard does not answer everything.
It answers the most common next question fast.

---

## 📊 Why Tableau and Power BI Win

Not because they are magical.

Because organizations value:

* fast build cycles
* business-user interaction
* low engineering overhead
* repeatable publishing
* compatibility with existing data stacks

They are workflow tools as much as visualization tools.

---

## ✅ What Makes a Dashboard Good

A strong dashboard has:

* a useful default view
* obvious controls
* one primary comparison
* one or two secondary details
* no confusion about what is updating

If a stakeholder needs instructions, the design is already weak.

---

## 📈 Example: One Chart, Two Stories

```{r story-plot, eval=TRUE, fig.width=8, fig.height=4.8}
ggplot(story_df, aes(x = county, y = total_evs, fill = bev_share)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_gradient(
    low = "#c9d9d3",
    high = "#145a52",
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Large counties dominate EV volume, but composition still varies",
    subtitle = "Top Washington counties by registration volume; color shows BEV share",
    x = NULL,
    y = "Registered EVs",
    fill = "BEV share"
  ) +
  theme_minimal(base_size = 12)
```

---

## 💡 Why This Works Better Than a Generic Plot

The chart does two jobs at once:

* size answers "where is the volume?"
* color answers "what kind of adoption is it?"

That is more useful than stacking three disconnected charts.

Good storytelling often comes from combining roles, not adding slides.

---

## ✨ Shiny's Real Value

Shiny is not just "R, but interactive."

It forces you to think about:

* user intent
* reactive dependencies
* defaults
* output coordination

That makes it useful even if students never become full-time app builders.

---

## 🛠️ Shiny Design Rule

Before writing UI code, define:

* the question
* the filter set
* the primary output
* the supporting output
* the sentence the app should be able to say

If you cannot write that sentence, the app is not conceptually ready.

---

## 🧩 Minimal but Real Shiny Pattern

```{r shiny-skeleton, eval=FALSE}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("county_pick", "County", choices = county_choices),
      sliderInput("year_range", "Model year range", min_year, max_year, c(min_year, max_year))
    ),
    mainPanel(
      textOutput("story_kpi"),
      plotOutput("make_plot"),
      tableOutput("make_table")
    )
  )
)

server <- function(input, output, session) {
  filtered_story <- reactive({
    ev_df %>%
      filter(
        county == input$county_pick,
        between(model_year, input$year_range[1], input$year_range[2])
      )
  })
}
```

---

## 📝 Quarto in One Slide

Quarto matters because delivery format changes the writing job.

* Slides optimize pacing
* Articles optimize readability
* Reports optimize traceability

Same analysis, different communication contract.

That is the point.
Not the syntax.

---

## 👀 What I Want Students To Notice

Not "dashboards exist."

I want them to notice that:

* communication is part of analysis
* default state is a design decision
* interactivity can clarify or distract
* the medium changes what good work looks like

That is the transferable skill.

---

## 🧪 Lab Framing

In lab we are not building a toy.

We are building a dashboard that can answer a real question with:

* three meaningful controls
* one defensible main view
* one supporting table
* one text takeaway

That is enough complexity to be real without becoming messy.

---

## 📌 Final Point

Weak storytelling usually comes from refusing to choose.

Strong storytelling comes from choosing:

* what matters
* what waits
* what the audience should leave with

That is the job.
