---
title: "Session 8: Clustering"
subtitle: "Finding Structure When No Labels Exist"
format:
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

## Live Setup (Run Once)

```{r setup-data, eval=TRUE, include=FALSE}
library(dplyr)
library(ggplot2)

resolve_data_path <- function(filename) {
  candidates <- c(
    file.path("data", filename),
    file.path("..", "..", "data", filename)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) stop("Could not find data file: ", filename)
  hit
}

aq <- read.csv(resolve_data_path("air_quality_sites.csv"), stringsAsFactors = FALSE)

cluster_df <- aq %>%
  select(
    site_id,
    region,
    traffic_index,
    population_density,
    industrial_index,
    pm25,
    ozone,
    asthma_rate,
    median_income,
    green_space_pct
  ) %>%
  filter(complete.cases(.))

cluster_scaled <- cluster_df %>%
  select(-site_id, -region) %>%
  scale()

set.seed(5120)
km3 <- kmeans(cluster_scaled, centers = 3, nstart = 25)

pcs <- prcomp(cluster_scaled)

plot_df <- data.frame(
  PC1 = pcs$x[, 1],
  PC2 = pcs$x[, 2],
  cluster = factor(km3$cluster),
  region = cluster_df$region
)

centroid_df <- data.frame(
  PC1 = tapply(plot_df$PC1, km3$cluster, mean),
  PC2 = tapply(plot_df$PC2, km3$cluster, mean),
  cluster = factor(sort(unique(km3$cluster)))
)

wcss_df <- data.frame(
  k = 1:8,
  wcss = sapply(1:8, function(k) {
    kmeans(cluster_scaled, centers = k, nstart = 25)$tot.withinss
  })
)

hc_sample_n <- min(250, nrow(cluster_scaled))
hc_idx <- sample(seq_len(nrow(cluster_scaled)), hc_sample_n)
hc_sample <- cluster_scaled[hc_idx, ]
hc_complete <- hclust(dist(hc_sample), method = "complete")
```

---

## What's Happening Today

- Last Week:

> We had labels, and we were trying to predict them.

- This Week:

> We do not have labels, and we are trying to discover structure.

---

## Why Do We Even Do Clustering?

Sometimes the data is messy, large, and unlabeled.

We still want to know:

* Are there natural groupings here?
* Which cases look similar to each other?
* Are some places meaningfully different from the rest?

Plain idea:

> Clustering is a way to organize observations into groups based on similarity.

---

## What Is Clustering Used For?

Common use cases:

* Group neighborhoods with similar pollution patterns
* Segment customers with similar behavior
* Find schools, hospitals, or counties with similar profiles
* Detect unusual cases that do not fit the rest

Why that matters:

* It helps us summarize complex data before making decisions.

---

## Clustering vs Classification

**Classification**

* We already know the labels
* Example: funded vs not funded
* Goal: predict the correct category

**Clustering**

* We do not know the labels
* Goal: discover possible groupings

Key line:

> Classification answers "Which known group is this?"  
> Clustering answers "Are there groups here at all?"

---

## A Practical Difference

With classification:

* There is usually a right answer in the training data.

With clustering:

* There may not be one single "correct" answer.

That means clustering is more exploratory.

It is often the start of analysis, not the end.

---

## What Does "Similar" Mean?

Clustering depends on a similarity rule.

Most of the time, that means a **distance** rule.

Simple intuition:

* Small distance -> observations are similar
* Large distance -> observations are different

If the distance rule changes, the clusters can change too.

---

## Euclidean Distance

The most common distance rule is **Euclidean distance**.

Plain idea:

* Draw a straight line from one point to another
* The length of that line is the distance

Formula:

$$
d(x, y) = \sqrt{\sum_{j=1}^{p}(x_j - y_j)^2}
$$

What matters for us:

* Bigger differences across variables mean bigger distance
* Variables on large scales can dominate unless we standardize first

---

## Why Scaling Matters So Much

Suppose one variable runs from 0 to 10000, and another runs from 0 to 10.

Without scaling:

* The large-scale variable dominates the distance calculation.

So before clustering, we often standardize variables so they are on comparable scales.

> In clustering, preprocessing is not optional. It often determines the answer.

---

## K-Means: The Basic Idea

K-means asks us to choose a number of clusters, called **K**.

Then it tries to:

1. Place K centers in the data
2. Assign each point to its nearest center
3. Recompute each center as the mean of the observations assigned to it
4. Repeat until assignments stop changing much

Plain language:

> K-means keeps redrawing group centers until the groups settle down.

---

## What K-Means Is Optimizing

K-means tries to make points within a cluster as close as possible to that cluster's center.

The objective is:

$$
\sum_{k=1}^{K}\sum_{i \in C_k} ||x_i - \mu_k||^2
$$

You do not need to memorize the formula.

The idea is enough:

* Tight clusters are good
* Large within-cluster spread is bad

---

## K-Means Live Example

```{r kmeans-plot, eval=TRUE, fig.width=8, fig.height=4.8}
ggplot(plot_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_point(
    data = centroid_df,
    aes(x = PC1, y = PC2),
    inherit.aes = FALSE,
    shape = 4,
    size = 5,
    stroke = 1.4,
    color = "black"
  ) +
  labs(
    title = "K-means example on scaled air-quality features",
    subtitle = "Colors show 3 clusters; black X marks the cluster centroids",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

---

## K-Means Limitations

K-means is useful, but it has strong assumptions built in.

Main limits:

* You must choose K ahead of time
* It works best when clusters are fairly round and compact
* It is sensitive to outliers
* Different starting points can give different results

Important reminder:

> K-means will always return clusters, even if the data does not truly have clean clusters.

---

## Hierarchical Clustering: The Basic Idea

Hierarchical clustering does not start with a fixed K.

Instead, it builds a tree of merges:

1. Start with every observation as its own cluster
2. Merge the two closest clusters
3. Repeat until everything is merged

Output:

* A hierarchy of possible groupings

---

## The Dendrogram

A dendrogram is the tree diagram from hierarchical clustering.

What it shows:

* Which observations merge early
* Which groups stay separate for a long time

Interpretation:

* Lower merge height -> more similar
* Higher merge height -> less similar

---

## Hierarchical Clustering Live Example

```{r hclust-plot, eval=TRUE, fig.width=9, fig.height=4.8}
plot(
  hc_complete,
  labels = FALSE,
  hang = -1,
  main = "Hierarchical clustering dendrogram",
  xlab = "Air-quality sites (sample)",
  ylab = "Merge height"
)
rect.hclust(hc_complete, k = 3, border = "firebrick")
```

---

## Linkage Choice Changes the Tree

When hierarchical clustering decides which clusters to merge, it needs a rule.

Common linkage choices:

* **Single linkage**: nearest two points
* **Complete linkage**: farthest two points
* **Average linkage**: average pairwise distance

This matters because:

* Different linkage rules can produce different dendrograms

In this example, we used **complete linkage**:

> The distance between two clusters is defined by the two points that are farthest apart.

That tends to prefer tighter, more compact clusters than single linkage.

---

## Hierarchical Clustering Limitations

Main limits:

* It can be sensitive to the distance rule and linkage choice
* Early bad merges cannot be undone
* It can become slow on very large datasets
* Outliers can distort the tree

Plain language:

> Hierarchical clustering is flexible, but it is not neutral.  
> Your design choices shape the result.

---

## When Do We Use K-Means?

K-means is usually a good choice when:

* You want a quick, simple baseline
* You have a rough idea of how many clusters you want
* The dataset is fairly large
* Variables are numeric and scaled

Good fit:

* Fast segmentation tasks
* First pass exploration

---

## When Do We Use Hierarchical Clustering?

Hierarchical clustering is often better when:

* You do not want to commit to one K immediately
* You want to inspect the structure visually
* The dataset is not huge
* You care about nested relationships

Good fit:

* Exploratory analysis
* Smaller datasets where interpretation matters

---

## Trade-Offs: K-Means vs Hierarchical

| Question | K-means | Hierarchical |
|---|---|---|
| Need to choose K first? | Yes | Not at first |
| Fast on larger data? | Usually yes | Usually less scalable |
| Easy to explain? | Yes | Fairly yes, with dendrogram |
| Handles weird cluster shapes well? | Not usually | Sometimes better, but depends |
| Sensitive to setup choices? | Yes | Yes |

There is no universal winner.

> The best method depends on the data and the goal.

---

## Choosing the Number of Clusters

This is one of the hardest parts.

Because in real data:

* There may not be one perfect answer

Common tools:

* Elbow plot
* Silhouette score
* Domain knowledge

---

## Elbow Plot Example

```{r elbow-plot, eval=TRUE, fig.width=8, fig.height=4.8}
ggplot(wcss_df, aes(x = k, y = wcss)) +
  geom_line(color = "#1f5c8b", linewidth = 1) +
  geom_point(color = "firebrick", size = 2) +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title = "Elbow plot for K-means",
    subtitle = "Look for the point where improvement starts slowing down",
    x = "Number of clusters (K)",
    y = "Total within-cluster sum of squares"
  ) +
  theme_minimal(base_size = 12)
```

---

## How Do We Evaluate Clustering?

This is different from regression or classification.

Why?

* We often do not have known labels
* So we cannot always talk about accuracy

Instead, we usually use three kinds of evaluation:

1. Internal evaluation
2. External evaluation
3. Substantive evaluation

---

## 1. Internal Evaluation

Internal evaluation asks:

> Are the clusters tight and well separated?

Examples:

* Within-cluster sum of squares
* Silhouette score

Good internal performance means:

* Members of the same cluster look similar
* Different clusters look meaningfully apart

---

## Silhouette Score

Silhouette score gives each point a simple report card.

It compares:

* How close the point is to its own cluster
* How close the point is to the nearest other cluster

Interpretation:

* Near **1** -> well matched to its own cluster
* Near **0** -> sitting near a boundary
* Below **0** -> may be in the wrong cluster

Plain takeaway:

> Higher average silhouette score usually means cleaner cluster separation.

---

## 2. External Evaluation

External evaluation is possible only if we have known labels somewhere.

Question:

> Do the discovered clusters line up with something real that we already know?

Examples:

* Known region types
* Known risk categories
* Previously defined site classes

This is not always available, but when it is, it helps.

---

## 3. Substantive Evaluation

Substantive evaluation asks:

> Are these clusters actually useful?

That means:

* Can we describe them clearly?
* Do they support a real policy or management decision?
* Do they reveal something we did not already know?

This is the evaluation step students often skip.

---

## A Big Warning About Interpretation

Clusters are not automatically "real" natural categories.

They are the result of:

* Variable choices
* Scaling choices
* Distance choices
* Algorithm choices

So be careful with language.

Better phrasing:

> "The algorithm identified three groups with different profiles."

Be cautious with:

> "There are truly three types of counties."

---

## Common Clustering Mistakes

The most common problems:

* Forgetting to scale variables
* Treating clusters as truth instead of a model output
* Choosing K with no justification
* Ignoring outliers
* Interpreting clusters without checking whether they are useful

---

## A Reasonable Workflow

When you do clustering:

1. Choose variables carefully
2. Scale them if distance is being used
3. Try more than one clustering setup
4. Check cluster quality
5. Interpret clusters in plain language
6. Ask whether the result helps a real decision

---

## Final Summary

Clustering:

* Finds patterns when labels are missing
* Depends heavily on similarity rules and scaling
* Uses K-means for fast partitioning
* Uses hierarchical clustering for tree-based structure
* Needs evaluation beyond "it made a plot"
* Should be treated as a decision aid, not automatic truth

---
