---
title: "Lab 06 – Logistic Regression in Practice"
subtitle: "Climate Policy Adoption: Baseline, Expanded Model, and Cutoff Tuning"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are advising a state climate office on which counties are likely to adopt a policy package.

**Your Goal:** Build and compare three logistic workflows:
- V1: simple baseline model
- V2: expanded model with more predictors
- V3: V2 with ROC-based best cutoff

---

## Dataset Variables (Used in This Lab)

- `policy_adopted`: Binary outcome (1 = adopted, 0 = not adopted)
- `compliance_score`: Compliance readiness score
- `public_support_pct`: Public support percentage
- `inspection_count`: Number of inspections
- `grants_million`: Grant amount (millions)
- `emissions_baseline_index`: Baseline emissions index
- `region_type`: Categorical county type
- `policy_type`: Categorical policy program type

---

## 0. Decision Question

> Which counties are likely to adopt the policy package, and what probability cutoff should be used to classify likely adoption?

---

## 1. Load Raw Data

**functions:**
- `read_csv()` – load dataset.
- `glimpse()` – inspect column types.

**logic steps:**
- Load county-level policy dataset.
- Verify variables needed for logistic regression.

```{r}
policy_raw <- read_csv("../../../data/climate_policy_adoption.csv", show_col_types = FALSE)
glimpse(policy_raw)
```

---

## 2. Clean Data

**functions:**
- `mutate()` – set variable types.
- `filter()` / `drop_na()` – keep usable rows.

**logic steps:**
- Keep modeling columns only.
- Convert `region_type` and `policy_type` to factors.
- Ensure outcome is numeric 0/1.

```{r}
policy_clean <- policy_raw %>%
  transmute(
    policy_adopted = as.integer(policy_adopted),
    compliance_score,
    public_support_pct,
    inspection_count,
    grants_million,
    emissions_baseline_index,
    region_type = as.factor(region_type),
    policy_type = as.factor(policy_type)
  ) %>%
  drop_na()

table(policy_clean$policy_adopted)
```

---

## 3. Train/Test Split (70/30)

**functions:**
- `set.seed()` – reproducible split.
- `sample.int()` – random row selection.

**logic steps:**
- Split data into 70% training and 30% testing.
- Save as `train_df` and `test_df`.

```{r}
set.seed(5120)
idx <- sample.int(nrow(policy_clean), size = floor(0.7 * nrow(policy_clean)))
train_df <- policy_clean[idx, ]
test_df <- policy_clean[-idx, ]

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")
```

---

## 4. Model V1: Simple Baseline Logistic Regression

**functions:**
- `glm(..., family = binomial())`
- `summary()`
- `exp(coef())`

**logic steps:**
- Fit V1 with one predictor: `compliance_score`.
- Interpret direction and odds-ratio meaning.

```{r}
model_v1 <- glm(
  policy_adopted ~ compliance_score,
  data = train_df,
  family = binomial()
)

summary(model_v1)
exp(coef(model_v1))
```

**Question 4.1:** Interpret the `compliance_score` coefficient in V1 using odds-ratio language.

---

## 5. Evaluate V1 (ROC, AUC, Accuracy, Precision, Recall)

**functions:**
- `predict(type = "response")`
- `roc()` / `auc()` from `pROC`
- `ggplot()`

**logic steps:**
- Generate V1 test probabilities.
- Compute ROC and AUC.
- Compute accuracy, precision, and recall at cutoff 0.50.

```{r}
compute_metrics <- function(actual, p_hat, cutoff) {
  pred <- ifelse(p_hat >= cutoff, 1L, 0L)

  TP <- sum(pred == 1 & actual == 1)
  FP <- sum(pred == 1 & actual == 0)
  TN <- sum(pred == 0 & actual == 0)
  FN <- sum(pred == 0 & actual == 1)

  data.frame(
    accuracy = (TP + TN) / length(actual),
    precision = ifelse((TP + FP) == 0, 0, TP / (TP + FP)),
    recall = ifelse((TP + FN) == 0, 0, TP / (TP + FN))
  )
}

roc_to_df <- function(roc_obj, model_name) {
  data.frame(
    fpr = 1 - roc_obj$specificities,
    tpr = roc_obj$sensitivities,
    model = model_name
  )
}

p_v1 <- predict(model_v1, newdata = test_df, type = "response")
roc_v1 <- roc(response = test_df$policy_adopted, predictor = p_v1, quiet = TRUE)
auc_v1 <- as.numeric(auc(roc_v1))

metrics_v1 <- compute_metrics(test_df$policy_adopted, p_v1, cutoff = 0.50)
metrics_v1$model <- "V1 (cutoff=0.50)"
metrics_v1$cutoff <- 0.50
metrics_v1$AUC <- auc_v1
metrics_v1 <- metrics_v1[, c("model", "cutoff", "accuracy", "precision", "recall", "AUC")]
metrics_v1

roc_df_v1 <- roc_to_df(roc_v1, "V1")

ggplot(roc_df_v1, aes(x = fpr, y = tpr)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = paste0("ROC Curve - V1 (AUC = ", round(auc_v1, 3), ")"),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal()
```

**Question 5.1:** Interpret the ROC curve for V1 in plain language.

**Question 5.2:** Interpret the AUC value for V1 in plain language.

---

## 6. Model V2: Expanded Logistic Regression

**functions:**
- `glm(..., family = binomial())`
- `summary()`
- `exp(coef())`
- `predict(type = "response")`

**logic steps:**
- Add multiple predictors and categorical controls.
- Fit V2 and evaluate at cutoff 0.50.

```{r}
model_v2 <- glm(
  policy_adopted ~ compliance_score + public_support_pct + inspection_count + grants_million + emissions_baseline_index + region_type + policy_type,
  data = train_df,
  family = binomial()
)

summary(model_v2)
exp(coef(model_v2))

p_v2 <- predict(model_v2, newdata = test_df, type = "response")
roc_v2 <- roc(response = test_df$policy_adopted, predictor = p_v2, quiet = TRUE)
auc_v2 <- as.numeric(auc(roc_v2))

metrics_v2 <- compute_metrics(test_df$policy_adopted, p_v2, cutoff = 0.50)
metrics_v2$model <- "V2 (cutoff=0.50)"
metrics_v2$cutoff <- 0.50
metrics_v2$AUC <- auc_v2
metrics_v2 <- metrics_v2[, c("model", "cutoff", "accuracy", "precision", "recall", "AUC")]
metrics_v2
```

**Question 6.1:** Which variable in V2 appears most influential, and why?

**Question 6.2:** Choose one numeric coefficient in V2 and interpret it as an odds ratio.

---

## 7. Best Cutoff for V2 (Using ROC)

Use Youden's J:

$$
J = TPR - FPR
$$

**functions:**
- `coords()` from `pROC` with `x = "best"` and `best.method = "youden"`.

**logic steps:**
- Extract the best threshold from V2 ROC.
- Save as `best_cutoff_v2`.

```{r}
best_coords <- coords(
  roc_v2,
  x = "best",
  best.method = "youden",
  ret = c("threshold", "sensitivity", "specificity"),
  transpose = FALSE
)

best_cutoff_v2 <- as.numeric(best_coords["threshold"])
best_cutoff_v2
best_coords
```

**Question 7.1:** Why is this cutoff better than 0.50 for this model?

---

## 8. Model V3: V2 with Best Cutoff + Confusion Matrix

**functions:**
- `compute_metrics()`
- `table()`

**logic steps:**
- Keep model form same as V2.
- Reclassify predictions using `best_cutoff_v2`.
- Build confusion matrix for V3.

```{r}
model_v3 <- model_v2
p_v3 <- p_v2
auc_v3 <- auc_v2

metrics_v3 <- compute_metrics(test_df$policy_adopted, p_v3, cutoff = best_cutoff_v2)
metrics_v3$model <- paste0("V3 (cutoff=", round(best_cutoff_v2, 2), ")")
metrics_v3$cutoff <- best_cutoff_v2
metrics_v3$AUC <- auc_v3
metrics_v3 <- metrics_v3[, c("model", "cutoff", "accuracy", "precision", "recall", "AUC")]
metrics_v3

pred_v3 <- ifelse(p_v3 >= best_cutoff_v2, 1L, 0L)
cm_v3 <- table(
  Predicted = factor(pred_v3, levels = c(1, 0), labels = c("Yes", "No")),
  Actual = factor(test_df$policy_adopted, levels = c(1, 0), labels = c("Yes", "No"))
)

cm_v3
```

**Question 8.1:** Interpret the V3 confusion matrix: what do false positives and false negatives mean in this policy context?

---

## 9. Compare V1 vs V2 vs V3

**functions:**
- `rbind()`
- `order()` / `arrange()`
- `ggplot()`

**logic steps:**
- Build one comparison table with cutoff, accuracy, precision, recall, and AUC.
- Include AUC for all three versions (V3 has same ranking as V2).
- Compare model tradeoffs and recommend one setup.

```{r}
model_comparison <- rbind(metrics_v1, metrics_v2, metrics_v3)
model_comparison <- model_comparison[order(-model_comparison$recall, -model_comparison$precision), ]
model_comparison

roc_df_compare <- rbind(
  roc_to_df(roc_v1, "V1"),
  roc_to_df(roc_v2, "V2"),
  roc_to_df(roc_v2, "V3 (same ranking as V2)")
)

ggplot(roc_df_compare, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 1.05) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  labs(
    title = "ROC Comparison Across Model Setups",
    subtitle = paste0(
      "AUC V1 = ", round(auc_v1, 3),
      ", AUC V2 = ", round(auc_v2, 3),
      ", AUC V3 = ", round(auc_v3, 3)
    ),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal()
```

**Question 9.1:** Which model setup is best for this context, and why?

**Question 9.2:** Should the policy team prioritize precision or recall here? Explain with context.

**Question 9.3:** Is accuracy alone enough in this dataset? Briefly justify your answer.

---

## 10. Final Reflection

**Summary Question 10.1:**
If you had to brief leadership in two minutes, which model/cutoff would you deploy, and what is one limitation they should remember?
