---
title: "Lab 04 – Proving the Pattern"
subtitle: "Hypothesis Testing with EV Infrastructure"
author: "ENIA-5120"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
---

## 🧭 Mission Background
You are a policy analyst evaluating whether observed EV patterns are likely to reflect **real differences** or just **random variation**.

**Your Goal:** Use hypothesis tests to compare EV demand and charging infrastructure patterns across groups.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)

# Load data
ev_pop <- read_csv("../../../data/wa_ev_population.csv") 
ev_stations <- read_csv("../../../data/wa_ev_charge_stations.csv") 

# Clean station groups
stations_clean <- ev_stations %>%
  mutate(
    zip = as.character(zip),
    station_group = case_when(
      str_detect(station_group, "Public") ~ "Public",
      str_detect(station_group, "Private") ~ "Private",
      TRUE ~ "Unknown"
    ),
    station_group = as.factor(station_group)
  )
```

---

## 1. Distribution Check (Normality & CLT)

**functions:**
- `ggplot()` + `geom_histogram()` – Visualize distribution.

**logic steps:**
- Plot the distribution of `electric_range` and identify skewness.
- Explain why CLT still supports testing with large samples.

```{r}
"<write your code here>"
```

**Question 1.1.1:** Is this distribution approximately normal or skewed?  
**Question 1.1.2:** Why can we still use a t-test with a very large sample?

---

## 2. Standardizing with Z-scores

**Scenario:** A local official claims a station with 15 Level 2 ports is “unusually large.” Use a z-score to check that claim.

**functions:**
- `mean()` / `sd()` – compute average and spread.

**logic steps:**
- Compute the statewide mean and standard deviation for `ev_lvl_2`.
- Calculate the z-score for 15 ports.
- Interpret how unusual that value is.

```{r}
"<write your code here>"
```

**Question 2.1.1:** What is the z-score for 15 ports?  
**Question 2.1.2:** Is this typical, somewhat unusual, or extreme?

---

## 3. Comparing Two Means (t-test)

**Scenario:** A policymaker believes BEVs have a higher average range than PHEVs. Test whether the averages differ.

**Hypotheses:**  
$H_0$: $\mu_{BEV} = \mu_{PHEV}$  
$H_a$: $\mu_{BEV} \neq \mu_{PHEV}$  

**functions:**
- `filter()` – remove obvious missing values.
- `t.test()` – compare means.

**logic steps:**
- Filter out zero ranges (treated as missing).
- Run a two-sample t-test by `ev_type`.
- Use the p-value and confidence interval to interpret the result.

```{r}
"<write your code here>"
```

**Question 3.1.1:** What is the p-value?  
**Question 3.1.2:** Does the 95% confidence interval include 0?  
**Question 3.1.3:** At $\alpha = 0.05$, what decision would you make?  
**Question 3.1.4:** Which group appears to have the higher average range?

---

## 4. Comparing Multiple Groups (ANOVA)

**Scenario:** Are Level 2 ports distributed differently across Public, Private, and Unknown station groups?

**Hypotheses:**  
$H_0$: $\mu_{Public} = \mu_{Private} = \mu_{Unknown}$  
$H_a$: At least one group mean differs  

**functions:**
- `ggplot()` + `geom_boxplot()` – visualize group differences.
- `aov()` – run ANOVA.

**logic steps:**
- Make a boxplot to compare distributions by station group.
- Run ANOVA to test whether any group mean differs.
- Interpret the F statistic and p-value.

```{r}
"<write your code here>"
```

**Question 4.1.1:** What is the F value?  
**Question 4.1.2:** What does the p-value imply at $\alpha = 0.05$?  
**Question 4.1.3:** What can ANOVA not tell you?

---

## 5. Testing Relationships (Chi-square)

**Scenario:** Do the top 3 counties have the same mix of EV types, or different mixes?

**Hypotheses:**  
$H_0$: EV type and county are independent  
$H_a$: They are associated  

**functions:**
- `count()` – find top counties.
- `filter()` – keep top counties.
- `table()` – create contingency table.
- `chisq.test()` – test association.

**logic steps:**
- Find the top 3 counties by EV count.
- Build a county × EV type table.
- Run a chi-square test and interpret the p-value.

```{r}
"<write your code here>"
```

**Question 5.1.1:** What is the $\chi^2$ statistic?  
**Question 5.1.2:** What does the p-value suggest?  
**Question 5.1.3:** If expected counts are very small, what should you do?

---
