---
title: "Session 3 – Data Visualization with ggplot2"
format: 
  revealjs:
    slide-number: true
    smaller: true
    scrollable: true    
    footer: "ENIA-5120 · Veo Chae · Georgetown University"
    self-contained: true
    theme: serif     
    fig-width: 8
    fig-height: 5    
execute:
  echo: true    # Shows the code on the slide
  eval: true    # RUNS the code
  warning: false
  message: false 

---
```{r setup, include=FALSE}
# This chunk runs but is hidden from the slides
library(readr)
library(dplyr)
library(ggplot2)
players <- read_csv("laser_league_players.csv")
```


# Today's Agenda 🗺️

1. **The Impact:** Why visuals matter in the history of policy.   
2. **Visual Literacy:** How to spot (and avoid) misleading charts.   
3. **The Grammar of Graphics:** The logic and syntax of `ggplot2`.   
4. **The Workflow:** Interactive "Build-a-Plot" coding.   
5. **Professional Polish:** Transforming drafts into report-ready assets.   

---

# Learning Goals 🎯
By the end of this session, you will be able to:

-   **Translate** raw data into a visual story that supports a policy claim.
-   **Diagnose** common visual "lies" like truncated axes and 3D distortions.
-   **Implement** the layered grammar of graphics using `ggplot2`.
-   **Construct** accessible and professional charts that follow high data-to-ink standards.
-   **Export** high-resolution visuals for inclusion in policy briefs and memos.

---

# 2. The Power of the Visual 🔎

## Why Visualize? 

> "A bad visual is worse than no visual at all."

-   **Discovery:** Finding patterns your brain can't see in a spreadsheet.
-   **Translation:** Taking 10,000 rows of data and making them understandable in 10 seconds.
-   **Influence:** Policy isn't changed by numbers; it's changed by the *stories* numbers tell.

---

## Hero Example: John Snow (1854) 🗺️
**The Problem:** Cholera was thought to be "bad air" (miasma).
**The Visual:** Mapping deaths to street addresses and water pumps.



**The Result:** The visual clustering around the Broad Street pump proved it was waterborne. Officials removed the handle, and modern epidemiology was born.

---

## Hero Example: Florence Nightingale (1858) 🌹
**The Problem:** Soldiers were dying in hospitals, not on the battlefield.
**The Visual:** The "Rose Diagram."



**The Result:** It proved that poor sanitation was the primary killer. It revolutionized hospital hygiene policy across the British Empire.

---

# 3. Visual Literacy: Spot the Lie ⚠️

## The "Dirty Tricks" of Visualization
1.  **Truncated Y-Axes:** Starting at 90 instead of 0 to exaggerate small differences.
2.  **The 3D Trap:** 3D bars and pie charts distort proportions—objects in the "front" always look bigger.
3.  **Chartjunk:** Unnecessary decorations that distract from the data.



---

## Ethics & Accessibility ♿
-   **Color Blindness:** ~8% of men have some form of color deficiency.
-   **The Solution:** Use palettes like `viridis` that map clearly even in grayscale or for color-blind readers.
-   **The "Squint Test":** If you squint at your graph and the main point disappears, the visual is too cluttered.

---

# 4. The Grammar of Graphics 🧩

## ggplot2: The Layered Logic 🧱  
Think of a plot like a **sandwich**:   

1.  **Data:** The bread (The `laser_league_players` CSV).   
2.  **Aesthetics (`aes`):** The "mapping" (X = Stamina, Y = Score, Color = Role).   
3.  **Geometries (`geoms`):** The "shape" (Dots, Bars, or Lines).   
 

---

## Choosing Your Geometry 📐
| Goal | Question | Geometry |
|:---|:---|:---|
| **Distribution** | How spread out is the Wellbeing Index? | `geom_histogram()` |
| **Comparison** | Which squad has the highest reflex scores? | `geom_boxplot()` |
| **Relationship** | Does stamina predict season score? | `geom_point()` |

---

# 5. Interactive Build-a-Plot 🛠️

## Step 1: The Canvas 🎨
We start with nothing—just the coordinate system.

```{r}
#| echo: true
#| eval: true
library(ggplot2)
ggplot(data = players, mapping = aes(x = tag_accuracy_pct, y = season_score))
```

## Step 2: Adding the Points ✨
Now we add the "Geometric" layer. Note the `+` sign!
```{r}
#| echo: true
#| eval: true
ggplot(players, aes(x = tag_accuracy_pct, y = season_score)) +
  geom_point()
```

## Step 3: Handling Overplotting 🌫️
Points are stacking up! We use Alpha (transparency) to see the density.

```{r}
#| echo: true
#| eval: true
ggplot(players, aes(x = tag_accuracy_pct, y = season_score)) +
  geom_point(alpha = 0.5) # 50% transparent
```

## Step 4: Adding the "Policy Story" 🧭
Let's add a statistical trend line and color by role.

```{r}
#| echo: true
#| eval: true
ggplot(players, aes(x = tag_accuracy_pct, y = season_score, color = role)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) # Adds a linear trend line
```

## Step 5: Professional Polish ✨  
- From "Draft" to "Briefing-Ready"
- We must remove the "Code-speak" (underscores) from our labels.

```{r}
#| echo: true
#| eval: true
ggplot(players, aes(x = tag_accuracy_pct, y = season_score)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  theme_minimal() + 
  labs(
    title = "Does Tag Accuracy Predict Success?",
    subtitle = "Higher tag accuracy scores show a strong correlation with season performance",
    x = "Tag Accuracy Score (0-100)",
    y = "Total Season Score",
    caption = "Source: Laser League Player Performance Data (Session 03)"
  )
```

## Step 6: Faceting Small Multiples 🧪
Don't cram everything into one messy plot. Split it into "Small Multiples."

```{r}
#| echo: true
#| eval: true
ggplot(players, aes(x = tag_accuracy_pct, y = season_score)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~role) + # One mini-plot per player role
  theme_bw()
```

# Wrapping Up ✅
**The Professional Visual Checklist**  
  - Zero Baseline: Do my bars start at zero?   
  - Plain Language: Are titles and axes free of code-speak?   
  - Accessibility: Is my color choice readable (Viridis/Colorblind friendly)?   
  - Transparency: If I have many points, did I use alpha?   
  - The "So What?": Does the plot answer a specific policy question?   

---

# Next Steps 🚀
- Open Lab 03: The Commissioner's Visual Briefing (Advanced Storytelling) in RStudio.
- Load the laser_league_players.csv dataset.
